---
title: "Workshop Setup Guide"
subtitle: "RNA-seq Bioinformatics Workshop - File Structure and Configuration"
author: "UF Health Cancer Center Bioinformatics"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-copy: true
    code-overflow: wrap
    embed-resources: true
    theme: cosmo
    number-sections: true
---

# Overview

This document explains the workshop setup, how `here()` works with a cloned repository, and where your data files will live.

# Workshop File Layout

## What You'll Create (Your Working Space)

```
/blue/group/YOUR_USERNAME/
└── rnaseq_workshop/          # You create this and clone repo here
    ├── .git/                 # Created by git clone (here() detects this!)
    ├── data/                 # Processed data will be saved here
    │   ├── rsem.merged.gene_counts.tsv  (created by prep script)
    │   ├── sample_info.tsv              (created by prep script)
    │   └── gene_annotation.tsv          (created by prep script)
    ├── scripts/              # From cloned repo
    │   └── prepare_nfcore_data.R
    ├── analysis/             # From cloned repo
    │   └── rnaseq_workshop_session2.qmd
    ├── results/              # Created automatically during analysis
    └── figures/              # Created automatically during analysis
```

## Shared Workshop Data (Read-Only)

```
/blue/cancercenter-dept/TRAINING/
└── nfcore_output/            # Shared nf-core pipeline results
    └── star_rsem/
        └── rsem.merged.gene_counts.tsv  (and other pipeline outputs)
```

# How `here()` Works in This Setup

::: {.callout-note}
## The `here` Package

The `here` package will automatically detect your **cloned repository** as the project root because it contains a `.git` folder.
:::

**When you run code from RStudio:**

- `here()` returns: `/blue/group/YOUR_USERNAME/rnaseq_workshop/`
- `here("data", "counts.tsv")` → `/blue/group/YOUR_USERNAME/rnaseq_workshop/data/counts.tsv`
- `here("results", "output.csv")` → `/blue/group/YOUR_USERNAME/rnaseq_workshop/results/output.csv`

This works **the same for everyone**, regardless of their username or exact path!

# Setup Instructions

## Create Your Working Directory and Clone the Repo

On HiPerGator:

```bash
# Navigate to your group's blue storage
cd /blue/bioinf_workshop/$USER/

# Clone the workshop repository
git clone https://github.com/UFHCC-BCBSR/res-bioinfo-rnaseq-workshop.git rnaseq_workshop

# Navigate into the cloned repo
cd rnaseq_workshop

# Verify the structure
ls -la
# You should see: .git/, data/, scripts/, analysis/, etc.
```

::: {.callout-important}
## The `.git/` Folder is Key!

The `.git/` folder (created by `git clone`) is what makes `here()` work! Don't delete it.
:::

## Prepare the Data (Run Once)

The data preparation script will:

- Read from the shared nf-core output directory (external, read-only)
- Write processed files to YOUR repo's `data/` directory

```bash
# Still in /blue/bioinf_workshop/$USER/rnaseq_workshop/

# Open RStudio Server and navigate to this directory
# OR run from command line:
module load R
Rscript scripts/prepare_nfcore_data.R
```

This will create:

- `data/rsem.merged.gene_counts.tsv`
- `data/sample_info.tsv`
- `data/gene_annotation.tsv`

## Start RStudio Server

```bash
# Launch RStudio Server session
# Make sure you start it FROM your rnaseq_workshop directory
```

In RStudio:

1. File → Open Project
2. Select `rnaseq_workshop` (if not auto-detected)
3. Check that `here()` is working:

```{r}
#| eval: false
library(here)
here()  # Should show: /blue/bioinf_workshop/$USER/rnaseq_workshop
```

# Using the `here` Package

All workshop materials use the `here` package for file paths within the repository. This ensures:

- **Portability**: Works on any computer/HPC system
- **Reproducibility**: Same code works for everyone
- **No hardcoded paths**: No need to change file paths in the code

## Two Types of Paths in This Workshop

::: {.callout-tip}
## Understanding the Two-Path System

This workshop uses two different path strategies:
:::

**1. External Input (Absolute Path):**

```{r}
#| eval: false
# The nf-core output is OUTSIDE your repo
# The prep script uses an absolute path to read it
nfcore_results_dir <- "/blue/cancercenter-dept/TRAINING/nfcore_output"
```

**2. Repo-Relative Paths (Using `here()`):**

```{r}
#| eval: false
# Everything else uses here() relative to your cloned repo
counts <- read.table(here("data", "rsem.merged.gene_counts.tsv"))
ggsave(here("figures", "plot.png"))
write.csv(results, here("results", "output.csv"))
```

## How `here()` Works

::: {.callout-note}
## Before and After

**Instead of this** (breaks when you move files or share with others):
```r
counts <- read.table("/blue/alice/rnaseq_workshop/data/counts.tsv")
```

**Use this** (works for everyone):
```r
counts <- read.table(here("data", "counts.tsv"))
```
:::

The `here()` function automatically finds your project root by looking for:

1. `.git` folder ← **Your cloned repo has this!**
2. `.Rproj` file (optional)
3. `.here` file (optional)

Since you cloned the workshop repo, the `.git` folder serves as the marker.

# Running the Analysis

## Option 1: In RStudio

1. Open `rnaseq_workshop.Rproj` (if you created one)
2. Open `analysis/rnaseq_workshop_session2.qmd`
3. Click "Render" button or press `Ctrl+Shift+K`

## Option 2: Command Line

```bash
cd /path/to/rnaseq_workshop

# Render the Quarto document
module load quarto
quarto render analysis/rnaseq_workshop_session2.qmd
```

# Troubleshooting

## `here()` is pointing to the wrong directory

Check what `here()` returns:

```{r}
#| eval: false
library(here)
here()
```

**Should return:** `/blue/bioinf_workshop/$USER/rnaseq_workshop`

::: {.callout-warning}
## If it returns something else:

- Make sure you're working inside the cloned repo directory
- Check that the `.git` folder exists: `list.files(all.files = TRUE)`
- Make sure you **started R/RStudio from inside the workshop repo**, not from a parent directory
- Try restarting R/RStudio
:::

## "Cannot find file" errors

```{r}
#| eval: false
# Debug: check where here() thinks you are
library(here)
cat("Project root:", here(), "\n")

# Check if data files exist
file.exists(here("data", "rsem.merged.gene_counts.tsv"))
file.exists(here("data", "sample_info.tsv"))

# If FALSE, you may need to run the data prep script first
```

## Data prep script can't find nf-core output

Make sure the path in `scripts/prepare_nfcore_data.R` matches the actual location:

```{r}
#| eval: false
# Check this line in the script:
nfcore_results_dir <- "/blue/cancercenter-dept/TRAINING/nfcore_output"

# Ask instructors for the correct path if different
```

## Package not installed

```{r}
#| eval: false
# Install here if needed
install.packages("here")

# Or all workshop packages at once
install.packages(c("tidyverse", "ggrepel", "pheatmap", "RColorBrewer", "here"))
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(c("limma", "edgeR"))
```

# Output Files

After running the analysis, you'll find:

- **figures/**: PNG files of all plots
- **results/**: CSV files with DE results
- **analysis/*.html**: Rendered HTML report

::: {.callout-tip}
## Portability

All paths are relative, so you can:

- Move the entire `rnaseq_workshop` folder anywhere
- Share it with collaborators
- Run it on different computers
- Everything will still work!
:::

# Questions?

Contact: [Your Email]

---

# Appendix: Common `here()` Examples

Here are common patterns you'll see throughout the workshop:

```{r}
#| eval: false

# Load data
counts <- read.table(here("data", "rsem.merged.gene_counts.tsv"))
metadata <- read.table(here("data", "sample_info.tsv"))

# Save plots
ggsave(here("figures", "volcano_plot.png"), width = 8, height = 6)
ggsave(here("figures", "heatmap.png"), width = 10, height = 8)

# Save results
write.csv(results, here("results", "de_genes.csv"), row.names = TRUE)
write.csv(top_genes, here("results", "top_50_genes.csv"))

# Check if directory exists
if (!dir.exists(here("results"))) {
  dir.create(here("results"), recursive = TRUE)
}

# List files in a directory
list.files(here("data"))
```

The pattern is always: `here("folder", "subfolder", "filename")`
