---
title: "Part 2: Differential Expression Analysis"
author: "UFHCI BCB-SR Bioinformatics Unit"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

In this notebook we test for differentially expressed genes between PRMT7
knockdown and wildtype samples using the limma-voom framework.

**Requires:** `DGE_filtered_normalized.rds` produced by `02_quality_control.Rmd`

---

## Setup

```{r libraries}
library(limma)
library(edgeR)
library(tidyverse)
library(ggrepel)
library(pheatmap)
library(RColorBrewer)
library(here)

set.seed(42)

results_dir <- here("demo-analysis", "output", "differential-expression", "results")
figures_dir <- here("demo-analysis", "output", "differential-expression", "figures")

dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(figures_dir, showWarnings = FALSE, recursive = TRUE)
```

---

## Load Data

Load the filtered and normalized DGEList object saved at the end of Part 1.

```{r load-rds}
DGE <- readRDS(here("demo-analysis", "output", "differential-expression",
                    "DGE_filtered_normalized.rds"))

cat("Genes:", nrow(DGE), "| Samples:", ncol(DGE), "\n")
cat("Groups:", paste(levels(DGE$samples$group), collapse = ", "), "\n")
DGE$samples
```

---

## Design Matrix

The design matrix tells limma which samples belong to which groups and what
we want to test. The first factor level (`PRMT7_WT`) is the reference — fold
changes are expressed relative to WT.

- **Intercept**: Average expression in PRMT7_WT
- **genotypePRMT7_KD**: The log2 fold-change between KD and WT — this is what
  we are testing

```{r design-matrix}
# WT is the reference level (listed first)
DGE$samples$genotype <- factor(DGE$samples$genotype,
                                levels = c("PRMT7_WT", "PRMT7_KD"))

design <- model.matrix(~genotype, data = DGE$samples)
design
colnames(design)
```

---

## voom Transformation

voom converts counts to log2-CPM and calculates a precision weight for each
observation based on the mean-variance relationship. The resulting plot should
show a smooth downward-sloping red curve — highly expressed genes have more
stable variance and receive higher weights in the model.

```{r voom, fig.width=8, fig.height=5}
# normalize.method = "none" because TMM was already applied in Part 1
v <- voom(DGE,
          design           = design,
          normalize.method = "none",
          plot             = TRUE)
```

---

## Fit Linear Model

`lmFit` fits a linear model to each gene. `eBayes` applies empirical Bayes
moderation, which borrows information across genes to stabilize variance
estimates — particularly important for small sample sizes.

```{r fit-model}
fit <- lmFit(v, design)
fit <- eBayes(fit, robust = TRUE)

# Summary of DE genes at FDR < 0.05
# Down = lower in KD, Up = higher in KD
summary(decideTests(fit, adjust.method = "fdr", p.value = 0.05, lfc = 1))
```

---

## Extract Results

`topTable` extracts the full results table. `coef = 2` refers to the
`genotypePRMT7_KD` coefficient — the KD vs WT comparison.

```{r extract-results}
top_genes_strict <- topTable(fit, 
                             coef = 2, 
                             adjust.method = "BH", 
                             number = 20, 
                             sort.by = "P", 
                             lfc = 1)
cat("\nTop 20 genes (FDR < 0.05, |logFC| > 1):\n")
print(top_genes_strict)

cat("Total genes tested:    ", nrow(top_genes_strict), "\n")
cat("DE genes (FDR < 0.05):", sum(top_genes_strict$adj.P.Val < 0.05), "\n")
cat("Up in KD:             ", sum(top_genes_strict$adj.P.Val < 0.05 & top_genes_strict$logFC > 0), "\n")
cat("Down in KD:           ", sum(top_genes_strict$adj.P.Val < 0.05 & top_genes_strict$logFC < 0), "\n\n")

# Top 20 by p-value
topTable(fit, coef = 2, adjust.method = "BH", number = 20, sort.by = "P", lfc = 1)



```

---

## Volcano Plot

A volcano plot shows log-fold change (x-axis) against statistical significance
(y-axis). Genes in the upper corners are both large in effect size and highly
significant. Dashed lines mark a 2-fold change threshold and p = 0.05.

```{r volcano, fig.width=8, fig.height=6}
# All results for visualization
all_results <- topTable(fit,
                        coef = 2,
                        adjust.method = "BH",
                        number = Inf,
                        sort.by = "none")

all_results$category <- case_when(
  all_results$adj.P.Val < 0.05 & all_results$logFC >  1 ~ "Up (>2-fold)",
  all_results$adj.P.Val < 0.05 & all_results$logFC < -1 ~ "Down (>2-fold)",
  all_results$adj.P.Val < 0.05                           ~ "Significant",
  TRUE                                                    ~ "Not Significant"
)

table(all_results$category)

ggplot(all_results, aes(x = logFC, y = -log10(P.Value), color = category)) +
  geom_point(alpha = 0.5, size = 1) +
  scale_color_manual(values = c("Up (>2-fold)" = "red",
                                "Down (>2-fold)" = "blue",
                                "Significant" = "orange",
                                "Not Significant" = "grey80")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", alpha = 0.5) +
  labs(title = "Volcano Plot: PRMT7 KD vs WT",
       subtitle = "Thresholds: FDR < 0.05, |logFC| > 1 (2-fold change)",
       x = "Log2 Fold Change",
       y = "-Log10 P-value",
       color = "Category") +
  theme_minimal() +
  theme(legend.position = "right")

ggsave(here(figures_dir, "volcano_plot.png"), width = 8, height = 6, dpi = 300)
```

---

## MA Plot

An MA plot shows log-fold change (y-axis) against average expression (x-axis).
The bulk of genes should sit around logFC = 0, with DE genes scattered above
and below. A systematic trend would suggest a normalization problem.

```{r ma-plot, fig.width=8, fig.height=6}
ggplot(all_results, aes(x = AveExpr, y = logFC, color = category)) +
  geom_point(alpha = 0.5, size = 1) +
  scale_color_manual(values = c("Up (>2-fold)" = "red",
                                "Down (>2-fold)" = "blue",
                                "Significant" = "orange",
                                "Not Significant" = "grey80")) +
  geom_hline(yintercept = 0, linetype = "solid", alpha = 0.5) +
  geom_hline(yintercept = c(-1, 1), linetype = "dashed", alpha = 0.3) +
  labs(title = "MA Plot: PRMT7 KD vs WT",
       subtitle = "Thresholds: FDR < 0.05, |logFC| > 1 (2-fold change)",
       x = "Average Expression (log2-CPM)",
       y = "Log2 Fold Change",
       color = "Category") +
  theme_minimal()

ggsave(here(figures_dir, "ma_plot.png"), width = 8, height = 6, dpi = 300)
```

---

## Heatmap of Top 50 DE Genes

Expression values are z-scored per gene (row) so that expression patterns are
comparable across genes regardless of absolute expression level. WT and KD
samples should separate clearly if the DE signal is strong.

```{r heatmap, fig.width=8, fig.height=10}
top50        <- rownames(topTable(fit, coef = 2, number = 50, lfc = 1))
top50_logcpm <- cpm(DGE[top50, ], log = TRUE)
top50_scaled <- t(scale(t(top50_logcpm)))  # Z-score per gene

annotation_col <- data.frame(
  Genotype  = DGE$samples$genotype,
  row.names = colnames(top50_logcpm)
)

# Display in document
pheatmap(top50_scaled,
         annotation_col = annotation_col,
         cluster_rows   = TRUE,
         cluster_cols   = TRUE,
         show_rownames  = TRUE,
         show_colnames  = TRUE,
         main           = "Top 50 DE Genes",
         fontsize_row   = 6,
         fontsize_col   = 8,
         color          = colorRampPalette(c("blue", "white", "red"))(100))

# Save to file
pheatmap(top50_scaled,
         annotation_col = annotation_col,
         cluster_rows   = TRUE,
         cluster_cols   = TRUE,
         show_rownames  = TRUE,
         show_colnames  = TRUE,
         main           = "Top 50 DE Genes",
         fontsize_row   = 6,
         fontsize_col   = 8,
         color          = colorRampPalette(c("blue", "white", "red"))(100),
         filename       = here(figures_dir, "heatmap_top50.png"),
         width          = 8,
         height         = 10)
```

---

## Export Results

```{r export}
all_results$gene <- rownames(all_results)

write.csv(all_results,
          here(results_dir, "de_results_all.csv"),
          row.names = FALSE)

# Significant genes only
sig_results <- all_results %>% filter(adj.P.Val < 0.05 & abs(logFC) > 1)

write.csv(sig_results,
          here(results_dir, "de_results_significant.csv"),
          row.names = FALSE)

cat("✓ de_results_all.csv:        ", nrow(all_results), "genes\n")
cat("✓ de_results_significant.csv:", nrow(sig_results), "genes\n")
```

---

## Session Information

```{r sessioninfo}
writeLines(capture.output(sessionInfo()),
           here(results_dir, "sessionInfo_limma.txt"))
sessionInfo()
```
