---
title: "Part 1: Quality Control and Normalization"
author: "UFHCI BCB-SR Bioinformatics Unit"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

Before testing for differential expression, we need to assess sample quality,
remove genes that are too lowly expressed to test reliably, and normalize the
data so that samples are comparable.

**In this notebook you will:**

- Load count data and sample metadata
- Create a DGEList object
- Filter lowly expressed genes
- Apply TMM normalization
- Visualize sample quality

**Output:** A normalized DGEList saved as `DGE_filtered_normalized.rds` for
use in the differential expression script.

---

## Setup

Load required packages and define output directories.

```{r libraries}
library(limma)
library(edgeR)
library(tidyverse)
library(pheatmap)
library(RColorBrewer)
library(ggrepel)
library(here)

set.seed(42)

figures_dir <- here("demo-analysis", "output", "02-differential-expression", "figures")
rds_dir     <- here("demo-analysis", "output", "02-differential-expression")

dir.create(figures_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(rds_dir,     showWarnings = FALSE, recursive = TRUE)
```

```{r paths}
counts_file   <- here("demo-analysis", "output", "01-prepared-data",
                      "rsem.merged.gene_counts.tsv")
metadata_file <- here("demo-analysis", "data", "metadata", "sample_metadata.csv")

if (!file.exists(counts_file))   stop("Count file not found: ",    counts_file)
if (!file.exists(metadata_file)) stop("Metadata file not found: ", metadata_file)

cat("✓ Count file found\n")
cat("✓ Metadata file found\n")
```

---

## Load Data

### Gene Counts

The nf-core/rnaseq pipeline with RSEM produces a merged count matrix. We need
the raw expected counts (`rsem.merged.gene_counts.tsv`), not the TPM or scaled
variants.

```{r load-counts}
expression_data <- read.table(counts_file,
                               header      = TRUE,
                               row.names   = 1,
                               check.names = FALSE)

cat("Genes:", nrow(expression_data), "| Samples:", ncol(expression_data), "\n")
head(expression_data, n = 3)
```

### Sample Metadata

```{r load-metadata}
sample_info <- read.csv(metadata_file,
                         header           = TRUE,
                         stringsAsFactors = FALSE)

sample_info

# Confirm sample names match count matrix columns
all(sample_info$sample %in% colnames(expression_data))
```

---

## Create DGEList Object

The `DGEList` object bundles counts and sample metadata into a single
container used throughout the edgeR and limma-voom workflow.

```{r create-dge}
DGE <- DGEList(counts  = expression_data,
               samples = sample_info)

# Set the group factor explicitly — WT is the reference level (listed first)
DGE$samples$group <- factor(DGE$samples$genotype,
                              levels = c("PRMT7_WT", "PRMT7_KD"))
DGE
```

---

## Filter Lowly Expressed Genes

Genes with very low counts across all samples add noise without contributing
useful signal and inflate the multiple testing burden. `filterByExpr()` applies
a principled heuristic: a gene must reach a minimum expression level in enough
samples given the experimental design.

```{r filter}
keep.exprs <- filterByExpr(DGE, group = DGE$samples$group)
table(keep.exprs)

DGE <- DGE[keep.exprs, , keep.lib.sizes = FALSE]
cat("Genes remaining after filtering:", nrow(DGE), "\n")
```

---

## TMM Normalization

Even after filtering, samples differ in library size (total reads) and
composition bias (a few highly expressed genes can dominate the library).
TMM normalization calculates a scaling factor for each sample to correct
for both. Factors close to 1.0 indicate little correction was needed.

```{r normalize}
DGE <- calcNormFactors(DGE, method = "TMM")
DGE$samples
```

---

## Quality Control

### Library Sizes

Library size is the total number of reads per sample after filtering. Large
differences (>2–3×) between samples may indicate technical issues. Moderate
variation is normal and is handled by normalization.

```{r library-sizes, fig.width=8, fig.height=5}
lib_size_df <- data.frame(
  sample   = rownames(DGE$samples),
  lib_size = DGE$samples$lib.size,
  group    = DGE$samples$group
)

ggplot(lib_size_df, aes(x = sample, y = lib_size / 1e6, fill = group)) +
  geom_col() +
  labs(title = "Library Sizes After Filtering",
       x = "Sample", y = "Library Size (Millions of Reads)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Set2")
```

### MDS Plot

Multidimensional Scaling (MDS) places samples in two-dimensional space such
that similar samples are closer together. Samples from the same group should
cluster together and separate clearly from the other group. Outlier samples
or unexpected clustering patterns are red flags.

```{r mds-plot, fig.width=8, fig.height=6}
mds <- plotMDS(DGE, top = 500, plot = FALSE, gene.selection = "common")

mds_df <- data.frame(
  sample   = rownames(DGE$samples),
  dim1     = mds$x,
  dim2     = mds$y,
  genotype = DGE$samples$genotype
)

ggplot(mds_df, aes(x = dim1, y = dim2, color = genotype, label = sample)) +
  geom_point(size = 4, alpha = 0.8) +
  geom_text_repel(size = 3, max.overlaps = 20) +
  labs(title    = "MDS Plot — Sample Similarity",
       subtitle = "Top 500 most variable genes",
       x = paste0("Dim 1 (", round(mds$var.explained[1] * 100, 1), "%)"),
       y = paste0("Dim 2 (", round(mds$var.explained[2] * 100, 1), "%)")) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")

ggsave(here(figures_dir, "mds_plot.png"), width = 8, height = 6, dpi = 300)
```

### Sample Correlation Heatmap

Pairwise Spearman correlations between all samples. Samples in the same group
should show higher correlations with each other than with samples in the other
group. Typical within-group correlations are 0.95–0.99.

```{r correlation-heatmap, fig.width=8, fig.height=7}
logcpm     <- cpm(DGE, log = TRUE)
cor_matrix <- cor(logcpm, method = "spearman")

annotation_col <- data.frame(
  Genotype  = DGE$samples$genotype,
  row.names = colnames(logcpm)
)

# Save to file
pheatmap(cor_matrix,
         annotation_col = annotation_col,
         annotation_row = annotation_col,
         main     = "Sample Correlation Heatmap",
         color    = colorRampPalette(c("blue", "white", "red"))(100),
         breaks   = seq(0.7, 1, length.out = 101),
         fontsize = 8,
         filename = here(figures_dir, "correlation_heatmap.png"),
         width = 8, height = 7)

# Display in document
pheatmap(cor_matrix,
         annotation_col = annotation_col,
         annotation_row = annotation_col,
         main     = "Sample Correlation Heatmap",
         color    = colorRampPalette(c("blue", "white", "red"))(100),
         breaks   = seq(0.7, 1, length.out = 101),
         fontsize = 8)
```

---

## Save Processed Data

Save the filtered and normalized DGEList so the differential expression script
can load it directly without re-running QC.

```{r save-rds}
saveRDS(DGE, file = here(rds_dir, "DGE_filtered_normalized.rds"))

cat("✓ Saved:", here(rds_dir, "DGE_filtered_normalized.rds"), "\n")
cat("  Genes:", nrow(DGE), "| Samples:", ncol(DGE), "\n")
cat("  Groups:", paste(levels(DGE$samples$group), collapse = ", "), "\n")
```

---

## Session Information

```{r sessioninfo}
sessionInfo()
```
