---
title: "Part 1: Quality Control and Normalization"
subtitle: "UF Health Cancer Institute Bioinformatics Workshop - Differential Expression Analysis"
author: "UFHCI BCB-SR Bioinformatics Unit"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: false
    code-tools: true
    code-overflow: wrap
    embed-resources: true
    theme: cosmo
    df-print: paged
execute:
  warning: false
  message: false
---

# Introduction

Before we can test for differentially expressed genes, we need to assess
sample quality, remove genes that are too lowly expressed to test reliably,
and normalize the data so that samples are comparable.

**In this notebook you will:**

- Load count data and sample metadata
- Create a DGEList object — the core data structure for this analysis
- Filter lowly expressed genes
- Apply TMM normalization
- Visualize sample quality and relationships

**Output:** A normalized, filtered DGEList object saved as
`DGE_filtered_normalized.rds` for use in the differential expression analysis.

---

# Setup

## Load Required Libraries

```{r setup}
library(limma)
library(edgeR)
library(tidyverse)
library(pheatmap)
library(RColorBrewer)
library(here)

set.seed(42)

# Define output directories
figures_dir <- here("demo-analysis", "output", "02-differential-expression", "figures")
rds_dir <- here("demo-analysis", "output", "02-differential-expression")

dir.create(figures_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(rds_dir, showWarnings = FALSE, recursive = TRUE)

cat("Project root:", here(), "\n")
cat("Figures directory:", figures_dir, "\n")
```

::: {.callout-tip}
## Package Installation
If you haven't installed these packages yet:
```r
install.packages(c("tidyverse", "pheatmap", "RColorBrewer"))
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(c("limma", "edgeR"))
```
:::

## Set File Paths

```{r paths}
counts_file <- here("demo-analysis", "output", "01-prepared-data",
                    "rsem.merged.gene_counts.tsv")
metadata_file <- here("demo-analysis", "data", "metadata", "sample_metadata.csv")

if (!file.exists(counts_file)) stop("Count file not found: ", counts_file)
if (!file.exists(metadata_file)) stop("Metadata file not found: ", metadata_file)

cat("✓ Count file found\n")
cat("✓ Metadata file found\n")
```

---

# Load Data

## Load Gene Counts

```{r load-counts}
expression_data <- read.table(counts_file,
                              header = TRUE,
                              row.names = 1,
                              check.names = FALSE)

cat("Dimensions:", nrow(expression_data), "genes x",
    ncol(expression_data), "samples\n\n")

head(expression_data, n = 3)
```

::: {.callout-important}
## RSEM Expected Counts
RSEM produces **expected counts** — suitable for limma-voom or DESeq2.
These are not TPM values. Make sure you are using the
`rsem.merged.gene_counts.tsv` file (not the scaled or TPM variants).
:::

## Load Sample Metadata

```{r load-metadata}
sample_info <- read.csv(metadata_file,
                        header = TRUE,
                        stringsAsFactors = FALSE)

sample_info

# Verify sample names match count data
all(sample_info$sample %in% colnames(expression_data))
```

---

# Create DGEList Object

The `DGEList` object bundles counts and sample metadata into a single
container used throughout the edgeR and limma-voom workflow.

```{r create-dge}
DGE <- DGEList(counts = expression_data,
               samples = sample_info)

# Set the group factor explicitly — required for filtering and DE testing
DGE$samples$group <- factor(DGE$samples$genotype,
                             levels = c("PRMT7_WT", "PRMT7_KD"))
DGE
```

::: {.callout-note}
## The DGEList Object
`DGE$counts` contains the count matrix. `DGE$samples` contains sample
metadata including library sizes and normalization factors. As we process
the data, everything stays bundled together in this object.
:::

---

# Filter Lowly Expressed Genes

Genes with very low counts across all samples:

- Add noise without contributing useful signal
- Inflate the multiple testing burden
- Are often unreliable measurements

`filterByExpr()` applies a principled heuristic: a gene must reach a
minimum expression level in enough samples to be testable given your
experimental design.

```{r filter-genes}
keep.exprs <- filterByExpr(DGE, group = DGE$samples$group)

table(keep.exprs)

DGE <- DGE[keep.exprs, , keep.lib.sizes = FALSE]

cat("Genes remaining after filtering:", nrow(DGE), "\n")
```

::: {.callout-note}
## filterByExpr
By default, `filterByExpr()` requires a minimum count of 10 reads in at
least as many samples as the smallest group. This is more principled than
a simple count cutoff because it accounts for your experimental design.
:::

---

# TMM Normalization

Even after filtering, samples differ in:

1. **Library size** — total number of reads
2. **Composition bias** — a few highly expressed genes can dominate the
   library, making other genes appear lower than they are

TMM (Trimmed Mean of M-values) normalization corrects for both by
calculating a scaling factor for each sample.

```{r normalize}
DGE <- calcNormFactors(DGE, method = "TMM")

DGE$samples
```

::: {.callout-note}
## Normalization Factors
- Factors close to **1.0**: little correction needed
- Factors **< 1**: sample has upward composition bias
- Factors **> 1**: sample has downward composition bias

These factors are applied internally by limma-voom — you do not need to
manually adjust your counts.
:::

---

# Quality Control Visualizations

## Library Sizes

This plot shows the total number of reads per sample after filtering.
The samples are colored by experimental group.

```{r library-sizes, fig.width=8, fig.height=5}
lib_size_df <- data.frame(
  sample   = rownames(DGE$samples),
  lib_size = DGE$samples$lib.size,
  group    = DGE$samples$group
)

ggplot(lib_size_df, aes(x = sample, y = lib_size / 1e6, fill = group)) +
  geom_col() +
  labs(title = "Library Sizes After Filtering",
       x = "Sample",
       y = "Library Size (Millions of Reads)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Set2")
```

::: {.callout-warning}
## What to Look For
- Large differences (>2–3×) in library sizes may indicate technical issues
- Very low libraries (<5M reads for human/mouse) may lack sufficient depth
- Moderate variation is normal and handled by normalization
:::

## MDS Plot

Multidimensional Scaling (MDS) is similar to PCA — it places samples in
two-dimensional space such that similar samples are closer together.
Samples from the same experimental group should cluster together.

```{r mds-plot, fig.width=8, fig.height=6}
mds <- plotMDS(DGE, top = 500, plot = FALSE, gene.selection = "common")

mds_df <- data.frame(
  sample   = rownames(DGE$samples),
  dim1     = mds$x,
  dim2     = mds$y,
  genotype = DGE$samples$genotype
)

ggplot(mds_df, aes(x = dim1, y = dim2, color = genotype, label = sample)) +
  geom_point(size = 4, alpha = 0.8) +
  ggrepel::geom_text_repel(size = 3, max.overlaps = 20) +
  labs(title = "MDS Plot — Sample Similarity",
       subtitle = "Based on top 500 most variable genes",
       x = paste0("Leading logFC dim 1 (",
                  round(mds$var.explained[1] * 100, 1), "%)"),
       y = paste0("Leading logFC dim 2 (",
                  round(mds$var.explained[2] * 100, 1), "%)")) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")

ggsave(here(figures_dir, "mds_plot.png"), width = 8, height = 6, dpi = 300)
```

::: {.callout-important}
## Interpreting the MDS Plot
**Good signs:**
- Samples from the same group cluster together
- Clear separation between conditions

**Red flags:**
- Outlier samples far from their group
- No separation between conditions (weak signal)
- Samples clustering by batch instead of biology
:::

## Sample Correlation Heatmap

This heatmap shows pairwise Spearman correlations between all samples.
Samples in the same group should show higher correlations with each other
than with samples in the other group.

```{r correlation-heatmap, fig.width=8, fig.height=7}
logcpm <- cpm(DGE, log = TRUE)
cor_matrix <- cor(logcpm, method = "spearman")

annotation_col <- data.frame(
  Genotype = DGE$samples$genotype,
  row.names = colnames(logcpm)
)

# Save to file
pheatmap(cor_matrix,
         annotation_col = annotation_col,
         annotation_row = annotation_col,
         main = "Sample Correlation Heatmap",
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(0.7, 1, length.out = 101),
         fontsize = 8,
         filename = here(figures_dir, "correlation_heatmap.png"),
         width = 8,
         height = 7)

# Display in document
pheatmap(cor_matrix,
         annotation_col = annotation_col,
         annotation_row = annotation_col,
         main = "Sample Correlation Heatmap",
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(0.7, 1, length.out = 101),
         fontsize = 8)
```

::: {.callout-tip}
## Expected Correlation Values
- **Within-group**: typically 0.95–0.99
- **Between-group**: typically 0.85–0.98 (depends on effect size)
- **Outlier samples**: consistently < 0.90 correlation with all others
:::

---

# Save Processed Data

Save the filtered and normalized DGEList object so the differential
expression script can load it directly without re-running QC.

```{r save-rds}
saveRDS(DGE, file = here(rds_dir, "DGE_filtered_normalized.rds"))

cat("✓ DGE object saved to:",
    here(rds_dir, "DGE_filtered_normalized.rds"), "\n\n")

cat("Summary:\n")
cat("  Genes after filtering:", nrow(DGE), "\n")
cat("  Samples:", ncol(DGE), "\n")
cat("  Groups:", paste(levels(DGE$samples$group), collapse = ", "), "\n")
```

---

# Session Information

```{r sessioninfo}
sessionInfo()
```
