---
title: "Part 2: Differential Expression Analysis"
subtitle: "UF Health Cancer Institute Bioinformatics Workshop - Differential Expression Analysis"
author: "UFHCI BCB-SR Bioinformatics Unit"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: false
    code-tools: true
    code-overflow: wrap
    embed-resources: true
    theme: cosmo
    df-print: paged
execute:
  warning: false
  message: false
---

# Introduction

In this notebook we test for differentially expressed genes between PRMT7
knockdown and wildtype samples using the limma-voom framework.

**You will:**

- Load the normalized DGEList object from Part 1
- Build a design matrix describing the experimental groups
- Apply the voom transformation
- Fit a linear model and apply empirical Bayes moderation
- Extract, visualize, and export results

**Requires:** `DGE_filtered_normalized.rds` produced by
`02_quality_control.qmd/.Rmd`

---

# Setup

```{r setup}
library(limma)
library(edgeR)
library(tidyverse)
library(ggrepel)
library(pheatmap)
library(RColorBrewer)
library(here)

set.seed(42)

results_dir <- here("demo-analysis", "output", "02-differential-expression", "results")
figures_dir <- here("demo-analysis", "output", "02-differential-expression", "figures")

dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(figures_dir, showWarnings = FALSE, recursive = TRUE)
```

---

# Load Data

Load the filtered and normalized DGEList object saved at the end of Part 1.

```{r load-rds}
DGE <- readRDS(here("demo-analysis", "output", "02-differential-expression",
                    "DGE_filtered_normalized.rds"))

cat("Genes:", nrow(DGE), "\n")
cat("Samples:", ncol(DGE), "\n")
cat("Groups:", paste(levels(DGE$samples$group), collapse = ", "), "\n")

DGE$samples
```

---

# Design Matrix

The design matrix tells limma which samples belong to which groups and
what we want to test.

```{r design-matrix}
# Ensure factor levels are correct — WT is the reference
DGE$samples$genotype <- factor(DGE$samples$genotype,
                                levels = c("PRMT7_WT", "PRMT7_KD"))

# Create design matrix
design <- model.matrix(~genotype, data = DGE$samples)

design
colnames(design)
```

::: {.callout-note}
## Understanding the Design Matrix

- **Intercept**: Average expression in the reference group (PRMT7_WT)
- **genotypePRMT7_KD**: The difference between KD and WT — this is the
  log2 fold-change we are testing

A positive coefficient means higher expression in KD; negative means
lower expression in KD.
:::

---

# voom Transformation

voom converts counts to log2-CPM and calculates a precision weight for
each observation based on the mean-variance relationship. This is what
allows a method designed for normally distributed data (linear models)
to work well with count data.

```{r voom, fig.width=8, fig.height=5}
v <- voom(DGE,
          design = design,
          normalize.method = "none",
          plot = TRUE)
```

::: {.callout-note}
## Interpreting the voom Plot

- **Red curve**: The mean-variance trend being modeled
- Lowly expressed genes (left) have high variance — voom down-weights these
- Highly expressed genes (right) have stable variance — voom gives these more weight

A smooth, downward-sloping curve indicates the data are well-behaved.
:::

---

# Fit Linear Model

```{r fit-model}
# Fit a linear model to each gene
fit <- lmFit(v, design)

# Empirical Bayes moderation borrows information across genes to
# stabilize variance estimates, especially for small sample sizes
fit <- eBayes(fit, robust = TRUE)

# Summary of DE genes at FDR < 0.05
summary(decideTests(fit, adjust.method = "fdr", p.value = 0.05))
```

::: {.callout-note}
## decideTests Output

- **Down**: Significantly lower in KD vs WT
- **NotSig**: Not significantly different
- **Up**: Significantly higher in KD vs WT
:::

---

# Extract Results

```{r extract-results}
# Extract all results (not just the top genes)
all_results <- topTable(fit,
                        coef = 2,
                        adjust.method = "BH",
                        number = Inf,
                        sort.by = "none")

# Summary
cat("Total genes tested:", nrow(all_results), "\n")
cat("DE genes (FDR < 0.05):", sum(all_results$adj.P.Val < 0.05), "\n")
cat("Upregulated in KD:", sum(all_results$adj.P.Val < 0.05 & all_results$logFC > 0), "\n")
cat("Downregulated in KD:", sum(all_results$adj.P.Val < 0.05 & all_results$logFC < 0), "\n\n")

# Top 20 genes by p-value
topTable(fit, coef = 2, adjust.method = "BH", number = 20, sort.by = "P")
```

---

# Visualization

## Volcano Plot

A volcano plot shows the relationship between fold change (x-axis) and
statistical significance (y-axis). Genes in the upper corners are both
large in effect size and highly significant.

```{r volcano, fig.width=8, fig.height=6}
all_results$category <- case_when(
  all_results$adj.P.Val < 0.05 & all_results$logFC >  1 ~ "Up",
  all_results$adj.P.Val < 0.05 & all_results$logFC < -1 ~ "Down",
  all_results$adj.P.Val < 0.05                           ~ "Significant",
  TRUE                                                    ~ "Not Significant"
)

ggplot(all_results, aes(x = logFC, y = -log10(P.Value), color = category)) +
  geom_point(alpha = 0.5, size = 1) +
  scale_color_manual(values = c("Up"              = "red",
                                "Down"            = "blue",
                                "Significant"     = "orange",
                                "Not Significant" = "grey80")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", alpha = 0.5) +
  labs(title  = "Volcano Plot: PRMT7_KD vs PRMT7_WT",
       x      = "Log2 Fold Change",
       y      = "-Log10 P-value",
       color  = "Category") +
  theme_minimal()

ggsave(here(figures_dir, "volcano_plot.png"), width = 8, height = 6, dpi = 300)
```

::: {.callout-tip}
## Reading a Volcano Plot

- **Dashed vertical lines**: ±1 log2FC (2-fold change)
- **Dashed horizontal line**: p = 0.05
- **Upper right**: Strongly and significantly upregulated in KD
- **Upper left**: Strongly and significantly downregulated in KD
:::

## MA Plot

An MA plot shows log-fold change (y-axis) versus average expression
(x-axis). Ideally the bulk of genes sit around logFC = 0, with DE genes
scattered above and below.

```{r ma-plot, fig.width=8, fig.height=6}
ggplot(all_results, aes(x = AveExpr, y = logFC, color = category)) +
  geom_point(alpha = 0.5, size = 1) +
  scale_color_manual(values = c("Up"              = "red",
                                "Down"            = "blue",
                                "Significant"     = "orange",
                                "Not Significant" = "grey80")) +
  geom_hline(yintercept = 0, linetype = "solid", alpha = 0.5) +
  labs(title = "MA Plot: PRMT7_KD vs PRMT7_WT",
       x     = "Average Expression (log2-CPM)",
       y     = "Log2 Fold Change",
       color = "Category") +
  theme_minimal()

ggsave(here(figures_dir, "ma_plot.png"), width = 8, height = 6, dpi = 300)
```

::: {.callout-note}
## Interpreting the MA Plot

The MA plot is useful for checking whether fold changes are balanced
around zero across expression levels. A systematic trend (e.g. all
highly expressed genes are positive) would suggest a normalization
problem.
:::

## Heatmap of Top 50 DE Genes

```{r heatmap, fig.width=8, fig.height=10}
top50 <- rownames(topTable(fit, coef = 2, number = 50))

top50_logcpm <- cpm(DGE[top50, ], log = TRUE)
top50_scaled <- t(scale(t(top50_logcpm)))

annotation_col <- data.frame(
  Genotype  = DGE$samples$genotype,
  row.names = colnames(top50_logcpm)
)

pheatmap(top50_scaled,
         annotation_col = annotation_col,
         cluster_rows   = TRUE,
         cluster_cols   = TRUE,
         show_rownames  = TRUE,
         show_colnames  = TRUE,
         main           = "Top 50 DE Genes",
         fontsize_row   = 6,
         fontsize_col   = 8,
         color          = colorRampPalette(c("blue", "white", "red"))(100))

pheatmap(top50_scaled,
         annotation_col = annotation_col,
         cluster_rows   = TRUE,
         cluster_cols   = TRUE,
         show_rownames  = TRUE,
         show_colnames  = TRUE,
         main           = "Top 50 DE Genes",
         fontsize_row   = 6,
         fontsize_col   = 8,
         color          = colorRampPalette(c("blue", "white", "red"))(100),
         filename       = here(figures_dir, "heatmap_top50.png"),
         width          = 8,
         height         = 10)
```

::: {.callout-tip}
## Reading the Heatmap

Each row is a gene, each column is a sample. Colors show z-scored
expression (red = above average, blue = below average for that gene).
Samples and genes are clustered by similarity. You should see the WT and
KD samples separate clearly if the DE signal is strong.
:::

---

# Export Results

```{r export}
# All tested genes
all_results$gene <- rownames(all_results)

write.csv(all_results,
          here(results_dir, "de_results_all.csv"),
          row.names = FALSE)

# Significant genes only (FDR < 0.05)
sig_results <- all_results %>% filter(adj.P.Val < 0.05)

write.csv(sig_results,
          here(results_dir, "de_results_significant.csv"),
          row.names = FALSE)

cat("✓ de_results_all.csv:", nrow(all_results), "genes\n")
cat("✓ de_results_significant.csv:", nrow(sig_results), "genes\n")
```

---

# Session Information

```{r sessioninfo}
writeLines(capture.output(sessionInfo()),
           here(results_dir, "sessionInfo_limma.txt"))

sessionInfo()
```
