---
title: "Part 3: Pathway Analysis"
author: "UFHCI BCB-SR Bioinformatics Unit"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

Having identified differentially expressed genes, we now ask what they mean
biologically. Pathway analysis uses over-representation testing to ask: are
our DE genes enriched in particular biological processes or pathways compared
to what we would expect by chance?

**Requires:** `de_results_all.csv` produced by `03_differential_expression.Rmd`

---

## Setup

```{r libraries}
library(clusterProfiler)
library(org.Mm.eg.db)   # Mouse annotation — use org.Hs.eg.db for human
library(enrichplot)
library(tidyverse)
library(patchwork)
library(here)

set.seed(42)

results_dir <- here("demo-analysis", "output", "differential-expression", "results")
figures_dir <- here("demo-analysis", "output", "differential-expression", "figures")

dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(figures_dir, showWarnings = FALSE, recursive = TRUE)
```

---

## Load DE Results

```{r load-results}
all_results <- read.csv(here(results_dir, "de_results_all.csv"))
rownames(all_results) <- all_results$gene

sig_genes  <- all_results %>% filter(adj.P.Val < 0.05)
up_genes   <- sig_genes  %>% filter(logFC > 0)
down_genes <- sig_genes  %>% filter(logFC < 0)

cat("Total DE genes (FDR < 0.05):", nrow(sig_genes), "\n")
cat("Upregulated in KD:          ", nrow(up_genes), "\n")
cat("Downregulated in KD:        ", nrow(down_genes), "\n")
```

---

## Convert Gene IDs

Most pathway databases use Entrez Gene IDs rather than gene symbols. We convert
here and also define a background gene set — all genes that were actually tested.
Using the correct background is important for accurate enrichment statistics:
comparing to all genes in the genome rather than genes in your experiment
inflates false positives.

A conversion rate of 85–95% is normal. Genes that fail to convert are typically
novel genes, pseudogenes, or non-coding RNAs not yet in the annotation database.

```{r convert-ids}
# Convert DE gene symbols to Entrez IDs
gene_entrez <- bitr(rownames(sig_genes),
                    fromType = "SYMBOL",
                    toType   = "ENTREZID",
                    OrgDb    = org.Mm.eg.db)

# Background: all genes tested
background_entrez <- bitr(rownames(all_results),
                           fromType = "SYMBOL",
                           toType   = "ENTREZID",
                           OrgDb    = org.Mm.eg.db)

universe <- background_entrez$ENTREZID

cat("DE genes submitted:     ", nrow(sig_genes), "\n")
cat("Successfully converted: ", nrow(gene_entrez), "\n")
cat("Background genes:       ", length(universe), "\n")
```

---

## GO Enrichment — Biological Process

Gene Ontology (GO) Biological Process terms describe what genes do at the
cellular and systems level (e.g., "DNA repair", "cell cycle arrest"). Key
output columns:

- **GeneRatio**: Fraction of your DE genes annotated to this term
- **BgRatio**: Fraction of background genes annotated to this term
- **p.adjust**: FDR-corrected p-value
- **geneID**: Gene symbols driving the enrichment

```{r go-bp}
ego_bp <- enrichGO(gene          = gene_entrez$ENTREZID,
                   universe      = universe,
                   OrgDb         = org.Mm.eg.db,
                   ont           = "BP",
                   pAdjustMethod = "BH",
                   pvalueCutoff  = 0.05,
                   qvalueCutoff  = 0.05,
                   readable      = TRUE)  # Convert Entrez IDs back to symbols

cat("Enriched GO-BP terms:", nrow(ego_bp), "\n\n")
head(ego_bp, 10)
```

### Dot Plot

The dot plot shows the top enriched terms. X-axis is the gene ratio, dot size
is the number of genes, and dot color is the adjusted p-value.

```{r go-dotplot, fig.width=10, fig.height=8}
dotplot(ego_bp,
        showCategory = 20,
        title        = "GO Biological Process Enrichment",
        font.size    = 10)

ggsave(here(figures_dir, "GO_BP_dotplot.png"),
       width = 10, height = 8, dpi = 300)
```

### Enrichment Map

Shows relationships between enriched terms — terms sharing many genes are
connected and cluster around common biological themes.

```{r go-emap, fig.width=12, fig.height=10}
ego_bp_sim <- pairwise_termsim(ego_bp)

emapplot(ego_bp_sim, showCategory = 30, layout = "kk") +
  labs(title = "GO Enrichment Map")

ggsave(here(figures_dir, "GO_BP_emap.png"),
       width = 12, height = 10, dpi = 300)
```

### Gene-Concept Network

Shows which specific genes belong to which GO terms. Brown nodes are GO terms,
colored nodes are genes, and hub genes appear in multiple terms.

```{r go-cnetplot, fig.width=12, fig.height=10}
cnetplot(ego_bp, showCategory = 5, node_label = "all") +
  labs(title = "Gene-Concept Network: Top 5 GO Terms")

ggsave(here(figures_dir, "GO_BP_cnetplot.png"),
       width = 12, height = 10, dpi = 300)
```

### Simplify Redundant Terms

GO is hierarchical and many terms overlap. Simplification removes redundant
terms while keeping the most significant representative from each cluster.

```{r go-simplify}
# Simplify GO results by removing redundant terms
ego_bp_simple <- clusterProfiler::simplify(ego_bp,
                          cutoff = 0.7,      # Similarity threshold
                          by = "p.adjust",   # Keep most significant
                          select_fun = min)

cat("Original GO terms:", nrow(ego_bp), "\n")
cat("After simplification:", nrow(ego_bp_simple), "\n\n")

# Compare
cat("Top 10 simplified terms:\n")
head(ego_bp_simple, 10)
```

```{r go-simplified-dotplot, fig.width=10, fig.height=6}
dotplot(ego_bp_simple,
        showCategory = 20,
        title        = "Simplified GO-BP Enrichment")

ggsave(here(figures_dir, "GO_BP_simplified_dotplot.png"),
       width = 10, height = 6, dpi = 300)
```

---

## KEGG Pathway Enrichment

KEGG provides curated pathway maps for metabolism, signaling, and disease.
It complements GO by providing a different, more focused view of the same
biology. Common organism codes: `mmu` = mouse, `hsa` = human.

```{r enrichKEGG}
# KEGG uses organism codes: "mmu" = mouse, "hsa" = human
kegg_result <- enrichKEGG(gene = gene_entrez$ENTREZID,
                          universe = universe,
                          organism = "mmu",      # Mouse
                          pAdjustMethod = "BH",
                          pvalueCutoff = 0.05,
                          qvalueCutoff = 0.05)

# Add gene symbols (KEGG uses Entrez IDs)
kegg_result <- setReadable(kegg_result, 
                           OrgDb = org.Mm.eg.db, 
                           keyType = "ENTREZID")

# Summary
cat("Enriched KEGG pathways:", nrow(kegg_result), "\n\n")

# Show results
head(kegg_result, 15)
```

```{r kegg-dotplot, fig.width=10, fig.height=6}
# Dot plot
dotplot(kegg_result, 
        showCategory = 20,
        title = "KEGG Pathway Enrichment")

ggsave(here("demo-analysis", "output", "differential-expression", "figures", "KEGG_dotplot.png"), width = 10, height = 6, dpi = 300)
```

### GO vs KEGG Comparison

```{r go-vs-kegg, fig.width=10, fig.height=8}
ego_df <- as.data.frame(ego_bp) %>%
  mutate(Database = "GO-BP", Term = Description) %>%
  select(Database, Term, Count, FDR = p.adjust) %>%
  head(10)

kegg_df <- as.data.frame(kegg_result) %>%
  mutate(Database = "KEGG", Term = Description) %>%
  select(Database, Term, Count, FDR = p.adjust) %>%
  head(10)

bind_rows(ego_df, kegg_df) %>%
  ggplot(aes(x = Count, y = reorder(Term, Count), fill = Database)) +
  geom_col() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Top Enriched Terms: GO vs KEGG",
       x = "Number of Genes", y = "Term") +
  theme_minimal() +
  facet_wrap(~Database, scales = "free_y", ncol = 1)

ggsave(here(figures_dir, "GO_vs_KEGG_comparison.png"),
       width = 10, height = 8, dpi = 300)
```

---

## Upregulated vs Downregulated Genes

Analyzing genes by direction of change often reveals more specific biological
themes than treating all DE genes together. For example, upregulated genes may
reflect stress responses while downregulated genes reflect disrupted normal
functions.

```{r up-vs-down}
up_entrez <- gene_entrez %>%
  filter(SYMBOL %in% rownames(up_genes)) %>%
  pull(ENTREZID)

down_entrez <- gene_entrez %>%
  filter(SYMBOL %in% rownames(down_genes)) %>%
  pull(ENTREZID)

ego_up <- enrichGO(gene         = up_entrez,
                   universe     = universe,
                   OrgDb        = org.Mm.eg.db,
                   ont          = "BP",
                   pvalueCutoff = 0.05,
                   readable     = TRUE)

ego_down <- enrichGO(gene         = down_entrez,
                     universe     = universe,
                     OrgDb        = org.Mm.eg.db,
                     ont          = "BP",
                     pvalueCutoff = 0.05,
                     readable     = TRUE)

cat("GO terms — upregulated:  ", nrow(ego_up), "\n")
cat("GO terms — downregulated:", nrow(ego_down), "\n")
```

```{r up-vs-down-plot, fig.width=12, fig.height=8}
p1 <- dotplot(ego_up,   showCategory = 10, title = "Upregulated in KD") +
      theme(axis.text.y = element_text(size = 8))

p2 <- dotplot(ego_down, showCategory = 10, title = "Downregulated in KD") +
      theme(axis.text.y = element_text(size = 8))

p1 | p2

ggsave(here(figures_dir, "GO_up_vs_down.png"),
       width = 12, height = 8, dpi = 300)
```

---

## Export Results

```{r export}
write.csv(as.data.frame(ego_bp),
          here(results_dir, "GO_BP_enrichment.csv"),           row.names = FALSE)
write.csv(as.data.frame(ego_bp_simple),
          here(results_dir, "GO_BP_enrichment_simplified.csv"), row.names = FALSE)
write.csv(as.data.frame(kegg_result),
          here(results_dir, "KEGG_enrichment.csv"),             row.names = FALSE)
write.csv(as.data.frame(ego_up),
          here(results_dir, "GO_BP_upregulated.csv"),           row.names = FALSE)
write.csv(as.data.frame(ego_down),
          here(results_dir, "GO_BP_downregulated.csv"),         row.names = FALSE)

cat("✓ GO_BP_enrichment.csv\n")
cat("✓ GO_BP_enrichment_simplified.csv\n")
cat("✓ KEGG_enrichment.csv\n")
cat("✓ GO_BP_upregulated.csv\n")
cat("✓ GO_BP_downregulated.csv\n")
```

---

## Session Information

```{r sessioninfo}
sessionInfo()
```
