---
title: "Part 3: Pathway Analysis"
author: "UFHCI BCB-SR Bioinformatics Unit"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

Having identified differentially expressed genes, we now ask what they mean
biologically. Pathway analysis uses over-representation testing to ask: are
our DE genes enriched in particular biological processes or pathways compared
to what we would expect by chance?

**Requires:** `de_results_all.csv` produced by `03_differential_expression.Rmd`

---

## Setup

```{r libraries}
library(clusterProfiler)
library(org.Mm.eg.db)   # Mouse annotation — use org.Hs.eg.db for human
library(enrichplot)
library(tidyverse)
library(patchwork)
library(here)

set.seed(42)

results_dir <- here("demo-analysis", "output", "differential-expression", "results")
figures_dir <- here("demo-analysis", "output", "differential-expression", "figures")

dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(figures_dir, showWarnings = FALSE, recursive = TRUE)
```

---

## Load DE Results

```{r load-results}
all_results <- read.csv(here(results_dir, "de_results_all.csv"))
rownames(all_results) <- all_results$gene

sig_genes  <- all_results %>% filter(adj.P.Val < 0.05 & abs(logFC) > 1)
up_genes   <- sig_genes  %>% filter(logFC > 1)
down_genes <- sig_genes  %>% filter(logFC < 1)

cat("Total DE genes (FDR < 0.05) & FC > 2x:", nrow(sig_genes), "\n")
cat("Upregulated in KD:          ", nrow(up_genes), "\n")
cat("Downregulated in KD:        ", nrow(down_genes), "\n")
```

---

## Convert Gene IDs

Most pathway databases use Entrez Gene IDs rather than gene symbols. We convert
here and also define a background gene set — all genes that were actually tested.
Using the correct background is important for accurate enrichment statistics:
comparing to all genes in the genome rather than genes in your experiment
inflates false positives.

A conversion rate of 85–95% is normal. Genes that fail to convert are typically
novel genes, pseudogenes, or non-coding RNAs not yet in the annotation database.

```{r convert-ids}
# Convert DE gene symbols to Entrez IDs
gene_entrez <- bitr(rownames(sig_genes),
                    fromType = "SYMBOL",
                    toType   = "ENTREZID",
                    OrgDb    = org.Mm.eg.db)

# Background: all genes tested
background_entrez <- bitr(rownames(all_results),
                           fromType = "SYMBOL",
                           toType   = "ENTREZID",
                           OrgDb    = org.Mm.eg.db)

universe <- background_entrez$ENTREZID

cat("DE genes submitted:     ", nrow(sig_genes), "\n")
cat("Successfully converted: ", nrow(gene_entrez), "\n")
cat("Background genes:       ", length(universe), "\n")
```

---

## Upregulated vs Downregulated Genes

Analyzing genes by direction of change often reveals more specific biological
themes than treating all DE genes together. For example, upregulated genes may
reflect stress responses while downregulated genes reflect disrupted normal
functions.

```{r up-vs-down}
up_entrez <- gene_entrez %>%
  filter(SYMBOL %in% rownames(up_genes)) %>%
  pull(ENTREZID)

down_entrez <- gene_entrez %>%
  filter(SYMBOL %in% rownames(down_genes)) %>%
  pull(ENTREZID)

```

## GO Enrichment — Biological Process

Gene Ontology (GO) Biological Process terms describe what genes do at the
cellular and systems level (e.g., "DNA repair", "cell cycle arrest"). Key
output columns:

- **GeneRatio**: Fraction of your DE genes annotated to this term
- **BgRatio**: Fraction of background genes annotated to this term
- **p.adjust**: FDR-corrected p-value
- **geneID**: Gene symbols driving the enrichment

```{r go-bp}
ego_up <- enrichGO(gene         = up_entrez,
                   universe     = universe,
                   OrgDb        = org.Mm.eg.db,
                   ont          = "BP",
                   pvalueCutoff = 0.05,
                   readable     = TRUE)

ego_down <- enrichGO(gene         = down_entrez,
                     universe     = universe,
                     OrgDb        = org.Mm.eg.db,
                     ont          = "BP",
                     pvalueCutoff = 0.05,
                     readable     = TRUE)

cat("GO terms — upregulated:  ", nrow(ego_up), "\n")
cat("GO terms — downregulated:", nrow(ego_down), "\n")
```

### Dot Plot

The dot plot shows the top enriched terms. X-axis is the gene ratio, dot size
is the number of genes, and dot color is the adjusted p-value.

```{r up-vs-down-plot, fig.width=12, fig.height=8}
p1 <- dotplot(ego_up,   showCategory = 10, title = "Upregulated in KD") +
      theme(axis.text.y = element_text(size = 8))

p2 <- dotplot(ego_down, showCategory = 10, title = "Downregulated in KD") +
      theme(axis.text.y = element_text(size = 8))

p1 | p2

ggsave(here(figures_dir, "GO_up_vs_down.png"),
       width = 12, height = 8, dpi = 300)
```


### Gene-Concept Network

Shows which specific genes belong to which GO terms. Brown nodes are GO terms,
colored nodes are genes, and hub genes appear in multiple terms.

Gene-concept networks show which genes belong to which terms. We can color the genes by their fold-change to see expression patterns.
```{r prepare-foldchange-vectors}
# Extract fold changes for coloring
# Need a named vector: names = gene symbols, values = logFC

# For upregulated genes
up_gene_symbols <- rownames(up_genes)
up_foldchanges <- up_genes$logFC
names(up_foldchanges) <- up_gene_symbols

# For downregulated genes
down_gene_symbols <- rownames(down_genes)
down_foldchanges <- down_genes$logFC
names(down_foldchanges) <- down_gene_symbols

# Check the format
head(up_foldchanges)
head(down_foldchanges)

# Check ranges
cat("Upregulated logFC range:", 
    round(min(up_foldchanges), 2), "to", 
    round(max(up_foldchanges), 2), "\n")
cat("Downregulated logFC range:", 
    round(min(down_foldchanges), 2), "to", 
    round(max(down_foldchanges), 2), "\n")
```

```{r go-cnetplot, fig.width=12, fig.height=10}
p_up <- cnetplot(ego_up, 
                 showCategory = 5, 
                 foldChange = up_foldchanges, 
                 node_label = "all", 
                 cex_label_gene = 0.7, 
                 cex_label_category = 1.2, 
                 colorEdge = TRUE) + 
  scale_color_gradient(low = "white",
                       high = "red",
                       name = "Log2FC\n(Upregulated)") +
  labs(title = "Gene-Concept Network: Upregulated DEGS - Top 5 GO Terms") + 
  theme(legend.position = "right")

# Display the plot
print(p_up)

# Save the plot
ggsave(here(figures_dir, "GO_up_cnetplot.png"), plot = p_up, 
       width = 12, height = 10, dpi = 300)

p_down <- cnetplot(ego_down, 
                   showCategory = 5, 
                   foldChange = down_foldchanges, 
                   node_label = "all", 
                   cex_label_gene = 0.7, 
                   cex_label_category = 1.2, 
                   colorEdge = TRUE) + 
  scale_color_gradient(low = "white",
                       high = "blue",
                       name = "Log2FC\n(Downregulated)") +
  labs(title = "Gene-Concept Network: Downregulated DEGs - Top 5 GO Terms") + 
  theme(legend.position = "right")

# Display the plot
print(p_down)

# Save the plot
ggsave(here(figures_dir, "GO_down_cnetplot.png"), plot = p_down,
       width = 12, height = 10, dpi = 300)
```

### Simplify Redundant Terms

GO is hierarchical and many terms overlap. Simplification removes redundant
terms while keeping the most significant representative from each cluster.

```{r go-simplify}
# Simplify GO results by removing redundant terms
ego_up_simple <- clusterProfiler::simplify(ego_up,
                          cutoff = 0.7,      # Similarity threshold
                          by = "p.adjust",   # Keep most significant
                          select_fun = min)

cat("Original GO terms:", nrow(ego_up), "\n")
cat("After simplification:", nrow(ego_up_simple), "\n\n")

# Compare
cat("Top 10 simplified terms:\n")
head(ego_up_simple, 10)
```

```{r go-simplified-dotplot, fig.width=10, fig.height=6}
dotplot(ego_up_simple,
        showCategory = 10,
        title        = "Simplified GO-BP Enrichment")

ggsave(here(figures_dir, "GO_BP_simplified_dotplot.png"),
       width = 10, height = 6, dpi = 300)
```

---

## KEGG Pathway Enrichment

KEGG provides curated pathway maps for metabolism, signaling, and disease.
It complements GO by providing a different, more focused view of the same
biology. Common organism codes: `mmu` = mouse, `hsa` = human.

```{r enrichKEGG}
# KEGG uses organism codes: "mmu" = mouse, "hsa" = human
kegg_up <- enrichKEGG(gene = up_entrez,
                      universe = universe, 
                      organism = "mmu",      # Mouse
                      pAdjustMethod = "BH",
                      pvalueCutoff = 0.05,
                      qvalueCutoff = 0.05)

kegg_down <- enrichKEGG(gene = down_entrez,
                        universe = universe, 
                        organism = "mmu",      # Mouse
                        pAdjustMethod = "BH",
                        pvalueCutoff = 0.05,
                        qvalueCutoff = 0.05)

# Add gene symbols (KEGG uses Entrez IDs)
kegg_up <- setReadable(kegg_up, 
                       OrgDb = org.Mm.eg.db, 
                       keyType = "ENTREZID")

kegg_down <- setReadable(kegg_down, 
                         OrgDb = org.Mm.eg.db, 
                         keyType = "ENTREZID")

# Clean pathway names - remove " - Mus musculus" suffix
kegg_up@result$Description <- gsub(" - .*$", "", kegg_up@result$Description)
kegg_down@result$Description <- gsub(" - .*$", "", kegg_down@result$Description)

# Summary
cat("Enriched KEGG pathways - Upregulated DEGs: ", nrow(kegg_up), "\n")
cat("Enriched KEGG pathways - Downregulated DEGs: ", nrow(kegg_down), "\n\n")

# Show results
head(kegg_up, 10)
```

```{r kegg-dotplot, fig.width=10, fig.height=6}
# Dot plot
dotplot(kegg_up, 
        showCategory = 15,
        title = "KEGG Pathway Enrichment")

ggsave(here("demo-analysis", "output", "differential-expression", "figures", "KEGG_dotplot.png"), width = 10, height = 6, dpi = 300)
```

### GO vs KEGG Comparison (Up and Down Separately)

```{r compare-databases-direction, fig.width=12, fig.height=10}
# Prepare GO data (up and down)
ego_up_df <- as.data.frame(ego_up) %>%
  mutate(Database = "GO-BP", 
         Direction = "Up",
         Term = Description) %>%
  dplyr::select(Database, Direction, Term, Count, FDR = p.adjust) %>%
  head(10)

ego_down_df <- as.data.frame(ego_down) %>%
  mutate(Database = "GO-BP", 
         Direction = "Down",
         Term = Description) %>%
  dplyr::select(Database, Direction, Term, Count, FDR = p.adjust) %>%
  head(10)

# Prepare KEGG data (up and down)
kegg_up_df <- as.data.frame(kegg_up) %>%
  mutate(Database = "KEGG", 
         Direction = "Up",
         Term = Description) %>%
  dplyr::select(Database, Direction, Term, Count, FDR = p.adjust) %>%
  head(10)

kegg_down_df <- as.data.frame(kegg_down) %>%
  mutate(Database = "KEGG", 
         Direction = "Down",
         Term = Description) %>%
  dplyr::select(Database, Direction, Term, Count, FDR = p.adjust) %>%
  head(10)

# Combine all
comparison_df <- bind_rows(ego_up_df, ego_down_df, 
                           kegg_up_df, kegg_down_df)

# Plot: facet by database, color by direction
ggplot(comparison_df, aes(x = Count, y = reorder(Term, Count), fill = Direction)) +
  geom_col() +
  scale_fill_manual(values = c("Up" = "red", "Down" = "blue")) +
  labs(title = "Top Enriched Terms: GO vs KEGG (Up and Down)",
       x = "Number of Genes", 
       y = "Term",
       fill = "Direction") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 8)) +
  facet_wrap(~Database, scales = "free_y", ncol = 1)

ggsave(here(figures_dir, "GO_vs_KEGG_comparison_direction.png"), 
       width = 12, height = 10, dpi = 300)
```

---

## Export Results

```{r export-pathway-results-final}
# ===== Export GO-BP Results =====
write.csv(as.data.frame(ego_up),
          here(results_dir, "GO_BP_upregulated.csv"),
          row.names = FALSE)

write.csv(as.data.frame(ego_down),
          here(results_dir, "GO_BP_downregulated.csv"),
          row.names = FALSE)

# ===== Export KEGG Results =====
write.csv(as.data.frame(kegg_up),
          here(results_dir, "KEGG_upregulated.csv"),
          row.names = FALSE)

write.csv(as.data.frame(kegg_down),
          here(results_dir, "KEGG_downregulated.csv"),
          row.names = FALSE)

# ===== Create Summary Table =====
pathway_summary <- data.frame(
  Database = rep(c("GO-BP", "KEGG"), each = 2),
  Direction = rep(c("Up", "Down"), 2),
  N_Terms = c(nrow(ego_up), nrow(ego_down), 
              nrow(kegg_up), nrow(kegg_down)),
  Top_Term = c(
    ego_up$Description[1],
    ego_down$Description[1],
    kegg_up$Description[1],
    kegg_down$Description[1]
  ),
  Top_FDR = formatC(c(
    ego_up$p.adjust[1],
    ego_down$p.adjust[1],
    kegg_up$p.adjust[1],
    kegg_down$p.adjust[1]
  ), format = "e", digits = 2)
)

write.csv(pathway_summary,
          here(results_dir, "pathway_summary.csv"),
          row.names = FALSE)

# ===== Print Summary =====
cat("\n=== Pathway Enrichment Results Exported ===\n\n")
print(pathway_summary)

cat("\nFiles saved:\n")
cat("  ✓ GO_BP_upregulated.csv (", nrow(ego_up), " terms)\n", sep = "")
cat("  ✓ GO_BP_downregulated.csv (", nrow(ego_down), " terms)\n", sep = "")
cat("  ✓ KEGG_upregulated.csv (", nrow(kegg_up), " terms)\n", sep = "")
cat("  ✓ KEGG_downregulated.csv (", nrow(kegg_down), " terms)\n", sep = "")
cat("  ✓ pathway_summary.csv\n")
cat("\nLocation:", results_dir, "\n")
```

---

## Session Information

```{r sessioninfo}
sessionInfo()
```
