---
title: "Part 3: Pathway Analysis"
subtitle: "UF Health Cancer Institute Bioinformatics Workshop - Differential Expression Analysis"
author: "UFHCI BCB-SR Bioinformatics Unit"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: false
    code-tools: true
    code-overflow: wrap
    embed-resources: true
    theme: cosmo
    df-print: paged
execute:
  warning: false
  message: false
---

# Introduction

We've identified thousands of differentially expressed genes — but what does
this mean biologically? Pathway analysis helps us answer:

- What biological processes are affected?
- What pathways are dysregulated?
- Are there common themes among DE genes?

**In this notebook you will:**

- Convert gene identifiers for use with pathway databases
- Run GO (Gene Ontology) over-representation analysis
- Run KEGG pathway enrichment
- Visualize and simplify results
- Analyze upregulated and downregulated genes separately
- Export results tables and figures

**Requires:** `de_results_all.csv` produced by `03_differential_expression.qmd/.Rmd`

---

# Setup

```{r setup}
library(clusterProfiler)
library(org.Mm.eg.db)   # Mouse annotation — use org.Hs.eg.db for human
library(enrichplot)
library(tidyverse)
library(patchwork)
library(here)

set.seed(42)

results_dir <- here("demo-analysis", "output", "02-differential-expression", "results")
figures_dir <- here("demo-analysis", "output", "02-differential-expression", "figures")

dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(figures_dir, showWarnings = FALSE, recursive = TRUE)
```

::: {.callout-tip}
## Package Installation
```r
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(c("clusterProfiler", "org.Mm.eg.db", "enrichplot"))
install.packages("patchwork")
```
:::

---

# Load DE Results

```{r load-results}
all_results <- read.csv(here(results_dir, "de_results_all.csv"),
                        row.names = 1)

sig_genes  <- all_results %>% filter(adj.P.Val < 0.05)
up_genes   <- sig_genes  %>% filter(logFC > 0)
down_genes <- sig_genes  %>% filter(logFC < 0)

cat("Total DE genes (FDR < 0.05):", nrow(sig_genes), "\n")
cat("Upregulated in KD:           ", nrow(up_genes), "\n")
cat("Downregulated in KD:         ", nrow(down_genes), "\n")
```

---

# Convert Gene IDs

Most pathway databases use Entrez Gene IDs rather than gene symbols.
We convert here and also define a background gene set — the set of all
genes that were tested. Using the correct background is important for
accurate enrichment statistics.

```{r convert-ids}
# Convert DE genes
gene_entrez <- bitr(rownames(sig_genes),
                    fromType = "SYMBOL",
                    toType   = "ENTREZID",
                    OrgDb    = org.Mm.eg.db)

cat("DE genes submitted:     ", nrow(sig_genes), "\n")
cat("Successfully converted: ", nrow(gene_entrez), "\n")

# Background: all genes tested
background_entrez <- bitr(rownames(all_results),
                          fromType = "SYMBOL",
                          toType   = "ENTREZID",
                          OrgDb    = org.Mm.eg.db)

universe <- background_entrez$ENTREZID
cat("Background genes:       ", length(universe), "\n")
```

::: {.callout-note}
## Why Background Matters
Using all genes in the genome as background inflates enrichment. Always
use the set of genes that were actually tested in your experiment.
:::

::: {.callout-warning}
## Conversion Rate
A conversion rate of 85–95% is normal. Genes that fail to convert are
typically novel genes, pseudogenes, or non-coding RNAs not yet in the
annotation database.
:::

---

# GO Enrichment — Biological Process

Gene Ontology (GO) Biological Process terms describe what genes do at
the cellular and systems level (e.g., "DNA repair", "cell cycle arrest").

```{r go-bp}
ego_bp <- enrichGO(gene          = gene_entrez$ENTREZID,
                   universe      = universe,
                   OrgDb         = org.Mm.eg.db,
                   ont           = "BP",
                   pAdjustMethod = "BH",
                   pvalueCutoff  = 0.05,
                   qvalueCutoff  = 0.05,
                   readable      = TRUE)

cat("Enriched GO-BP terms:", nrow(ego_bp), "\n\n")
head(ego_bp, 10)
```

::: {.callout-note}
## Reading the Results Table
- **GeneRatio**: Fraction of your DE genes in this term
- **BgRatio**: Fraction of background genes in this term
- **p.adjust**: FDR-corrected p-value
- **geneID**: Gene symbols driving the enrichment
:::

## Dot Plot

```{r go-dotplot, fig.width=10, fig.height=8}
dotplot(ego_bp,
        showCategory = 20,
        title        = "GO Biological Process Enrichment",
        font.size    = 10)

ggsave(here(figures_dir, "GO_BP_dotplot.png"),
       width = 10, height = 8, dpi = 300)
```

::: {.callout-tip}
## Reading a Dot Plot
- **X-axis**: Gene ratio (proportion of DE genes in the term)
- **Dot size**: Number of genes
- **Dot color**: Adjusted p-value (darker = more significant)
:::

## Enrichment Map

Shows relationships between enriched terms — terms sharing many genes
are connected and tend to cluster around common biological themes.

```{r go-emap, fig.width=12, fig.height=10}
ego_bp_sim <- pairwise_termsim(ego_bp)

emapplot(ego_bp_sim,
         showCategory = 30,
         layout       = "kk") +
  labs(title = "GO Enrichment Map")

ggsave(here(figures_dir, "GO_BP_emap.png"),
       width = 12, height = 10, dpi = 300)
```

## Gene-Concept Network

Shows which specific genes belong to which GO terms.

```{r go-cnetplot, fig.width=12, fig.height=10}
cnetplot(ego_bp,
         showCategory = 5,
         node_label   = "all") +
  labs(title = "Gene-Concept Network: Top 5 GO Terms")

ggsave(here(figures_dir, "GO_BP_cnetplot.png"),
       width = 12, height = 10, dpi = 300)
```

::: {.callout-note}
## Gene-Concept Networks
- **Brown nodes**: GO terms
- **Colored nodes**: Genes
- **Edges**: Gene belongs to term
- **Hub genes**: Genes appearing in multiple terms
:::

## Simplify Redundant Terms

GO is hierarchical and many terms overlap. Simplification removes
redundant terms while keeping the most significant representative.

```{r go-simplify}
ego_bp_simple <- simplify(ego_bp,
                           cutoff     = 0.7,
                           by         = "p.adjust",
                           select_fun = min)

cat("Original terms:   ", nrow(ego_bp), "\n")
cat("Simplified terms: ", nrow(ego_bp_simple), "\n")
```

```{r go-simplified-dotplot, fig.width=10, fig.height=6}
dotplot(ego_bp_simple,
        showCategory = 20,
        title        = "Simplified GO-BP Enrichment")

ggsave(here(figures_dir, "GO_BP_simplified_dotplot.png"),
       width = 10, height = 6, dpi = 300)
```

---

# KEGG Pathway Enrichment

KEGG provides curated pathway maps for metabolism, signaling, and disease.
It complements GO by giving a different view of the same biology.

```{r kegg}
kegg_result <- enrichKEGG(gene          = gene_entrez$ENTREZID,
                           universe      = universe,
                           organism      = "mmu",
                           pAdjustMethod = "BH",
                           pvalueCutoff  = 0.05,
                           qvalueCutoff  = 0.05)

kegg_result <- setReadable(kegg_result,
                            OrgDb   = org.Mm.eg.db,
                            keyType = "ENTREZID")

cat("Enriched KEGG pathways:", nrow(kegg_result), "\n\n")
head(kegg_result, 15)
```

::: {.callout-tip}
## KEGG Organism Codes
`mmu` = mouse, `hsa` = human, `rno` = rat, `dme` = fly, `dre` = zebrafish
:::

```{r kegg-dotplot, fig.width=10, fig.height=6}
dotplot(kegg_result,
        showCategory = 20,
        title        = "KEGG Pathway Enrichment")

ggsave(here(figures_dir, "KEGG_dotplot.png"),
       width = 10, height = 6, dpi = 300)
```

## GO vs KEGG Comparison

```{r go-vs-kegg, fig.width=10, fig.height=8}
ego_df <- as.data.frame(ego_bp) %>%
  mutate(Database = "GO-BP", Term = Description) %>%
  select(Database, Term, Count, FDR = p.adjust) %>%
  head(10)

kegg_df <- as.data.frame(kegg_result) %>%
  mutate(Database = "KEGG", Term = Description) %>%
  select(Database, Term, Count, FDR = p.adjust) %>%
  head(10)

bind_rows(ego_df, kegg_df) %>%
  ggplot(aes(x = Count, y = reorder(Term, Count), fill = Database)) +
  geom_col() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Top Enriched Terms: GO vs KEGG",
       x     = "Number of Genes",
       y     = "Term") +
  theme_minimal() +
  facet_wrap(~Database, scales = "free_y", ncol = 1)

ggsave(here(figures_dir, "GO_vs_KEGG_comparison.png"),
       width = 10, height = 8, dpi = 300)
```

::: {.callout-note}
## GO vs KEGG
GO is broader and more comprehensive. KEGG is more focused on curated
signaling and metabolic pathways with visual pathway diagrams. Run both
— they complement each other.
:::

---

# Upregulated vs Downregulated Genes

Analyzing genes by direction of change often reveals more specific
biological themes than treating all DE genes together.

```{r up-vs-down}
up_entrez <- gene_entrez %>%
  filter(SYMBOL %in% rownames(up_genes)) %>%
  pull(ENTREZID)

down_entrez <- gene_entrez %>%
  filter(SYMBOL %in% rownames(down_genes)) %>%
  pull(ENTREZID)

ego_up <- enrichGO(gene          = up_entrez,
                   universe      = universe,
                   OrgDb         = org.Mm.eg.db,
                   ont           = "BP",
                   pvalueCutoff  = 0.05,
                   readable      = TRUE)

ego_down <- enrichGO(gene          = down_entrez,
                     universe      = universe,
                     OrgDb         = org.Mm.eg.db,
                     ont           = "BP",
                     pvalueCutoff  = 0.05,
                     readable      = TRUE)

cat("GO terms enriched in upregulated genes:  ", nrow(ego_up), "\n")
cat("GO terms enriched in downregulated genes:", nrow(ego_down), "\n")
```

```{r up-vs-down-plot, fig.width=12, fig.height=8}
p1 <- dotplot(ego_up,   showCategory = 10,
              title = "Upregulated in KD") +
      theme(axis.text.y = element_text(size = 8))

p2 <- dotplot(ego_down, showCategory = 10,
              title = "Downregulated in KD") +
      theme(axis.text.y = element_text(size = 8))

p1 | p2

ggsave(here(figures_dir, "GO_up_vs_down.png"),
       width = 12, height = 8, dpi = 300)
```

---

# Export Results

```{r export}
write.csv(as.data.frame(ego_bp),
          here(results_dir, "GO_BP_enrichment.csv"),
          row.names = FALSE)

write.csv(as.data.frame(ego_bp_simple),
          here(results_dir, "GO_BP_enrichment_simplified.csv"),
          row.names = FALSE)

write.csv(as.data.frame(kegg_result),
          here(results_dir, "KEGG_enrichment.csv"),
          row.names = FALSE)

write.csv(as.data.frame(ego_up),
          here(results_dir, "GO_BP_upregulated.csv"),
          row.names = FALSE)

write.csv(as.data.frame(ego_down),
          here(results_dir, "GO_BP_downregulated.csv"),
          row.names = FALSE)

cat("✓ GO_BP_enrichment.csv\n")
cat("✓ GO_BP_enrichment_simplified.csv\n")
cat("✓ KEGG_enrichment.csv\n")
cat("✓ GO_BP_upregulated.csv\n")
cat("✓ GO_BP_downregulated.csv\n")
```

---

# Session Information

```{r sessioninfo}
sessionInfo()
```
