---
title: "RNA-seq Differential Expression Analysis with limma-voom"
subtitle: "UF Health Cancer Institute Bioinformatics Workshop - Session 2"
author: "UFHCI BCB-SR Bioinformatics Unit"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: false
    code-tools: true
    code-overflow: wrap
    embed-resources: true
    theme: cosmo
    df-print: paged
execute:
  warning: false
  message: false
---

# Introduction

In this session, we'll work through a complete differential expression (DE) analysis using RNA-seq data. We'll be using the output from the nf-core/rnaseq pipeline (STAR alignment + RSEM quantification) that was configured in Session 1.

**Learning Objectives:**

- Load and prepare RNA-seq count data from RSEM
- Perform quality control and exploratory data analysis
- Understand filtering, normalization, and the limma-voom workflow
- Conduct differential expression analysis for simple and complex designs
- Create publication-quality visualizations
- Interpret and export results

## The Analysis Workflow

```{mermaid}
%%| fig-width: 8
flowchart TD
    A[RSEM Gene Counts] --> B[Create DGEList Object]
    B --> C[Filter Low Expression]
    C --> D[TMM Normalization]
    D --> E[Quality Control & Visualization]
    E --> F[Design Matrix]
    F --> G[voom Transformation + Quality Weights]
    G --> H[Linear Model Fitting]
    H --> I[Empirical Bayes]
    I --> J[Extract Results]
    J --> K[Visualization & Interpretation]
```

# Setup

## Load Required Libraries

```{r setup}
# Load required packages
library(limma)       # Differential expression analysis
library(edgeR)       # RNA-seq data structures and filtering
library(tidyverse)   # Data manipulation and visualization
library(ggrepel)     # Better text labels on plots
library(pheatmap)    # Heatmaps
library(RColorBrewer) # Color palettes
library(here)

# Set a seed for reproducibility
set.seed(42)

# Define directory structure using here() for portability
# here() finds the project root automatically
data_dir <- here("demo-analysis", "output", "01-prepared-data")
results_dir <- here("demo-analysis", "output", "differential-expression", "results")
figures_dir <- here("demo-analysis", "output", "differential-expression", "figures")

# Create output directories if they don't exist
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(figures_dir, showWarnings = FALSE, recursive = TRUE)

# Display the project root for verification
cat("Project root:", here(), "\n")
cat("Data directory:", data_dir, "\n")
cat("Results directory:", results_dir, "\n")
cat("Figures directory:", figures_dir, "\n")
```

::: {.callout-tip}
## Package Installation
If you haven't installed these packages yet, you can do so with:
```r
install.packages(c("tidyverse", "ggrepel", "pheatmap", "RColorBrewer"))
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(c("limma", "edgeR"))
```
:::

## Set File Paths

```{r paths}
# Define paths to your data files
# These should be outputs from the nf-core/rnaseq pipeline
counts_file <- here("demo-analysis", "output", "01-prepared-data", "rsem.merged.gene_counts.tsv")
sample_info_file <- here("demo-analysis", "data", "metadata", "sample_metadata.csv")

# Verify files exist
if (!file.exists(counts_file)) {
  stop("Count file not found: ", counts_file, "\n",
       "Please ensure your data is in the 'data/' directory")
}

if (!file.exists(sample_info_file)) {
  stop("Sample info file not found: ", sample_info_file, "\n",
       "Please ensure your metadata is in the 'data/' directory")
}

cat("✓ Count file found:", counts_file, "\n")
cat("✓ Sample info found:", sample_info_file, "\n")
```

::: {.callout-note}
## nf-core Output Structure
The nf-core/rnaseq pipeline with RSEM produces a merged count matrix file typically named:
`rsem.merged.gene_counts.tsv` in the `star_rsem/` results directory. This file contains raw counts for all samples across all genes.
:::

# Part 1: Data Loading and Preparation

## Load Gene Count Data

```{r load-counts}
# Load the RSEM gene counts
# The nf-core pipeline produces a merged file with:
# - gene_id: Ensembl gene ID
# - gene_name: Gene symbol
# - One column per sample with count data

expression_data <- read.table(counts_file, 
                              header = TRUE, 
                              row.names = 1,
                              check.names = FALSE)

# Quick look at the data
head(expression_data, n = 3)
dim(expression_data)
```

::: {.callout-important}
## RSEM vs Salmon
Note that RSEM produces **expected counts** which are suitable for downstream analysis with limma-voom or DESeq2. These are not the same as TPM values - make sure you're using the `.genes.results` counts, not TPM values!
:::

## Load Sample Metadata

```{r load-metadata}
# Load sample information
# This file should contain:
# - sample: sample names (matching column names in expression_data)
# - Any experimental factors (treatment, genotype, time, etc.)

sample_info <- read.csv(sample_info_file, 
                          header = TRUE, 
                          stringsAsFactors = FALSE)

# Display the sample information
sample_info

# Quick check: do sample names match?
all(sample_info$sample %in% colnames(expression_data))
```

::: {.callout-tip}
## Sample Metadata Best Practices
Your sample metadata file should include:
- Clear, consistent sample names matching your count data
- All experimental factors (treatment, batch, individual, etc.)
- Any known technical variables (sequencing run, library prep date)
- Keep a data dictionary explaining what each column means!
:::

# Part 2: Quality Control and Exploratory Analysis

## Create DGEList Object

The `DGEList` object from edgeR is a convenient container for RNA-seq data that bundles counts, sample information, and gene annotations together.

```{r create-dge}
# Create DGEList object
# This combines count data with sample metadata
DGE <- DGEList(counts = expression_data,
               samples = sample_info)

# Explicitly add group column to DGE
DGE$samples$group <- factor(DGE$samples$genotype, levels = c("PRMT7_WT", "PRMT7_KD"))
# Inspect the object
DGE
```

## Filter Lowly Expressed Genes

Not all genes are expressed in your samples. Filtering out genes with very low or zero counts:

- Reduces noise in the analysis
- Improves statistical power (fewer tests = less multiple testing burden)
- Speeds up computation

```{r filter-genes}
# The filterByExpr function determines which genes have 
# sufficient counts for reliable DE testing
# It uses the experimental design (group) to determine 
# the minimum number of samples that should have adequate counts

keep.exprs <- filterByExpr(DGE, group = DGE$samples$group)

# How many genes are we keeping?
table(keep.exprs)

# Filter the DGEList
DGE <- DGE[keep.exprs, , keep.lib.sizes = FALSE]

cat("After filtering, we have", nrow(DGE), "genes remaining\n")
```

::: {.callout-note}
## Understanding filterByExpr
`filterByExpr()` uses a sensible heuristic: a gene should be expressed at a minimum level in at least some proportion of samples from the smallest group. By default:
- Minimum count: 10 reads
- Minimum proportion: Depends on library sizes and design
This is more sophisticated than simple count cutoffs!
:::

## Library Size and Composition

Before normalization, let's check library sizes (total read counts per sample).

```{r library-sizes}
# Calculate library sizes
lib_sizes <- DGE$samples$lib.size

# Create a data frame for plotting
lib_size_df <- data.frame(
  sample = rownames(DGE$samples),
  lib_size = lib_sizes,
  group = DGE$samples$group
)

# Plot library sizes
ggplot(lib_size_df, aes(x = sample, y = lib_size / 1e6, fill = group)) +
  geom_col() +
  labs(title = "Library Sizes After Filtering",
       x = "Sample",
       y = "Library Size (Millions of Reads)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Set2")
```

::: {.callout-warning}
## What to Look For
- **Large differences** (>2-3x) in library sizes might indicate technical issues
- **Very low libraries** (<5M reads for human/mouse) may have insufficient depth
- Some variation is normal and will be handled by normalization
:::

## TMM Normalization

Trimmed Mean of M-values (TMM) normalization accounts for:

1. **Library size differences** - some samples have more total reads
2. **Composition bias** - a few highly expressed genes can "steal" reads from other genes

```{r normalize}
# Calculate normalization factors
DGE <- calcNormFactors(DGE, method = "TMM")

# Check normalization factors
# Values close to 1 indicate little composition bias
DGE$samples
```

::: {.callout-note}
## TMM Normalization Factors
- Factors close to 1.0: little correction needed
- Factors <1: this sample has upward composition bias (effective library is smaller)
- Factors >1: this sample has downward composition bias (effective library is larger)
These factors are used internally by limma-voom; you don't manually apply them to counts.
:::

## Multidimensional Scaling (MDS) Plot

MDS is similar to PCA and shows overall similarity between samples. Samples that are similar should cluster together.

```{r mds-plot, fig.width=8, fig.height=6}
# Calculate MDS
mds <- plotMDS(DGE, top = 500, plot = FALSE, gene.selection = "common")

# Create a data frame for ggplot
mds_df <- data.frame(
  sample = rownames(DGE$samples),
  dim1 = mds$x,
  dim2 = mds$y, 
  genotype = DGE$samples$genotype, 
  treatment = DGE$samples$treatment, 
  Run = DGE$samples$Run
)

# Create MDS plot
ggplot(mds_df, aes(x = dim1, y = dim2, color = genotype, label = Run)) +
  geom_point(size = 4, alpha = 0.8) +
  geom_text_repel(size = 3, max.overlaps = 20) +
  labs(title = "MDS Plot - Sample Similarity",
       subtitle = "Based on top 500 most variable genes",
       x = paste0("Leading logFC dim 1 (", round(mds$var.explained[1] * 100, 1), "%)"),
       y = paste0("Leading logFC dim 2 (", round(mds$var.explained[2] * 100, 1), "%)")) +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_color_brewer(palette = "Set1")

# Save the plot
ggsave(here("demo-analysis", "output", "differential-expression", "figures", "mds_plot.png"), width = 8, height = 6, dpi = 300)
```

::: {.callout-important}
## Interpreting the MDS Plot
**Good signs:**
- Samples from the same group cluster together
- Clear separation between different conditions
- No obvious outliers

**Red flags:**
- Outlier samples far from their group
- No separation between conditions (weak signal)
- Samples clustering by batch instead of biology
:::

## Sample Correlation Heatmap

A complementary view showing pairwise sample correlations.

```{r correlation-heatmap, fig.width=8, fig.height=7}
# Get log-CPM values for visualization
logcpm <- cpm(DGE, log = TRUE)

# Calculate correlation matrix
cor_matrix <- cor(logcpm, method = "spearman")

# Create annotation for samples
annotation_col <- data.frame(
  Genotype = DGE$samples$genotype,
  row.names = colnames(logcpm)
)

# Create heatmap
pheatmap(cor_matrix,
         annotation_col = annotation_col,
         annotation_row = annotation_col,
         main = "Sample Correlation Heatmap",
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(0.7, 1, length.out = 101),
         fontsize = 8,
         filename = here("demo-analysis", "output", "differential-expression", "figures", "correlation_heatmap.png"),
         width = 8,
         height = 7)

# Also display in the document
pheatmap(cor_matrix,
         annotation_col = annotation_col,
         annotation_row = annotation_col,
         main = "Sample Correlation Heatmap",
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(0.7, 1, length.out = 101),
         fontsize = 8)
```

::: {.callout-tip}
## Correlation Expectations
- **Within-group correlation**: typically 0.95-0.99
- **Between-group correlation**: depends on how different conditions are (0.85-0.98)
- **Outliers**: samples with consistently lower correlations (<0.90) to all others
:::

# Part 3: Differential Expression - Simple Two-Group Design

Let's start with a simple comparison: two conditions, like "treated vs. control" or "mutant vs. wildtype".

## Create Design Matrix

The design matrix tells limma which samples belong to which groups.

```{r design-simple}
# For our PRMT7 knockdown experiment, we're comparing KD to WT
# The first level (PRMT7_WT) will be the reference

# Make sure the factor levels are set correctly
# The first level will be the reference (denominator in fold-change)
DGE$samples$genotype <- factor(DGE$samples$genotype, 
                                levels = c("PRMT7_WT", "PRMT7_KD"))

# Create design matrix
design_simple <- model.matrix(~genotype, data = DGE$samples)

# View the design matrix
design_simple

# Column names tell us what each coefficient represents
colnames(design_simple)
```

::: {.callout-note}
## Understanding the Design Matrix
- **Intercept (column 1)**: Average expression in the reference group (PRMT7_WT)
- **genotypePRMT7_KD (column 2)**: Difference between KD and WT (log2 fold-change)

If a gene has:
- Positive coefficient: higher expression in KD
- Negative coefficient: higher expression in WT (lower in KD)
:::

## Run limma-voom with Quality Weights

The voom transformation:
1. Converts counts to log2-CPM
2. Estimates the mean-variance relationship
3. Calculates precision weights for each observation



```{r voom-simple, fig.width=8, fig.height=5}
# Run voom with quality weights
# normalize.method = "none" because we already did TMM normalization
v_simple <- voom(DGE, 
                 design = design_simple,
                 normalize.method = "none",
                 plot = TRUE)
```

::: {.callout-note}
## Interpreting the voom Plot

The plot shows the relationship between average expression and variability:

- **Red curve**: The mean-variance trend being modeled
- **Low expression** (left side): High variance (noisy measurements)
- **High expression** (right side): Lower, more stable variance

voom uses this relationship to calculate precision weights for each gene, giving more weight to reliably measured genes in the statistical analysis.
:::

::: {.callout-tip}
## voom vs. voomWithQualityWeights

We're using basic `voom()` which handles gene-level variance. If you have concerns about outlier samples or variable sample quality, you can use `voomWithQualityWeights()` instead, which also calculates sample-level weights to down-weight problematic samples.
:::

## Fit Linear Model

```{r fit-simple}
# Fit linear model to each gene
fit_simple <- lmFit(v_simple, design_simple)

# Apply empirical Bayes moderation
# This borrows information across genes to improve variance estimates
fit_simple <- eBayes(fit_simple, robust = TRUE)

# Quick summary of DE genes at FDR < 0.05
summary(decideTests(fit_simple, adjust.method = "fdr", p.value = 0.05))
```

::: {.callout-tip}
## Understanding decideTests Output
- **Down**: Genes significantly downregulated in KD vs. WT (negative logFC = lower in KD)
- **NotSig**: Genes not significantly different between genotypes
- **Up**: Genes significantly upregulated in KD vs. WT (positive logFC = higher in KD)
:::

## Extract Top Results

```{r top-results}
# Get the top differentially expressed genes
# coef = 2 refers to the treatment effect (second column of design matrix)
top_genes <- topTable(fit_simple, 
                      coef = 2,
                      adjust.method = "BH",
                      number = 20,
                      sort.by = "P")

# Display top genes
top_genes
```

## Volcano Plot

A volcano plot shows log-fold change vs. statistical significance.

```{r volcano-simple, fig.width=8, fig.height=6}
# Extract all results
all_results <- topTable(fit_simple, 
                        coef = 2,
                        adjust.method = "BH",
                        number = Inf,
                        sort.by = "none")

# Add significance categories
all_results$significant <- case_when(
  all_results$adj.P.Val < 0.05 & all_results$logFC > 1 ~ "Up",
  all_results$adj.P.Val < 0.05 & all_results$logFC < -1 ~ "Down",
  all_results$adj.P.Val < 0.05 ~ "Significant",
  TRUE ~ "Not Significant"
)

# Create volcano plot
ggplot(all_results, aes(x = logFC, y = -log10(P.Value), color = significant)) +
  geom_point(alpha = 0.5, size = 1) +
  scale_color_manual(values = c("Up" = "red", 
                                "Down" = "blue", 
                                "Significant" = "orange",
                                "Not Significant" = "grey")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", alpha = 0.5) +
  labs(title = "Volcano Plot: PRMT7_KD vs WT",
       x = "Log2 Fold Change",
       y = "-Log10 P-value",
       color = "Significance") +
  theme_minimal() +
  theme(legend.position = "right")

ggsave(here("demo-analysis", "output", "differential-expression", "figures", "volcano_plot_simple.png"), width = 8, height = 6, dpi = 300)
```

## MA Plot

An MA plot shows log-fold change vs. average expression.

```{r ma-plot, fig.width=8, fig.height=6}
# Create MA plot
ggplot(all_results, aes(x = AveExpr, y = logFC, color = significant)) +
  geom_point(alpha = 0.5, size = 1) +
  scale_color_manual(values = c("Up" = "red", 
                                "Down" = "blue", 
                                "Significant" = "orange",
                                "Not Significant" = "grey")) +
  geom_hline(yintercept = 0, linetype = "solid", alpha = 0.5) +
  labs(title = "MA Plot: PRMT7_KD vs WT",
       x = "Average Expression (log2-CPM)",
       y = "Log2 Fold Change",
       color = "Significance") +
  theme_minimal()

ggsave(here("demo-analysis", "output", "differential-expression", "figures", "ma_plot_simple.png"), width = 8, height = 6, dpi = 300)
```

# Part 3B: Alternative Approach - edgeR Exact Test

::: {.callout-note}
## Why Use edgeR Instead of limma-voom?

While limma-voom is powerful and flexible, the classic **edgeR exact test** is:

- Often used by public databases like GREIN for processing GEO datasets
- Based on the negative binomial distribution (doesn't transform counts)
- Simpler for two-group comparisons
- Can give slightly different results due to different statistical frameworks

Let's run the same analysis using edgeR's approach to compare results and understand how to reproduce GREIN-style analyses.
:::

## Understanding GREIN's Filtering Approach

Public databases like GREIN use specific filtering criteria that aren't always fully documented. Through careful comparison, we determined their actual approach:
```{r grein-filtering}
DGE <- DGEList(counts = expression_data,
               samples = sample_info)

# Explicitly add group column to DGE
DGE$samples$group <- factor(DGE$samples$genotype, levels = c("PRMT7_WT", "PRMT7_KD"))

# GREIN's actual filtering (despite documentation saying "CPM > 0")
# They use: count >= 4 in at least min_group_size samples

# Calculate minimum group size
min_group_size <- min(table(sample_info$genotype))
cat("Minimum group size:", min_group_size, "\n")

# Apply GREIN-style filtering
keep_grein <- rowSums(DGE$counts >= 4) >= min_group_size
DGE_grein <- DGE[keep_grein, , keep.lib.sizes = FALSE]

cat("Genes before filtering:", nrow(DGE), "\n")
cat("Genes after GREIN-style filtering:", nrow(DGE_grein), "\n")
cat("(GREIN reports ~15,871 genes for comparison)\n")
```

::: {.callout-warning}
## Reproducibility Lesson #1

GREIN's documentation states "CPM > 0" but their actual implementation appears to use **count >= 4**. This is a common problem in bioinformatics:

- Methods descriptions are often incomplete
- Default parameters change between software versions
- Post-hoc filters may not be documented

**Always document your exact thresholds and test reproducibility!**
:::

## TMM Normalization
```{r grein-normalize}
# Apply TMM normalization (same as before)
DGE_grein <- calcNormFactors(DGE_grein, method = "TMM")

# Check normalization factors
summary(DGE_grein$samples$norm.factors)
```

## Estimate Dispersion

edgeR models count variability using dispersion parameters. We need to estimate:

1. **Common dispersion**: Overall variability across all genes
2. **Trended dispersion**: How dispersion changes with expression level  
3. **Tagwise dispersion**: Gene-specific dispersion estimates
```{r edger-dispersion, fig.width=8, fig.height=6}
# Estimate dispersions
DGE_grein <- estimateDisp(DGE_grein, design = design_simple)

# Plot the dispersion estimates
plotBCV(DGE_grein)
```

::: {.callout-note}
## Understanding the BCV Plot

**BCV** = Biological Coefficient of Variation = sqrt(dispersion)

- **Black dots**: Gene-specific dispersion estimates
- **Blue line**: Common dispersion (average across all genes)
- **Red curve**: Trended dispersion (how dispersion changes with expression)

Typical BCV values:
- **Well-controlled experiments**: 0.1-0.2 (10-20%)
- **Biological replicates**: 0.4-0.5 (40-50%)

Higher BCV = more biological variability between replicates
:::

## Perform Exact Test

The exact test compares two groups using the negative binomial distribution.
```{r edger-exact-test}
# Perform exact test for KD vs WT
# This compares the second group (KD) to the first group (WT)
et_grein <- exactTest(DGE_grein, pair = c("PRMT7_WT", "PRMT7_KD"))

# Get summary of DE genes at FDR < 0.05
summary(decideTests(et_grein, adjust.method = "BH", p.value = 0.05))

# Extract top genes
topTags(et_grein, n = 10)
```

## The Hidden Filter: Average Expression

Here's where it gets interesting. GREIN applies an **additional, undocumented filter**:
```{r grein-hidden-filter}
# Get all results with FDR values
all_edger_results <- topTags(et_grein, n = Inf)$table

# Count DE genes with just FDR < 0.05
cat("DE genes with FDR < 0.05:", sum(all_edger_results$FDR < 0.05), "\n")

# GREIN's actual criteria: FDR < 0.05 AND logCPM > 1
grein_de_genes <- all_edger_results[all_edger_results$FDR < 0.05 & 
                                    all_edger_results$logCPM > 1, ]

cat("DE genes with FDR < 0.05 AND logCPM > 1:", nrow(grein_de_genes), "\n")
cat("(GREIN reports 8,803 for this comparison)\n")

# This removes lowly expressed genes that are statistically significant
# but may be noisy or unreliable
```

::: {.callout-warning}
## Reproducibility Lesson #2

GREIN applies a **logCPM > 1 filter** to their DE gene counts that is **never mentioned** in their methods documentation. This removes ~1,700 lowly expressed genes from the reported DE list.

**Without this information, you cannot reproduce their exact results.**

This is why sharing actual code (not just prose descriptions) is critical!
:::

## Extract and Annotate Results
```{r edger-results}
# Use the GREIN-matched filtering for final results
edger_results <- grein_de_genes

# Add significance categories
edger_results$significant <- case_when(
  edger_results$logFC > 1 ~ "Up (>2-fold)",
  edger_results$logFC < -1 ~ "Down (>2-fold)",
  edger_results$logFC > 0 ~ "Up",
  edger_results$logFC < 0 ~ "Down"
)

# Summary statistics
cat("\n=== GREIN-Matched Results ===\n")
cat("Total DE genes (FDR < 0.05, logCPM > 1):", nrow(edger_results), "\n")
cat("Upregulated in KD:", sum(edger_results$logFC > 0), "\n")
cat("Downregulated in KD:", sum(edger_results$logFC < 0), "\n")
cat("Strong up (>2-fold):", sum(edger_results$logFC > 1), "\n")
cat("Strong down (>2-fold):", sum(edger_results$logFC < -1), "\n")

# Show top genes
head(edger_results[order(edger_results$FDR), ], 20)
```

## Visualize edgeR Results

### Volcano Plot
```{r edger-volcano, fig.width=8, fig.height=6}
ggplot(edger_results, aes(x = logFC, y = -log10(PValue), color = significant)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_color_manual(values = c("Up (>2-fold)" = "darkred",
                                "Down (>2-fold)" = "darkblue",
                                "Up" = "red", 
                                "Down" = "blue")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", alpha = 0.5) +
  labs(title = "Volcano Plot: KD vs WT (edgeR, GREIN-matched)",
       subtitle = paste0("FDR < 0.05 & logCPM > 1 (n = ", nrow(edger_results), " genes)"),
       x = "Log2 Fold Change",
       y = "-Log10 P-value",
       color = "Classification") +
  theme_minimal() +
  theme(legend.position = "right")

ggsave(here("demo-analysis", "output", "differential-expression", "figures", "volcano_plot_edger_grein.png"), width = 8, height = 6, dpi = 300)
```

### MA Plot
```{r edger-ma, fig.width=8, fig.height=6}
ggplot(edger_results, aes(x = logCPM, y = logFC, color = significant)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_color_manual(values = c("Up (>2-fold)" = "darkred",
                                "Down (>2-fold)" = "darkblue",
                                "Up" = "red", 
                                "Down" = "blue")) +
  geom_hline(yintercept = 0, linetype = "solid", alpha = 0.5) +
  geom_hline(yintercept = c(-1, 1), linetype = "dashed", alpha = 0.3) +
  labs(title = "MA Plot: KD vs WT (edgeR, GREIN-matched)",
       subtitle = paste0("FDR < 0.05 & logCPM > 1 (n = ", nrow(edger_results), " genes)"),
       x = "Average Expression (log2-CPM)",
       y = "Log2 Fold Change",
       color = "Classification") +
  theme_minimal()

ggsave(here("demo-analysis", "output", "differential-expression", "figures", "ma_plot_edger_grein.png"), width = 8, height = 6, dpi = 300)
```

## Compare edgeR vs limma-voom Results
```{r compare-methods}
# Get limma-voom results (from earlier analysis)
limma_results <- all_results  # From the limma-voom section

# For fair comparison, apply same filters to limma
limma_matched <- limma_results[limma_results$adj.P.Val < 0.05 & 
                               limma_results$AveExpr > 1, ]

# Create comparison data frame
comparison <- data.frame(
  gene = rownames(edger_results),
  edgeR_logFC = edger_results$logFC,
  edgeR_FDR = edger_results$FDR,
  limma_logFC = limma_results[rownames(edger_results), "logFC"],
  limma_FDR = limma_results[rownames(edger_results), "adj.P.Val"]
)

# Compare fold changes
cat("Correlation of logFC between methods:", 
    round(cor(comparison$edgeR_logFC, comparison$limma_logFC, use = "complete.obs"), 3), 
    "\n")

# Plot fold change comparison
ggplot(comparison, aes(x = limma_logFC, y = edgeR_logFC)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Fold Change Comparison: edgeR vs limma-voom",
       subtitle = "High correlation indicates robust findings",
       x = "limma-voom log2FC",
       y = "edgeR log2FC") +
  theme_minimal()

# Compare significance calls
cat("\n=== Method Comparison (with matched filters) ===\n")
cat("edgeR DE genes:", nrow(edger_results), "\n")
cat("limma-voom DE genes (FDR < 0.05, AveExpr > 1):", nrow(limma_matched), "\n")

# Gene overlap
edger_genes <- rownames(edger_results)
limma_genes <- rownames(limma_matched)

cat("\nOverlap:\n")
cat("  Both methods:", length(intersect(edger_genes, limma_genes)), "\n")
cat("  edgeR only:", length(setdiff(edger_genes, limma_genes)), "\n")
cat("  limma-voom only:", length(setdiff(limma_genes, edger_genes)), "\n")

# Calculate agreement percentage
overlap <- length(intersect(edger_genes, limma_genes))
total_unique <- length(union(edger_genes, limma_genes))
cat("  Agreement:", round(100 * overlap / total_unique, 1), "%\n")
```

::: {.callout-important}
## Which Method Should You Use?

**Use edgeR exact test when:**
- Simple two-group comparison
- Want to match published GREIN/GEO analyses
- Prefer count-based methods (no transformation)

**Use limma-voom when:**
- Complex designs (multiple factors, interactions, time series)
- Need flexibility with contrast matrices
- Sample quality varies (use voomWithQualityWeights)
- Want to leverage limma's mature ecosystem

**Both methods typically agree >95%** on which genes are DE for simple comparisons, which increases confidence in the findings!
:::

## Export edgeR Results
```{r export-edger}
# Prepare final results table with gene names
edger_export <- edger_results
edger_export$gene <- rownames(edger_export)

# Reorder columns for clarity
edger_export <- edger_export[, c("gene", "logFC", "logCPM", "PValue", "FDR", "significant")]

# Save results
write.csv(edger_export, 
          here("demo-analysis", "output", "differential-expression", "results", "edger_grein_matched_results.csv"),
          row.names = FALSE)

cat("edgeR results exported to:", here("demo-analysis", "output", "differential-expression", "results", "edger_grein_matched_results.csv"), "\n")
```

## Document Your Analysis (Reproducibility Best Practice!)

::: {.callout-tip}
## Make Your Analysis Reproducible

Always save your session information so others (including future you!) can reproduce your exact analysis:
```{r session-info}
# Save session info
writeLines(capture.output(sessionInfo()), 
           here("demo-analysis", "output", "differential-expression", "results", "sessionInfo_edger.txt"))

# Display key information
cat("R version:", R.version.string, "\n")
cat("edgeR version:", as.character(packageVersion("edgeR")), "\n")
cat("limma version:", as.character(packageVersion("limma")), "\n")
cat("Date:", Sys.Date(), "\n")
```

**What to document:**
- Exact software versions (R, edgeR, limma, etc.)
- All filtering thresholds (count >= 4, logCPM > 1, FDR < 0.05)
- Any post-hoc filters applied
- Random seeds if applicable
- Date of analysis
:::

---

## Key Takeaways

1. **edgeR and limma-voom give very similar results** for simple two-group comparisons (~95%+ agreement)

2. **Small differences exist** due to different statistical frameworks:
   - edgeR: Negative binomial model on counts
   - limma-voom: Linear model on log-transformed counts with precision weights

3. **Reproducibility requires complete documentation**:
   - GREIN's stated methods: "CPM > 0, FDR < 0.05"
   - GREIN's actual implementation: "count >= 4, FDR < 0.05, logCPM > 1"
   - Without testing, you wouldn't know this!

4. **Always share your actual code**, not just prose descriptions

5. **Use sessionInfo()** to document software versions


# Part 4: Complex Design - Two-Factor Analysis

::: {.callout-note}
## Switching Datasets to Teach Complex Designs

Our PRMT7 knockdown experiment has only **one factor** (genotype: WT vs KD). To demonstrate **two-factor designs** and **contrast matrices**, we'll switch to a published dataset that has a proper 2×2 factorial design.

**Dataset**: Drosophila melanogaster temperature adaptation study (Zhao et al. 2015, *PLoS Genetics*)

**Design:**
- **Population**: Maine (high latitude) vs Panama (low latitude)
- **Temperature**: Low vs High
- **Structure**: 2 populations × 2 temperatures × 3 replicates = 12 samples

This demonstrates how limma-voom handles complex experimental designs where you have multiple factors and want to test:
1. Main effects (population, temperature)
2. Interactions (does temperature response differ by population?)
3. Specific contrasts (temperature effect within each population)

These skills apply to ANY multi-factor experiment: genotype × treatment, tissue × disease, time × condition, etc.
:::

## Load Harvard Drosophila Data

```{r load-drosophila}
# Load the count matrix
drosophila_counts <- read.table(here("demo-analysis", "data", "two-factor-design", "salmon.merged.gene_counts.tsv"),
                                header = TRUE,
                                row.names = "gene_name")

# Remove the gene_id column (we're using gene names as row names)
drosophila_counts$gene_id <- NULL

# Load sample metadata
drosophila_info <- read.table(here("demo-analysis", "data", "two-factor-design", "dme_elev_samples.tsv"),
                              header = TRUE,
                              stringsAsFactors = FALSE)

# Create a combined factor for filtering
drosophila_info$pop_temp <- paste(drosophila_info$population, 
                                   drosophila_info$temp, 
                                   sep = "_")

# Display the experimental design
cat("Experimental Design:\n")
table(drosophila_info$population, drosophila_info$temp)

# Show sample info
drosophila_info
```

::: {.callout-note}
## About This Dataset

- **Species**: *Drosophila melanogaster* (fruit fly)
- **Question**: Do flies from different latitudes respond differently to temperature stress?
- **Samples**: Whole-body RNA-seq from adults
- **Design**: Perfectly balanced 2×2 factorial

This is a classic example of testing for **parallel vs. divergent** responses across populations.
:::

## Create DGE Object and Filter

```{r drosophila-dge}
# Create DGEList object
DGE_dros <- DGEList(counts = drosophila_counts,
                    samples = drosophila_info)

# Add group information
DGE_dros$samples$group <- factor(drosophila_info$pop_temp)

# Filter lowly expressed genes
keep_dros <- filterByExpr(DGE_dros, group = DGE_dros$samples$group)
DGE_dros <- DGE_dros[keep_dros, , keep.lib.sizes = FALSE]

cat("Genes after filtering:", nrow(DGE_dros), "\n")

# TMM normalization
DGE_dros <- calcNormFactors(DGE_dros, method = "TMM")
```

## Quick QC - MDS Plot

```{r drosophila-mds, fig.width=8, fig.height=6}
# MDS plot to check sample clustering
mds_dros <- plotMDS(DGE_dros, top = 500, plot = FALSE, gene.selection = "common")

mds_dros_df <- data.frame(
  sample = rownames(DGE_dros$samples),
  dim1 = mds_dros$x,
  dim2 = mds_dros$y,
  population = DGE_dros$samples$population,
  temperature = DGE_dros$samples$temp
)

ggplot(mds_dros_df, aes(x = dim1, y = dim2, 
                         color = population, 
                         shape = temperature)) +
  geom_point(size = 4) +
  geom_text_repel(aes(label = sample), size = 3) +
  labs(title = "MDS Plot: Drosophila Temperature Experiment",
       x = paste0("Leading logFC dim 1 (", round(mds_dros$var.explained[1]*100, 1), "%)"),
       y = paste0("Leading logFC dim 2 (", round(mds_dros$var.explained[2]*100, 1), "%)"),
       color = "Population",
       shape = "Temperature") +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")
```

::: {.callout-tip}
## What to Look For in MDS

- **Population separation**: Do Maine and Panama samples cluster apart?
- **Temperature effects**: Do low/high temperature samples separate?
- **Interaction**: Does the temperature separation differ between populations?

Good separation suggests strong biological effects that will be detected in DE analysis.
:::

## Create Two-Factor Design Matrix

```{r design-2factor-dros}
# Define factors with explicit levels
# First level becomes the reference (intercept)
DGE_dros$samples$population <- factor(DGE_dros$samples$population, 
                                      levels = c("maine", "panama"))
DGE_dros$samples$temp <- factor(DGE_dros$samples$temp, 
                                levels = c("high", "low"))

# Create additive two-factor design matrix
design_2factor_dros <- model.matrix(~population + temp, data = DGE_dros$samples)

# View the design
design_2factor_dros
colnames(design_2factor_dros)
```

::: {.callout-note}
## Understanding the Two-Factor Model Coefficients

This is an **additive model**: `~population + temp`

The coefficients represent:
- **(Intercept)**: Expression in Maine + High temperature (both reference levels)
- **populationpanama**: Effect of Panama vs. Maine (averaged across temperatures)
- **templow**: Effect of Low vs. High temperature (averaged across populations)

**Assumptions of the additive model:**
- Temperature effect is the same magnitude in both populations
- Population effect is the same magnitude in both temperatures
- No interaction between factors

If this assumption is violated (temperature effect differs by population), you'll need an interaction model (Part 5).
:::

## Fit Two-Factor Model

```{r voom-2factor-dros, fig.width=8, fig.height=5}
# Run voom with quality weights
vwts_2factor_dros <- voomWithQualityWeights(DGE_dros, 
                                            design = design_2factor_dros,
                                            normalize.method = "none",
                                            plot = TRUE)

# Fit linear model
fit_2factor_dros <- lmFit(vwts_2factor_dros, design_2factor_dros)
fit_2factor_dros <- eBayes(fit_2factor_dros, robust = TRUE)

# Get summary
summary(decideTests(fit_2factor_dros, adjust.method = "fdr", p.value = 0.05))
```

::: {.callout-tip}
## Interpreting Sample Quality Weights

The voomWithQualityWeights plot shows two panels:

**Top panel (voom)**: Mean-variance relationship (same as before)

**Bottom panel (sample weights)**: Weights assigned to each sample
- Weights near 1.0: Good quality samples
- Weights < 0.5: Potentially problematic samples
- Weights > 1.5: Exceptionally good samples

Outlier samples are automatically down-weighted in the analysis.
:::

## Extract Results for Each Coefficient

```{r results-2factor-dros}
# Temperature effect (main effect across both populations)
temp_results_dros <- topTable(fit_2factor_dros, 
                              coef = "templow",
                              adjust.method = "BH",
                              number = Inf)

cat("\n=== Temperature Main Effect ===\n")
cat("Genes DE for temperature (FDR < 0.05):", 
    sum(temp_results_dros$adj.P.Val < 0.05), "\n")
cat("Top 10 temperature-responsive genes:\n")
print(head(temp_results_dros, 10))

# Population effect (main effect across both temperatures)
pop_results_dros <- topTable(fit_2factor_dros, 
                             coef = "populationpanama",
                             adjust.method = "BH",
                             number = Inf)

cat("\n=== Population Main Effect ===\n")
cat("Genes DE for population (FDR < 0.05):", 
    sum(pop_results_dros$adj.P.Val < 0.05), "\n")
cat("Top 10 population-divergent genes:\n")
print(head(pop_results_dros, 10))
```

::: {.callout-note}
## Interpreting logFC in Two-Factor Models

**templow coefficient:**
- Positive logFC: Higher expression in low temperature (averaged across populations)
- Negative logFC: Higher expression in high temperature (averaged across populations)

**populationpanama coefficient:**
- Positive logFC: Higher expression in Panama (averaged across temperatures)
- Negative logFC: Higher expression in Maine (averaged across temperatures)

These are **main effects** - they average over the other factor.
:::

## Visualize Temperature Effect

```{r volcano-2factor-temp, fig.width=8, fig.height=6}
# Add significance categories
temp_results_dros$significant <- case_when(
  temp_results_dros$adj.P.Val < 0.05 & temp_results_dros$logFC > 1 ~ "Up in Low",
  temp_results_dros$adj.P.Val < 0.05 & temp_results_dros$logFC < -1 ~ "Up in High",
  temp_results_dros$adj.P.Val < 0.05 ~ "Significant",
  TRUE ~ "Not Significant"
)

# Volcano plot
ggplot(temp_results_dros, aes(x = logFC, y = -log10(P.Value), color = significant)) +
  geom_point(alpha = 0.5, size = 1) +
  scale_color_manual(values = c("Up in Low" = "blue", 
                                "Up in High" = "red", 
                                "Significant" = "orange",
                                "Not Significant" = "grey")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", alpha = 0.5) +
  labs(title = "Temperature Effect (Main Effect Across Populations)",
       subtitle = "Low vs High Temperature",
       x = "Log2 Fold Change (Low vs High)",
       y = "-Log10 P-value",
       color = "Category") +
  theme_minimal()

ggsave(here("demo-analysis", "output", "differential-expression", "figures", "volcano_drosophila_temperature.png"), 
       width = 8, height = 6, dpi = 300)
```

::: {.callout-warning}
## Limitation of Main Effects

The additive model assumes the temperature effect is **the same** in both Maine and Panama populations. But what if:

- Maine flies have a strong cold-adaptation response
- Panama flies (tropical) have a different response pattern
- The temperature effect differs by population?

This is an **interaction**, which requires a different model (see Part 5).
:::

---

# Part 5: Advanced - Interaction Model and Contrasts

The additive model in Part 4 assumes temperature effects are the same in both populations. But biological reality is often more nuanced - adaptive responses might differ between populations. 

**Interaction models** let us test whether the effect of one factor (temperature) depends on another factor (population).

## Create Interaction Design

Instead of separate coefficients for each factor, we'll create a **group-means parameterization** where each combination of factors gets its own coefficient.

```{r design-interaction-dros}
# Create a combined factor for all group combinations
DGE_dros$samples$group_combined <- factor(
  paste(DGE_dros$samples$population, DGE_dros$samples$temp, sep = "_"),
  levels = c("maine_high", "maine_low", "panama_high", "panama_low")
)

# Create design with no intercept (each group gets its own coefficient)
design_interaction_dros <- model.matrix(~0 + group_combined, 
                                        data = DGE_dros$samples)

# Clean up column names
colnames(design_interaction_dros) <- levels(DGE_dros$samples$group_combined)

design_interaction_dros
```

::: {.callout-note}
## Group-Means Parameterization

Using `~0 + group_combined` creates a coefficient for each group:

- **maine_high**: Mean expression in Maine at high temperature
- **maine_low**: Mean expression in Maine at low temperature  
- **panama_high**: Mean expression in Panama at high temperature
- **panama_low**: Mean expression in Panama at low temperature

**Advantage**: We can define any contrast we want between these groups!
:::

## Define Contrast Matrix

Contrasts let us test specific biological questions by comparing group means:

```{r contrasts-dros}
# Define contrasts of interest
cont_matrix_dros <- makeContrasts(
  # Temperature effect in Maine
  Maine_TempEffect = maine_low - maine_high,
  
  # Temperature effect in Panama
  Panama_TempEffect = panama_low - panama_high,
  
  # Population effect at high temperature
  Pop_at_High = panama_high - maine_high,
  
  # Population effect at low temperature
  Pop_at_Low = panama_low - maine_low,
  
  # INTERACTION: Does temperature effect differ by population?
  # This compares (maine_low - maine_high) vs (panama_low - panama_high)
  Interaction = (maine_low - maine_high) - (panama_low - panama_high),
  
  levels = design_interaction_dros
)

cont_matrix_dros
```

::: {.callout-important}
## Understanding Contrasts

Each contrast is a linear combination of group means:

**Maine_TempEffect**: `maine_low - maine_high`
- Tests temperature response specifically in Maine population
- Positive logFC = genes upregulated at low temperature in Maine

**Panama_TempEffect**: `panama_low - panama_high`
- Tests temperature response specifically in Panama population  
- Positive logFC = genes upregulated at low temperature in Panama

**Interaction**: `(maine_low - maine_high) - (panama_low - panama_high)`
- Tests whether temperature effect differs between populations
- Significant genes show **divergent** temperature responses
- Positive logFC = stronger low-temp response in Maine than Panama
- Negative logFC = stronger low-temp response in Panama than Maine

**Non-significant interaction** = parallel responses (same temperature effect in both populations)
:::

## Fit Model with Contrasts

```{r fit-contrasts-dros, fig.width=8, fig.height=5}
# Run voom with interaction design
vwts_interaction_dros <- voomWithQualityWeights(DGE_dros, 
                                                design = design_interaction_dros,
                                                normalize.method = "none",
                                                plot = TRUE)

# Fit linear model
fit_interaction_dros <- lmFit(vwts_interaction_dros, design_interaction_dros)

# Apply contrasts
fit_contrasts_dros <- contrasts.fit(fit_interaction_dros, cont_matrix_dros)

# Empirical Bayes
fit_contrasts_dros <- eBayes(fit_contrasts_dros, robust = TRUE)

# Summary of all contrasts
contrast_summary <- decideTests(fit_contrasts_dros, 
                                adjust.method = "fdr", 
                                p.value = 0.05)

summary(contrast_summary)
```

## Visualize Contrast Overlap with Venn Diagram

```{r venn-contrasts, fig.width=8, fig.height=6}
# Venn diagram showing overlap of significant genes
vennDiagram(contrast_summary[, c("Maine_TempEffect", 
                                  "Panama_TempEffect", 
                                  "Interaction")],
            circle.col = c("blue", "red", "green"),
            main = "Overlap of DE Genes Across Contrasts")
```

::: {.callout-tip}
## Interpreting the Venn Diagram

- **Overlap between Maine_TempEffect & Panama_TempEffect**: Genes showing parallel temperature response
- **Interaction only**: Genes with divergent responses
- **Maine_TempEffect only**: Genes responding to temperature in Maine but not Panama
- **Panama_TempEffect only**: Genes responding to temperature in Panama but not Maine

This visualization helps identify convergent vs. divergent adaptation patterns!
:::

## Extract Results for Each Contrast

```{r extract-contrasts}
# Temperature effect in Maine
maine_temp <- topTable(fit_contrasts_dros, 
                       coef = "Maine_TempEffect",
                       adjust.method = "BH",
                       number = Inf,
                       sort.by = "P")

# Temperature effect in Panama  
panama_temp <- topTable(fit_contrasts_dros, 
                        coef = "Panama_TempEffect",
                        adjust.method = "BH",
                        number = Inf,
                        sort.by = "P")

# Interaction
interaction <- topTable(fit_contrasts_dros, 
                        coef = "Interaction",
                        adjust.method = "BH",
                        number = Inf,
                        sort.by = "P")

# Summary
cat("\n=== Contrast Results (FDR < 0.05) ===\n")
cat("Maine temperature effect:", sum(maine_temp$adj.P.Val < 0.05), "genes\n")
cat("Panama temperature effect:", sum(panama_temp$adj.P.Val < 0.05), "genes\n")
cat("Interaction (divergent responses):", sum(interaction$adj.P.Val < 0.05), "genes\n")

# Show top genes for each
cat("\n=== Top 10 Maine Temperature-Responsive Genes ===\n")
print(head(maine_temp, 10))

cat("\n=== Top 10 Panama Temperature-Responsive Genes ===\n")
print(head(panama_temp, 10))

cat("\n=== Top 10 Genes with Divergent Responses (Interaction) ===\n")
print(head(interaction, 10))
```

## Create Merged Results Table

For complex questions, merge results from multiple contrasts:

```{r merge-contrasts}
# Create a unified results table with all contrasts
# Start with gene names
all_genes <- rownames(maine_temp)

# Initialize results data frame
contrast_results <- data.frame(
  gene = all_genes,
  
  # Maine temperature effect
  Maine_logFC = maine_temp[all_genes, "logFC"],
  Maine_FDR = maine_temp[all_genes, "adj.P.Val"],
  
  # Panama temperature effect
  Panama_logFC = panama_temp[all_genes, "logFC"],
  Panama_FDR = panama_temp[all_genes, "adj.P.Val"],
  
  # Interaction
  Interaction_logFC = interaction[all_genes, "logFC"],
  Interaction_FDR = interaction[all_genes, "adj.P.Val"]
)

# Calculate agreement
contrast_results <- contrast_results %>%
  mutate(
    Maine_sig = Maine_FDR < 0.05,
    Panama_sig = Panama_FDR < 0.05,
    Interaction_sig = Interaction_FDR < 0.05,
    
    # Classify response patterns
    pattern = case_when(
      # Parallel responses (both sig, no interaction)
      Maine_sig & Panama_sig & !Interaction_sig & 
        sign(Maine_logFC) == sign(Panama_logFC) ~ "Parallel",
      
      # Divergent responses (significant interaction)
      Interaction_sig ~ "Divergent",
      
      # Maine-specific
      Maine_sig & !Panama_sig ~ "Maine-specific",
      
      # Panama-specific
      Panama_sig & !Maine_sig ~ "Panama-specific",
      
      # Not significant
      TRUE ~ "Not significant"
    )
  )

# Summary of patterns
cat("\n=== Response Pattern Classification ===\n")
table(contrast_results$pattern)

# Show examples of each pattern
cat("\n=== Example Parallel Response Genes ===\n")
parallel_genes <- contrast_results %>% 
  filter(pattern == "Parallel") %>%
  arrange(Maine_FDR) %>%
  head(5)
print(parallel_genes)

cat("\n=== Example Divergent Response Genes ===\n")
divergent_genes <- contrast_results %>% 
  filter(pattern == "Divergent") %>%
  arrange(Interaction_FDR) %>%
  head(5)
print(divergent_genes)
```

::: {.callout-note}
## Biological Interpretation

**Parallel responses**: Genes showing conserved temperature adaptation
- Same direction and magnitude in both populations
- Likely core thermal stress responses

**Divergent responses**: Genes showing population-specific adaptation
- Different response patterns between Maine and Panama
- Evidence of evolutionary divergence in temperature response

**Population-specific**: Genes responding in one population only
- May reflect local adaptation
- Could be due to power (smaller effect in other population)

This framework applies to any 2-factor experiment asking: "Do these groups respond the same way to treatment?"
:::

## Visualize Divergent Genes

```{r heatmap-divergent, fig.width=10, fig.height=8}
# Get the top 50 most divergent genes
top_divergent <- divergent_genes %>% 
  head(50) %>% 
  pull(gene)

# Extract normalized expression for these genes
divergent_expr <- vwts_interaction_dros$E[top_divergent, ]

# Create annotation for columns
col_annotation <- data.frame(
  Population = DGE_dros$samples$population,
  Temperature = DGE_dros$samples$temp,
  row.names = colnames(divergent_expr)
)

# Create heatmap
pheatmap(divergent_expr,
         scale = "row",
         annotation_col = col_annotation,
         cluster_cols = FALSE,  # Keep samples in order
         show_rownames = TRUE,
         show_colnames = TRUE,
         main = "Top 50 Genes with Divergent Temperature Responses",
         color = colorRampPalette(c("blue", "white", "red"))(100),
         fontsize_row = 6)
```


---

## Key Takeaways from Complex Designs

::: {.callout-important}
## When to Use Each Approach

**Additive Model** (`~factor1 + factor2`):
- ✓ Tests main effects
- ✓ Simpler, more power if no interaction
- ✗ Assumes effects are independent
- **Use when**: You expect similar responses across groups

**Interaction Model** (`~0 + group_combined` + contrasts):
- ✓ Tests specific hypotheses with contrasts
- ✓ Detects divergent responses
- ✓ Maximum flexibility
- ✗ More complex to interpret
- **Use when**: You expect group-specific responses

**General Principle**: Start with additive model, check for interaction, then use contrasts if needed.
:::

::: {.callout-tip}
## Applying to Your Own Data

The contrast approach works for ANY factorial design:

- **Genotype × Treatment**: Does treatment work differently in mutant vs. wildtype?
- **Tissue × Disease**: Are disease genes different between tissues?
- **Time × Condition**: Do treated and control diverge over time?
- **Cell Type × Stimulus**: Cell-type-specific responses?

The workflow is always:
1. Create group-means design (`~0 + group`)
2. Define contrasts for your questions
3. Test and visualize

This is one of limma-voom's greatest strengths!
:::

---

## Return to Your PRMT7 Data

Now that you've seen complex designs in action, return to your PRMT7 analysis for final visualization and export (Part 6). The skills you learned here - thinking about contrasts, interactions, and response patterns - will help you design better experiments in the future!



# Part 6: Visualization and Export

## Heatmap of Top DE Genes

```{r heatmap-top-genes, fig.width=8, fig.height=10}
# Get top 50 genes from simple treatment comparison
top50_genes <- rownames(topTable(fit_simple, coef = 2, number = 50))

# Extract log-CPM values for these genes
top50_logcpm <- cpm(DGE[top50_genes, ], log = TRUE)

# Z-score normalize for better visualization
top50_scaled <- t(scale(t(top50_logcpm)))

# Create annotation
annotation_col <- data.frame(
  Treatment = DGE$samples$treatment,
  row.names = colnames(top50_logcpm)
)

# Create heatmap
pheatmap(top50_scaled,
         annotation_col = annotation_col,
         scale = "none",
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         show_rownames = TRUE,
         show_colnames = TRUE,
         main = "Top 50 DE Genes (Treatment Effect)",
         fontsize_row = 6,
         fontsize_col = 8,
         color = colorRampPalette(c("blue", "white", "red"))(100))

# Save
pheatmap(top50_scaled,
         annotation_col = annotation_col,
         scale = "none",
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         show_rownames = TRUE,
         show_colnames = TRUE,
         main = "Top 50 DE Genes (Treatment Effect)",
         fontsize_row = 6,
         fontsize_col = 8,
         color = colorRampPalette(c("blue", "white", "red"))(100),
         filename = here("demo-analysis", "output", "differential-expression", "figures","heatmap_top50.png"),
         width = 8,
         height = 10)
```

## Export Results Tables

```{r export-contrasts}
# Save the merged contrast results
write.csv(contrast_results, 
          here("demo-analysis", "output", "differential-expression", "results", "drosophila_contrast_results.csv"),
          row.names = FALSE)

# Save individual contrast tables
write.csv(maine_temp, 
          here("demo-analysis", "output", "differential-expression", "results", "drosophila_maine_temperature.csv"),
          row.names = TRUE)

write.csv(panama_temp, 
          here("demo-analysis", "output", "differential-expression", "results", "drosophila_panama_temperature.csv"),
          row.names = TRUE)

write.csv(interaction, 
          here("demo-analysis", "output", "differential-expression", "results", "drosophila_interaction.csv"),
          row.names = TRUE)

# Optional: Export just the parallel response genes
parallel_response <- contrast_results %>% 
  filter(pattern == "Parallel")

write.csv(parallel_response, 
          here("demo-analysis", "output", "differential-expression", "results", "drosophila_parallel_response_genes.csv"),
          row.names = FALSE)

# Optional: Export just the divergent response genes
divergent_response <- contrast_results %>% 
  filter(pattern == "Divergent")

write.csv(divergent_response, 
          here("demo-analysis", "output", "differential-expression", "results", "drosophila_divergent_response_genes.csv"),
          row.names = FALSE)

cat("All contrast results exported!\n")
cat("Files created:\n")
cat("  - drosophila_contrast_results.csv (all genes with pattern classification)\n")
cat("  - drosophila_maine_temperature.csv (Maine temperature effect)\n")
cat("  - drosophila_panama_temperature.csv (Panama temperature effect)\n")
cat("  - drosophila_interaction.csv (interaction term)\n")
cat("  - drosophila_parallel_response_genes.csv (parallel patterns only)\n")
cat("  - drosophila_divergent_response_genes.csv (divergent patterns only)\n")
```

# Summary and Key Takeaways

## What We Covered

1. **Data Loading**: RSEM counts from nf-core pipeline
2. **Quality Control**: Library sizes, MDS plots, sample correlations
3. **Filtering & Normalization**: Removing low counts, TMM normalization
4. **Simple Design**: Two-group comparison with limma-voom
5. **Complex Design**: Two-factor model with main effects
6. **Advanced Contrasts**: Interaction model for specific biological questions
7. **Visualization**: Volcano plots, MA plots, heatmaps

## Key Concepts to Remember

::: {.callout-important}
## Critical Points

1. **Filter before normalizing** - remove genes with insufficient counts
2. **TMM normalization** handles library size and composition bias
3. **voom transformation** converts counts to log-scale with precision weights
4. **Quality weights** handle outlier samples without removing them
5. **Design matrix** defines what you're testing
6. **Contrasts** let you ask specific biological questions
7. **Multiple testing correction** (FDR/BH) is essential
8. **LogFC and p-value together** tell the full story (effect size + significance)
:::

## Next Steps

- **Gene Set Enrichment Analysis**: Are DE genes enriched for specific pathways/functions?
- **Validation**: qRT-PCR or additional experiments to confirm key findings
- **Biological Interpretation**: Connect results back to your biological question
- **Reproducibility**: Document your analysis, save your code, version control!

## Resources

- [limma User's Guide](https://bioconductor.org/packages/release/bioc/vignettes/limma/inst/doc/usersguide.pdf)
- [edgeR User's Guide](https://bioconductor.org/packages/release/bioc/vignettes/edgeR/inst/doc/edgeRUsersGuide.pdf)
- [RNA-seq analysis in R (Harvard Chan Bioinformatics Core)](https://hbctraining.github.io/DGE_workshop/)
- UF Research Computing: [https://www.rc.ufl.edu/](https://www.rc.ufl.edu/)

---

# Part 7: Pathway Enrichment Analysis

::: {.callout-note}
## From Genes to Biology

We've identified thousands of differentially expressed genes, but what does this mean biologically? 

**Pathway enrichment analysis** helps us answer:
- What biological processes are affected?
- What pathways are dysregulated?
- Are there common themes among DE genes?

This transforms gene lists into biological insights!
:::

## Conceptual Overview: Over-Representation Analysis

**The Basic Question**: Are my DE genes **enriched** in certain pathways/processes compared to what we'd expect by chance?

::: {.columns}
::: {.column width="50%"}
**Example:**

- Total genes in genome: 20,000
- Genes in "DNA Repair" pathway: 200 (1%)
- Your DE genes: 1,000
- Expected in DNA Repair: 10 genes (1% of 1,000)
- **Actual in DNA Repair: 50 genes**
:::

::: {.column width="50%"}
**Statistical Test:**

Is 50 significantly different from 10?

→ Use **hypergeometric test** (Fisher's exact test)

If p < 0.05: "DNA Repair is enriched!"

**Biological interpretation**: Your treatment affects DNA repair processes
:::
:::

::: {.callout-warning}
## Important Assumption

ORA assumes all DE genes are equally important. It treats a gene with logFC = 5 (32-fold change) the same as logFC = 0.6 (1.5-fold change).

**GSEA** (Gene Set Enrichment Analysis) addresses this by using ranked gene lists, but it's more complex. We'll mention it at the end.
:::

## Setup: Load Required Packages

```{r load-pathway-packages}
# Install if needed (uncomment):
# BiocManager::install(c("clusterProfiler", "org.Mm.eg.db", "enrichplot", "pathview"))

library(clusterProfiler)
library(org.Mm.eg.db)  # Mouse annotation (use org.Hs.eg.db for human)
library(enrichplot)
library(ggplot2)
library(here)

# Check organism database
org.Mm.eg.db
```

::: {.callout-tip}
## Species-Specific Annotation Packages

- **Human**: `org.Hs.eg.db`
- **Mouse**: `org.Mm.eg.db`
- **Rat**: `org.Rn.eg.db`
- **Fly**: `org.Dm.eg.db`
- **Zebrafish**: `org.Dr.eg.db`
- **C. elegans**: `org.Ce.eg.db`

Install the appropriate one for your species!
:::

## Prepare Gene Lists

We'll use the limma-voom results from Part 3 (the simple PRMT7_KD vs WT comparison).

```{r prepare-gene-lists}
# From Part 3, we have: all_results (limma-voom output)
# Let's refresh it if needed
if (!exists("all_results")) {
  all_results <- topTable(fit_simple, 
                          coef = 2,
                          adjust.method = "BH",
                          number = Inf,
                          sort.by = "none")
}

# Extract significant genes (FDR < 0.05)
sig_genes <- all_results %>%
  filter(adj.P.Val < 0.05)

cat("Total DE genes:", nrow(sig_genes), "\n")

# Separate up and down-regulated genes
up_genes <- sig_genes %>% filter(logFC > 0)
down_genes <- sig_genes %>% filter(logFC < 0)

cat("Upregulated in KD:", nrow(up_genes), "\n")
cat("Downregulated in KD:", nrow(down_genes), "\n")
```

## Convert Gene Symbols to Entrez IDs

Most pathway databases use **Entrez Gene IDs**, not gene symbols. We need to convert:

```{r convert-to-entrez}
# Get all gene symbols from our results
gene_symbols <- rownames(sig_genes)

# Convert symbols to Entrez IDs
gene_entrez <- bitr(gene_symbols, 
                    fromType = "SYMBOL",
                    toType = "ENTREZID", 
                    OrgDb = org.Mm.eg.db)

# Check conversion
cat("Original genes:", length(gene_symbols), "\n")
cat("Successfully converted:", nrow(gene_entrez), "\n")
cat("Failed to convert:", length(gene_symbols) - nrow(gene_entrez), "\n")

# Show examples
head(gene_entrez)
```

::: {.callout-warning}
## Gene ID Conversion Issues

Not all genes convert successfully:

- **Novel genes**: Not yet in annotation databases
- **Non-coding RNAs**: May lack Entrez IDs
- **Pseudogenes**: Often not annotated
- **Outdated symbols**: Gene nomenclature changes

**Typical conversion rate**: 85-95%

This is normal and expected!
:::

## Background Gene Set

For accurate enrichment testing, we need to define the **background** (universe) - all genes that COULD have been detected:

```{r prepare-background}
# All genes that passed filtering and were tested
background_symbols <- rownames(all_results)

# Convert background to Entrez IDs
background_entrez <- bitr(background_symbols,
                          fromType = "SYMBOL",
                          toType = "ENTREZID",
                          OrgDb = org.Mm.eg.db)

cat("Background genes:", nrow(background_entrez), "\n")

# Extract just the Entrez IDs
universe <- background_entrez$ENTREZID
```

::: {.callout-important}
## Why Background Matters

**Wrong**: Compare to all genes in genome (20,000+)

**Right**: Compare to genes expressed in your experiment (13,000)

Using the correct background prevents:
- False positives (from untested genes)
- Biased enrichment (from tissue-specific expression)

Always use your filtered gene set as background!
:::

## Gene Ontology (GO) Enrichment

GO provides a standardized vocabulary for gene function across three domains:

- **Biological Process (BP)**: What the gene does (e.g., "DNA repair", "cell cycle")
- **Molecular Function (MF)**: Biochemical activity (e.g., "kinase activity", "DNA binding")
- **Cellular Component (CC)**: Where it acts (e.g., "nucleus", "mitochondrion")

### Biological Process Enrichment

```{r enrichGO-BP}
# Run GO enrichment for Biological Process
ego_bp <- enrichGO(gene = gene_entrez$ENTREZID,
                   universe = universe,
                   OrgDb = org.Mm.eg.db,
                   ont = "BP",           # Biological Process
                   pAdjustMethod = "BH",
                   pvalueCutoff = 0.05,
                   qvalueCutoff = 0.05,
                   readable = TRUE)      # Convert back to gene symbols

# Summary
cat("Enriched GO-BP terms:", nrow(ego_bp), "\n\n")

# Show top 10 terms
head(ego_bp, 10)
```

::: {.callout-tip}
## Understanding the Output

Key columns:

- **Description**: GO term name
- **GeneRatio**: X/Y = X genes in your list / Y genes in term
- **BgRatio**: A/B = A genes in term / B genes in background
- **pvalue**: Hypergeometric test p-value
- **p.adjust**: FDR-corrected p-value
- **Count**: Number of your genes in this term
- **geneID**: Which genes (gene symbols if readable=TRUE)

**Rich Factor** = GeneRatio / BgRatio  
→ Higher = more enriched
:::

### Visualize GO Results: Dot Plot

```{r dotplot-GO, fig.width=10, fig.height=8}
# Dot plot of top 20 terms
dotplot(ego_bp, 
        showCategory = 20,
        title = "GO Biological Process Enrichment",
        font.size = 10)

ggsave(here("demo-analysis", "output", "differential-expression", "figures", "GO_BP_dotplot.png"), width = 10, height = 8, dpi = 300)
```

::: {.callout-note}
## Reading the Dot Plot

- **X-axis**: Gene Ratio (proportion of genes in term)
- **Y-axis**: GO terms (sorted by p-value)
- **Dot size**: Number of genes (Count)
- **Dot color**: Adjusted p-value (darker = more significant)

**Look for**:
- Large dots = many genes
- Dark colors = highly significant
- High GeneRatio = large proportion of term covered
:::

### Visualize GO Results: Enrichment Map

Shows relationships between enriched terms (terms sharing genes are connected):

```{r enrichment-map, fig.width=12, fig.height=10}
# Convert to enrichment map format
ego_bp_simplified <- pairwise_termsim(ego_bp)

# Create enrichment map
emapplot(ego_bp_simplified, 
         showCategory = 30,
         #cex_label_category = 0.6,
         layout = "kk") +
  labs(title = "GO Enrichment Map: Related Terms Share Genes")

ggsave(here("demo-analysis", "output", "differential-expression", "figures", "GO_BP_emap.png"), width = 12, height = 10, dpi = 300)
```

::: {.callout-tip}
## Interpreting Enrichment Maps

- **Nodes**: GO terms
- **Node size**: Number of genes
- **Node color**: Significance (p-value)
- **Edges**: Shared genes between terms
- **Clusters**: Related biological processes

**Use this to**:
- Identify major themes
- Reduce redundancy (many similar terms)
- Find unexpected connections
:::

### Visualize GO Results: Gene-Concept Network

Shows which genes belong to which terms:

```{r cnetplot, fig.width=12, fig.height=10}
# Show top 5 terms and their genes
cnetplot(ego_bp, 
         showCategory = 5,
        # circular = FALSE,
         # colorEdge = TRUE,
         node_label = "all") +
         # cex_label_gene = 0.7,
         #cex_label_category = 1.2) 
  labs(title = "Gene-Concept Network: Top 5 GO Terms")

ggsave(here("demo-analysis", "output", "differential-expression", "figures", "GO_BP_cnetplot.png"), width = 12, height = 10, dpi = 300)
```

::: {.callout-note}
## Gene-Concept Networks

- **Brown nodes**: GO terms
- **Colored nodes**: Genes (color = fold change if provided)
- **Edges**: Gene belongs to term

**Useful for**:
- Seeing which genes drive enrichment
- Identifying "hub genes" (in multiple terms)
- Understanding term overlap
:::

### Other GO Ontologies (Optional)

```{r enrichGO-other, eval=FALSE}
# Molecular Function
ego_mf <- enrichGO(gene = gene_entrez$ENTREZID,
                   universe = universe,
                   OrgDb = org.Mm.eg.db,
                   ont = "MF",
                   pAdjustMethod = "BH",
                   pvalueCutoff = 0.05,
                   readable = TRUE)

# Cellular Component
ego_cc <- enrichGO(gene = gene_entrez$ENTREZID,
                   universe = universe,
                   OrgDb = org.Mm.eg.db,
                   ont = "CC",
                   pAdjustMethod = "BH",
                   pvalueCutoff = 0.05,
                   readable = TRUE)

# View results
head(ego_mf, 10)
head(ego_cc, 10)
```

## KEGG Pathway Enrichment

KEGG provides curated pathway maps for metabolism, signaling, disease, etc.

```{r enrichKEGG}
# KEGG uses organism codes: "mmu" = mouse, "hsa" = human
kegg_result <- enrichKEGG(gene = gene_entrez$ENTREZID,
                          universe = universe,
                          organism = "mmu",      # Mouse
                          pAdjustMethod = "BH",
                          pvalueCutoff = 0.05,
                          qvalueCutoff = 0.05)

# Add gene symbols (KEGG uses Entrez IDs)
kegg_result <- setReadable(kegg_result, 
                           OrgDb = org.Mm.eg.db, 
                           keyType = "ENTREZID")

# Summary
cat("Enriched KEGG pathways:", nrow(kegg_result), "\n\n")

# Show results
head(kegg_result, 15)
```

::: {.callout-tip}
## Common KEGG Organism Codes

- **mmu**: Mouse (*Mus musculus*)
- **hsa**: Human (*Homo sapiens*)
- **rno**: Rat (*Rattus norvegicus*)
- **dme**: Fly (*Drosophila melanogaster*)
- **dre**: Zebrafish (*Danio rerio*)
- **cel**: C. elegans

Find more: https://www.genome.jp/kegg/catalog/org_list.html
:::

### Visualize KEGG Results

```{r kegg-dotplot, fig.width=10, fig.height=6}
# Dot plot
dotplot(kegg_result, 
        showCategory = 20,
        title = "KEGG Pathway Enrichment")

ggsave(here("demo-analysis", "output", "differential-expression", "figures", "KEGG_dotplot.png"), width = 10, height = 6, dpi = 300)
```

### Compare GO vs KEGG

```{r compare-databases, fig.width=10, fig.height=8}
# Convert to comparable format
ego_df <- as.data.frame(ego_bp) %>%
  mutate(Database = "GO-BP",
         Term = Description,
         Count = Count,
         FDR = p.adjust) %>%
  dplyr::select(Database, Term, Count, FDR) %>%
  head(10)

kegg_df <- as.data.frame(kegg_result) %>%
  mutate(Database = "KEGG",
         Term = Description,
         Count = Count,
         FDR = p.adjust) %>%
  dplyr::select(Database, Term, Count, FDR) %>%
  head(10)

# Combine
comparison_df <- bind_rows(ego_df, kegg_df)

# Plot
ggplot(comparison_df, aes(x = Count, y = reorder(Term, Count), fill = Database)) +
  geom_col() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Top Enriched Terms: GO vs KEGG",
       x = "Number of Genes",
       y = "Term") +
  theme_minimal() +
  facet_wrap(~Database, scales = "free_y", ncol = 1)

ggsave(here("demo-analysis", "output", "differential-expression", "figures", "GO_vs_KEGG_comparison.png"), width = 10, height = 8, dpi = 300)
```

::: {.callout-note}
## GO vs KEGG: When to Use Each?

**Gene Ontology (GO):**
- ✓ Comprehensive (more terms)
- ✓ Broad biological processes
- ✓ Good for exploratory analysis
- ✗ Can be redundant (many overlapping terms)

**KEGG Pathways:**
- ✓ Curated signaling/metabolic pathways
- ✓ Pathway diagrams available
- ✓ Less redundancy
- ✗ Fewer pathways (more focused)

**Recommendation**: Run both! They complement each other.
:::

## Simplify Redundant GO Terms

GO terms are hierarchical and often redundant. Simplify using semantic similarity:

```{r simplify-GO}
# Simplify GO results by removing redundant terms
ego_bp_simple <- simplify(ego_bp,
                          cutoff = 0.7,      # Similarity threshold
                          by = "p.adjust",   # Keep most significant
                          select_fun = min)

cat("Original GO terms:", nrow(ego_bp), "\n")
cat("After simplification:", nrow(ego_bp_simple), "\n\n")

# Compare
cat("Top 10 simplified terms:\n")
head(ego_bp_simple, 10)
```

```{r simplified-dotplot, fig.width=10, fig.height=6}
# Plot simplified results
dotplot(ego_bp_simple, 
        showCategory = 20,
        title = "Simplified GO-BP Enrichment (Redundancy Removed)")

ggsave(here("demo-analysis", "output", "differential-expression", "figures", "GO_BP_simplified_dotplot.png"), width = 10, height = 6, dpi = 300)
```

## Up vs Down-Regulated Genes Separately

Sometimes different processes are enriched in up vs down-regulated genes:

```{r separate-directions}
# Upregulated genes
up_entrez <- gene_entrez %>%
  filter(SYMBOL %in% rownames(up_genes)) %>%
  pull(ENTREZID)

ego_up <- enrichGO(gene = up_entrez,
                   universe = universe,
                   OrgDb = org.Mm.eg.db,
                   ont = "BP",
                   pvalueCutoff = 0.05,
                   readable = TRUE)

# Downregulated genes
down_entrez <- gene_entrez %>%
  filter(SYMBOL %in% rownames(down_genes)) %>%
  pull(ENTREZID)

ego_down <- enrichGO(gene = down_entrez,
                     universe = universe,
                     OrgDb = org.Mm.eg.db,
                     ont = "BP",
                     pvalueCutoff = 0.05,
                     readable = TRUE)

cat("GO terms enriched in upregulated genes:", nrow(ego_up), "\n")
cat("GO terms enriched in downregulated genes:", nrow(ego_down), "\n")
```

```{r compare-up-down, fig.width=12, fig.height=8}
# Compare up vs down
p1 <- dotplot(ego_up, showCategory = 10, title = "Upregulated in KD") + 
  theme(axis.text.y = element_text(size = 8))

p2 <- dotplot(ego_down, showCategory = 10, title = "Downregulated in KD") + 
  theme(axis.text.y = element_text(size = 8))

# Combine plots
library(patchwork)
p1 | p2

ggsave(here("demo-analysis", "output", "differential-expression", "figures", "GO_up_vs_down.png"), width = 12, height = 8, dpi = 300)
```

::: {.callout-tip}
## Biological Interpretation

Analyzing up/down separately can reveal:

- **Upregulated**: Stress responses, inflammation, defense
- **Downregulated**: Normal functions disrupted, metabolic processes

This provides more nuanced biological insight than treating all DE genes the same!
:::

## Export Results

```{r export-pathway-results}
# Export GO results as tables
write.csv(as.data.frame(ego_bp), 
          here("demo-analysis", "output", "differential-expression", "results", "GO_BP_enrichment.csv"),
          row.names = FALSE)

write.csv(as.data.frame(ego_bp_simple), 
          here("demo-analysis", "output", "differential-expression", "results", "GO_BP_enrichment_simplified.csv"),
          row.names = FALSE)

# Export KEGG results
write.csv(as.data.frame(kegg_result), 
          here("demo-analysis", "output", "differential-expression", "results", "KEGG_enrichment.csv"),
          row.names = FALSE)

# Export separate up/down GO results
write.csv(as.data.frame(ego_up), 
          here("demo-analysis", "output", "differential-expression", "results", "GO_BP_upregulated.csv"),
          row.names = FALSE)

write.csv(as.data.frame(ego_down), 
          here("demo-analysis", "output", "differential-expression", "results", "GO_BP_downregulated.csv"),
          row.names = FALSE)

cat("Pathway enrichment results exported!\n")
```

---

## Advanced Topic: Gene Set Enrichment Analysis (GSEA)

::: {.callout-note}
## What is GSEA?

**Over-Representation Analysis (ORA)** uses a binary cutoff:
- Gene is DE (FDR < 0.05) → Include
- Gene is not DE → Exclude

**Gene Set Enrichment Analysis (GSEA)** uses ALL genes:
- Rank all genes by statistic (e.g., t-statistic, logFC)
- Test if a gene set is enriched at top or bottom of ranked list
- No arbitrary cutoff needed!

**Advantages:**
- Uses full dataset (more information)
- Detects coordinated changes (many small effects)
- No cutoff bias

**Disadvantages:**
- More complex to interpret
- Requires proper ranking metric
- Takes longer to run
:::

### When to Use GSEA vs ORA

::: {.columns}
::: {.column width="50%"}
**Use ORA when:**
- Clear DE cutoff makes sense (FDR < 0.05)
- Strong effect sizes
- Want simple interpretation
- Quick exploratory analysis
- Teaching beginners
:::

::: {.column width="50%"}
**Use GSEA when:**
- Subtle, coordinated changes
- No clear DE cutoff
- Small effect sizes across pathway
- More rigorous analysis needed
- Publishing results
:::
:::

### GSEA with clusterProfiler (Quick Example)

```{r gsea-example, eval=FALSE}
# Prepare ranked gene list (all genes, ranked by statistic)
# Use signed -log10(p-value) * sign(logFC) as ranking metric

gene_list <- all_results %>%
  mutate(rank_stat = -log10(P.Value) * sign(logFC)) %>%
  arrange(desc(rank_stat)) %>%
  pull(rank_stat, name = "SYMBOL")

# Convert symbols to Entrez IDs
gene_list_entrez <- bitr(names(gene_list),
                         fromType = "SYMBOL",
                         toType = "ENTREZID",
                         OrgDb = org.Mm.eg.db)

# Create named vector with Entrez IDs
gene_list_final <- gene_list[gene_list_entrez$SYMBOL]
names(gene_list_final) <- gene_list_entrez$ENTREZID

# Run GSEA for GO
gsea_go <- gseGO(geneList = gene_list_final,
                 OrgDb = org.Mm.eg.db,
                 ont = "BP",
                 pvalueCutoff = 0.05,
                 pAdjustMethod = "BH")

# Run GSEA for KEGG
gsea_kegg <- gseKEGG(geneList = gene_list_final,
                     organism = "mmu",
                     pvalueCutoff = 0.05)

# Visualize
gseaplot2(gsea_go, geneSetID = 1:3, title = "GSEA: Top 3 GO Terms")
```

::: {.callout-important}
## GSEA is Beyond This Workshop

GSEA requires understanding:
- Ranking metrics (which to use?)
- Enrichment scores (what do they mean?)
- Leading edge genes (core enrichment)
- Positive vs negative enrichment

**For today**: ORA is sufficient and easier to interpret!

**For future**: Consider GSEA for publication-quality analysis.
:::

---

## Common Pitfalls in Pathway Analysis

::: {.callout-warning}
## Mistakes to Avoid

**1. Wrong background**
❌ Using all genome genes  
✅ Use genes tested in your experiment

**2. Ignoring multiple testing**
❌ p-value < 0.05  
✅ FDR (q-value) < 0.05

**3. Over-interpreting single genes**
- One gene can appear in many terms
- Focus on biological themes, not single genes

**4. Cherry-picking results**
- Report all enriched terms, not just favorites
- Include "uninteresting" results too

**5. Ignoring effect size**
- Term with 3 genes vs 50 genes
- Both significant, but 50 is stronger evidence

**6. Not simplifying GO**
- Many redundant GO terms
- Use `simplify()` to reduce redundancy

**7. Database version matters**
- GO updates quarterly
- KEGG updates irregularly
- Document database versions!
:::

## Summary and Next Steps

::: {.callout-important}
## What We Learned

**Pathway Enrichment Analysis:**

1. ✓ Over-representation testing (hypergeometric test)
2. ✓ Gene ID conversion (Symbol → Entrez)
3. ✓ Proper background selection
4. ✓ GO enrichment (BP, MF, CC)
5. ✓ KEGG pathway enrichment
6. ✓ Visualization (dotplot, enrichment map, gene-concept network)
7. ✓ Simplifying redundant terms
8. ✓ Analyzing up/down genes separately

**Key Takeaways:**
- ORA is simple and effective for most analyses
- Always use proper background (genes tested)
- GO and KEGG complement each other
- Visualize to find biological themes
- GSEA is more advanced (future topic)
:::

### Recommended Workflow

```{mermaid}
%%| fig-width: 10
flowchart TD
    A[DE Gene List] --> B[Convert to Entrez IDs]
    B --> C[Define Background]
    C --> D[Run enrichGO]
    C --> E[Run enrichKEGG]
    D --> F[Simplify GO Terms]
    F --> G[Visualize Results]
    E --> G
    G --> H{Interpret Biology}
    H --> I[Export Tables & Figures]
    H --> J[Follow-up Experiments]
```

### Resources for Learning More

**clusterProfiler Documentation:**
- Book: https://yulab-smu.top/biomedical-knowledge-mining-book/
- Vignettes: https://bioconductor.org/packages/release/bioc/html/clusterProfiler.html

**Pathway Databases:**
- GO: http://geneontology.org/
- KEGG: https://www.genome.jp/kegg/
- Reactome: https://reactome.org/
- MSigDB: https://www.gsea-msigdb.org/gsea/msigdb/

**Further Reading:**
- Reimand et al. 2019 *Nature Protocols* - Pathway enrichment best practices
- Subramanian et al. 2005 *PNAS* - Original GSEA paper
- Yu et al. 2012 *OMICS* - clusterProfiler paper

---

**Congratulations!** You've completed a full RNA-seq differential expression workflow from raw counts to biological interpretation! 🎉


## Session Information

```{r sessioninfo}
sessionInfo()
```
