---
title: "RNA-seq Differential Expression Analysis with limma-voom"
subtitle: "UF Health Cancer Institute Bioinformatics Workshop - Session 2"
author: "UFHCI BCB-SR Bioinformatics Unit"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: false
    code-tools: true
    code-overflow: wrap
    embed-resources: true
    theme: cosmo
    df-print: paged
execute:
  warning: false
  message: false
---

# Introduction

In this session, we'll work through a complete differential expression (DE) analysis using RNA-seq data. We'll be using the output from the nf-core/rnaseq pipeline (STAR alignment + RSEM quantification) that was configured in Session 1.

**Learning Objectives:**

- Load and prepare RNA-seq count data from RSEM
- Perform quality control and exploratory data analysis
- Understand filtering, normalization, and the limma-voom workflow
- Conduct differential expression analysis for simple and complex designs
- Create publication-quality visualizations
- Interpret and export results

## The Analysis Workflow

```{mermaid}
%%| fig-width: 8
flowchart TD
    A[RSEM Gene Counts] --> B[Create DGEList Object]
    B --> C[Filter Low Expression]
    C --> D[TMM Normalization]
    D --> E[Quality Control & Visualization]
    E --> F[Design Matrix]
    F --> G[voom Transformation + Quality Weights]
    G --> H[Linear Model Fitting]
    H --> I[Empirical Bayes]
    I --> J[Extract Results]
    J --> K[Visualization & Interpretation]
```

# Setup

## Load Required Libraries

```{r setup}
# Load required packages
library(limma)       # Differential expression analysis
library(edgeR)       # RNA-seq data structures and filtering
library(tidyverse)   # Data manipulation and visualization
library(ggrepel)     # Better text labels on plots
library(pheatmap)    # Heatmaps
library(RColorBrewer) # Color palettes
library(here)

# Set a seed for reproducibility
set.seed(42)

# Define directory structure using here() for portability
# here() finds the project root automatically
results_dir <- here("demo-analysis", "output", "differential-expression", "results")
figures_dir <- here("demo-analysis", "output", "differential-expression", "figures")

dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(figures_dir, showWarnings = FALSE, recursive = TRUE)

# Display the project root for verification
cat("Project root:", here(), "\n")
cat("Results directory:", results_dir, "\n")
cat("Figures directory:", figures_dir, "\n")
```

::: {.callout-tip}
## Package Installation
If you haven't installed these packages yet, you can do so with:
```r
install.packages(c("tidyverse", "ggrepel", "pheatmap", "RColorBrewer"))
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(c("limma", "edgeR"))
```
:::

## Set File Paths

```{r paths}
# Define paths to your data files
# These should be outputs from the nf-core/rnaseq pipeline
counts_file <- "/blue/bioinf_workshop/share/nfcore_rnaseq_files/star_rsem/rsem.merged.gene_counts.tsv"
metadata_file <- here("demo-analysis", "data", "metadata", "sample_metadata.csv")

if (!file.exists(counts_file))   stop("Count file not found: ",    counts_file)
if (!file.exists(metadata_file)) stop("Metadata file not found: ", metadata_file)

cat("✓ Count file found\n")
cat("✓ Metadata file found\n")
```

::: {.callout-note}
## nf-core Output Structure
The nf-core/rnaseq pipeline with RSEM produces a merged count matrix file typically named:
`rsem.merged.gene_counts.tsv` in the `star_rsem/` results directory. This file contains raw counts for all samples across all genes.
:::

# Part 1: Data Loading and Preparation

## Load Gene Count Data

```{r load-counts}
# Load the RSEM gene counts
# The nf-core pipeline produces a merged file with:
# - gene_id: Ensembl gene ID
# - gene_name: Gene symbol
# - One column per sample with count data

counts_raw <- read.table(counts_file, 
                         header = TRUE, 
                         sep = "\t",
                         check.names = FALSE, 
                         stringsAsFactors = FALSE) %>%
  dplyr::select(-any_of("transcript_id(s)"))

cat("Genes:", nrow(counts_raw), "| Samples:", ncol(counts_raw) - 1, "\n")
head(counts_raw[, 1:4], n = 3)
```

::: {.callout-important}
## RSEM vs Salmon
Note that RSEM produces **expected counts** which are suitable for downstream analysis with limma-voom or DESeq2. These are not the same as TPM values - make sure you're using the `.genes.results` counts, not TPM values!
:::

### Convert Ensembl IDs to Gene Symbols

The count matrix uses Ensembl IDs (e.g. `ENSMUSG00000000003`) as gene
identifiers. We convert these to gene symbols using the mouse annotation
package `org.Mm.eg.db` so that results are easier to interpret downstream.
Genes with no matching symbol retain their Ensembl ID as a fallback.

```{r convert-ids}
library(org.Mm.eg.db)

symbols <- AnnotationDbi::mapIds(org.Mm.eg.db, keys = counts_raw$gene_id,
                                 column = "SYMBOL", keytype = "ENSEMBL",
                                 multiVals = "first")

counts_raw$gene_name <- ifelse(is.na(symbols), counts_raw$gene_id, symbols)

cat("✓ Mapped", sum(!is.na(symbols)), "of", length(symbols), "genes to symbols\n")
```

### Build Final Count Matrix

A small number of genes share the same symbol (e.g. pseudogenes). We make
these unique by appending the Ensembl ID, then set gene symbols as row names
to produce the final counts matrix used throughout the rest of the workshop.

```{r build-matrix}
dups <- duplicated(counts_raw$gene_name) | duplicated(counts_raw$gene_name, fromLast = TRUE)
counts_raw$gene_name[dups] <- paste0(counts_raw$gene_name[dups], "_", counts_raw$gene_id[dups])

expression_data <- counts_raw %>%
  dplyr::select(-gene_id) %>%
  column_to_rownames("gene_name")

cat("✓ Final matrix:", nrow(expression_data), "genes x", ncol(expression_data), "samples\n")
head(expression_data[, 1:4], n = 3)
```


## Load Sample Metadata

```{r load-metadata}
# Load sample information
# This file should contain:
# - sample: sample names (matching column names in expression_data)
# - Any experimental factors (treatment, genotype, time, etc.)

sample_info <- read.csv(metadata_file,
                         header           = TRUE,
                         stringsAsFactors = FALSE)

sample_info

# Confirm sample names match count matrix columns
all(sample_info$sample %in% colnames(expression_data))
```

::: {.callout-tip}
## Sample Metadata Best Practices
Your sample metadata file should include:
- Clear, consistent sample names matching your count data
- All experimental factors (treatment, batch, individual, etc.)
- Any known technical variables (sequencing run, library prep date)
- Keep a data dictionary explaining what each column means!
:::

# Part 2: Quality Control and Exploratory Analysis

## Create DGEList Object

The `DGEList` object from edgeR is a convenient container for RNA-seq data that bundles counts, sample information, and gene annotations together.

```{r create-dge}
# Create DGEList object
# This combines count data with sample metadata
DGE <- DGEList(counts = expression_data,
               samples = sample_info)

# Explicitly add group column to DGE
DGE$samples$group <- factor(DGE$samples$genotype, levels = c("PRMT7_WT", "PRMT7_KD"))

# Inspect the object
DGE
```

## Filter Lowly Expressed Genes

::: {.callout-note}
## Why Filter Lowly Expressed Genes?

Not all genes are expressed in your samples. Filtering out genes with very low or zero counts:

- **Reduces noise** in the analysis
- **Improves statistical power** (fewer tests = less multiple testing burden)
- **Speeds up computation**
:::

```{r filter-genes}
# The filterByExpr function determines which genes have 
# sufficient counts for reliable DE testing
# It uses the experimental design (group) to determine 
# the minimum number of samples that should have adequate counts

keep.exprs <- filterByExpr(DGE, group = DGE$samples$group)

# How many genes are we keeping?
table(keep.exprs)

# Filter the DGEList
DGE <- DGE[keep.exprs, , keep.lib.sizes = FALSE]

cat("After filtering, we have", nrow(DGE), "genes remaining\n")
```

::: {.callout-note}
## Understanding filterByExpr
`filterByExpr()` uses a sensible heuristic: a gene should be expressed at a minimum level in at least some proportion of samples from the smallest group. By default:
- Minimum count: 10 reads
- Minimum proportion: Depends on library sizes and design
This is more sophisticated than simple count cutoffs!
:::

## Library Size and Composition

Before normalization, let's check library sizes (total read counts per sample).

```{r library-sizes}
# Calculate library sizes
lib_sizes <- DGE$samples$lib.size

# Create a data frame for plotting
lib_size_df <- data.frame(
  sample = rownames(DGE$samples),
  lib_size = lib_sizes,
  group = DGE$samples$group
)

# Plot library sizes
ggplot(lib_size_df, aes(x = sample, y = lib_size / 1e6, fill = group)) +
  geom_col() +
  labs(title = "Library Sizes After Filtering",
       x = "Sample",
       y = "Library Size (Millions of Reads)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Set2")
```

::: {.callout-warning}
## What to Look For
- **Large differences** (>2-3x) in library sizes might indicate technical issues
- **Very low libraries** (<5M reads for human/mouse) may have insufficient depth
- Some variation is normal and will be handled by normalization
:::

## TMM Normalization

Trimmed Mean of M-values (TMM) normalization accounts for:

1. **Library size differences** - some samples have more total reads
2. **Composition bias** - a few highly expressed genes can "steal" reads from other genes

```{r normalize}
# Calculate normalization factors
DGE <- calcNormFactors(DGE, method = "TMM")

# Check normalization factors
# Values close to 1 indicate little composition bias
DGE$samples
```

## Multidimensional Scaling (MDS) Plot

::: {.callout-note}
## Understanding MDS Plots

MDS (Multidimensional Scaling) is similar to PCA and shows overall similarity between samples. 

**What to look for:**
- **Close together**: Samples are similar (should be true for biological replicates)
- **Far apart**: Samples are different (expected for different conditions)
- **Expected patterns**: 
  - Biological replicates should cluster
  - Different conditions/genotypes should separate
- **Unexpected patterns**: 
  - Outliers far from their group
  - Poor separation between conditions
  - Batch effects (samples clustering by prep date, not biology)
:::

```{r mds-plot, fig.width=8, fig.height=6}
# Calculate MDS
mds <- plotMDS(DGE, top = 500, plot = FALSE, gene.selection = "common")

# Create a data frame for ggplot
mds_df <- data.frame(
  sample = rownames(DGE$samples),
  dim1 = mds$x,
  dim2 = mds$y, 
  genotype = DGE$samples$genotype, 
  treatment = DGE$samples$treatment, 
  Run = DGE$samples$Run
)

# Create MDS plot
ggplot(mds_df, aes(x = dim1, y = dim2, color = genotype, label = Run)) +
  geom_point(size = 4, alpha = 0.8) +
  geom_text_repel(size = 3, max.overlaps = 20) +
  labs(title = "MDS Plot - Sample Similarity",
       subtitle = "Based on top 500 most variable genes",
       x = paste0("Leading logFC dim 1 (", round(mds$var.explained[1] * 100, 1), "%)"),
       y = paste0("Leading logFC dim 2 (", round(mds$var.explained[2] * 100, 1), "%)")) +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_color_brewer(palette = "Set1")

# Save the plot
ggsave(here("demo-analysis", "output", "differential-expression", "figures", "mds_plot.png"), width = 8, height = 6, dpi = 300)
```

::: {.callout-important}
## Interpreting the MDS Plot
**Good signs:**
- Samples from the same group cluster together
- Clear separation between different conditions
- No obvious outliers

**Red flags:**
- Outlier samples far from their group
- No separation between conditions (weak signal)
- Samples clustering by batch instead of biology
:::

## Sample Correlation Heatmap

A complementary view showing pairwise sample correlations.

```{r correlation-heatmap, fig.width=8, fig.height=7}
# Get log-CPM values for visualization
logcpm <- cpm(DGE, log = TRUE)

# Calculate correlation matrix
cor_matrix <- cor(logcpm, method = "spearman")

# Create annotation for samples
annotation_col <- data.frame(
  Genotype = DGE$samples$genotype,
  row.names = colnames(logcpm)
)

# Create heatmap
pheatmap(cor_matrix,
         annotation_col = annotation_col,
         annotation_row = annotation_col,
         main = "Sample Correlation Heatmap",
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(0.7, 1, length.out = 101),
         fontsize = 8,
         filename = here("demo-analysis", "output", "differential-expression", "figures", "correlation_heatmap.png"),
         width = 8,
         height = 7)

# Also display in the document
pheatmap(cor_matrix,
         annotation_col = annotation_col,
         annotation_row = annotation_col,
         main = "Sample Correlation Heatmap",
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(0.7, 1, length.out = 101),
         fontsize = 8)
```

::: {.callout-tip}
## Correlation Expectations
- **Within-group correlation**: typically 0.95-0.99
- **Between-group correlation**: depends on how different conditions are (0.85-0.98)
- **Outliers**: samples with consistently lower correlations (<0.90) to all others
:::

# Part 3: Differential Expression - Simple Two-Group Design

Let's start with a simple comparison: two conditions, like "treated vs. control" or "mutant vs. wildtype".

## Create Design Matrix

The design matrix tells limma which samples belong to which groups.

```{r design-simple}
# For our PRMT7 knockdown experiment, we're comparing KD to WT
# The first level (PRMT7_WT) will be the reference

# Make sure the factor levels are set correctly
# The first level will be the reference (denominator in fold-change)
DGE$samples$genotype <- factor(DGE$samples$genotype, 
                                levels = c("PRMT7_WT", "PRMT7_KD"))

# Create design matrix
design <- model.matrix(~genotype, data = DGE$samples)

# View the design matrix
design

# Column names tell us what each coefficient represents
colnames(design)
```

::: {.callout-note}
## Understanding the Design Matrix
- **Intercept (column 1)**: Average expression in the reference group (PRMT7_WT)
- **genotypePRMT7_KD (column 2)**: Difference between KD and WT (log2 fold-change)

If a gene has:
- Positive coefficient: higher expression in KD
- Negative coefficient: higher expression in WT (lower in KD)
:::

## Run limma-voom with Quality Weights

::: {.callout-note}
## The voom Transformation

voom performs three key steps:

1. **Converts counts to log2-CPM** - makes data more normally distributed
2. **Estimates mean-variance relationship** - models how variability changes with expression
3. **Calculates precision weights** - genes with stable expression get higher weight

This transformation allows us to use powerful linear model tools designed for continuous data!
:::

The voom transformation:


```{r voom-simple, fig.width=8, fig.height=5}
# Run voom with quality weights
# normalize.method = "none" because we already did TMM normalization
v <- voom(DGE, 
          design = design,
          normalize.method = "none",
          plot = TRUE)
```

::: {.callout-note}
## Interpreting the voom Plot

The plot shows the relationship between average expression and variability:

**Top panel (mean-variance trend):**
- **X-axis**: Average expression (log2-CPM)
- **Y-axis**: Square root of variance
- **Red curve**: The mean-variance trend being modeled
- **Black points**: Individual genes

**Pattern to expect:**
- **Low expression** (left): High variance (noisy measurements)
- **High expression** (right): Lower, more stable variance
- Smooth declining trend is normal

**What voom does:** Uses this relationship to calculate precision weights for each gene, giving more weight to reliably measured genes.

**Warning signs:**
- Outlier points far from curve (potential problem genes)
- Increasing variance at high expression (unusual)
:::

::: {.callout-tip}
## voom vs. voomWithQualityWeights

We're using basic `voom()` which handles gene-level variance. If you have concerns about outlier samples or variable sample quality, you can use `voomWithQualityWeights()` instead, which also calculates sample-level weights to down-weight problematic samples.
:::

## Fit Linear Model

```{r fit-simple}
# Fit linear model to each gene
fit <- lmFit(v, design)


# Apply empirical Bayes moderation
# This borrows information across genes to improve variance estimates
ffit <- eBayes(fit, robust = TRUE)

# Quick summary of DE genes at FDR < 0.05
# Down = lower in KD, Up = higher in KD
summary(decideTests(fit, adjust.method = "fdr", p.value = 0.05, lfc = 1))
```

::: {.callout-tip}
## Understanding decideTests Output
- **Down**: Genes significantly downregulated in KD vs. WT (negative logFC = lower in KD)
- **NotSig**: Genes not significantly different between genotypes
- **Up**: Genes significantly upregulated in KD vs. WT (positive logFC = higher in KD)
:::

## Extract Top Results

```{r top-results}
# Get the top differentially expressed genes
# coef = 2 refers to the treatment effect (second column of design matrix)
top_genes_strict <- topTable(fit, 
                             coef = 2, 
                             adjust.method = "BH", 
                             number = 20, 
                             sort.by = "P", 
                             lfc = 1)
cat("\nTop 20 genes (FDR < 0.05, |logFC| > 1):\n")
print(top_genes_strict)

cat("Total genes tested:    ", nrow(top_genes_strict), "\n")
cat("DE genes (FDR < 0.05):", sum(top_genes_strict$adj.P.Val < 0.05), "\n")
cat("Up in KD:             ", sum(top_genes_strict$adj.P.Val < 0.05 & top_genes_strict$logFC > 0), "\n")
cat("Down in KD:           ", sum(top_genes_strict$adj.P.Val < 0.05 & top_genes_strict$logFC < 0), "\n\n")

# Top 20 by p-value
topTable(fit, coef = 2, adjust.method = "BH", number = 20, sort.by = "P", lfc = 1)

```

## Volcano Plot

A volcano plot shows log-fold change vs. statistical significance.

```{r volcano-simple, fig.width=8, fig.height=6}
# Extract all results
all_results <- topTable(fit,
                        coef = 2,
                        adjust.method = "BH",
                        number = Inf,
                        sort.by = "none")

all_results$category <- case_when(
  all_results$adj.P.Val < 0.05 & all_results$logFC >  1 ~ "Up (>2-fold)",
  all_results$adj.P.Val < 0.05 & all_results$logFC < -1 ~ "Down (>2-fold)",
  all_results$adj.P.Val < 0.05                           ~ "Significant",
  TRUE                                                    ~ "Not Significant"
)

table(all_results$category)

ggplot(all_results, aes(x = logFC, y = -log10(P.Value), color = category)) +
  geom_point(alpha = 0.5, size = 1) +
  scale_color_manual(values = c("Up (>2-fold)" = "red",
                                "Down (>2-fold)" = "blue",
                                "Significant" = "orange",
                                "Not Significant" = "grey80")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", alpha = 0.5) +
  labs(title = "Volcano Plot: PRMT7 KD vs WT",
       subtitle = "Thresholds: FDR < 0.05, |logFC| > 1 (2-fold change)",
       x = "Log2 Fold Change",
       y = "-Log10 P-value",
       color = "Category") +
  theme_minimal() +
  theme(legend.position = "right")

ggsave(here(figures_dir, "volcano_plot.png"), width = 8, height = 6, dpi = 300)
```

## MA Plot

An MA plot shows log-fold change vs. average expression.

```{r ma-plot, fig.width=8, fig.height=6}
# Create MA plot
ggplot(all_results, aes(x = AveExpr, y = logFC, color = category)) +
  geom_point(alpha = 0.5, size = 1) +
  scale_color_manual(values = c("Up (>2-fold)" = "red",
                                "Down (>2-fold)" = "blue",
                                "Significant" = "orange",
                                "Not Significant" = "grey80")) +
  geom_hline(yintercept = 0, linetype = "solid", alpha = 0.5) +
  geom_hline(yintercept = c(-1, 1), linetype = "dashed", alpha = 0.3) +
  labs(title = "MA Plot: PRMT7 KD vs WT",
       subtitle = "Thresholds: FDR < 0.05, |logFC| > 1 (2-fold change)",
       x = "Average Expression (log2-CPM)",
       y = "Log2 Fold Change",
       color = "Category") +
  theme_minimal()

ggsave(here(figures_dir, "ma_plot.png"), width = 8, height = 6, dpi = 300)
```

## Heatmap of Top DE Genes

```{r heatmap-top-genes, fig.width=8, fig.height=10}
# Get top 50 genes from simple treatment comparison
top50        <- rownames(topTable(fit, coef = 2, number = 50, lfc = 1))
top50_logcpm <- cpm(DGE[top50, ], log = TRUE)
top50_scaled <- t(scale(t(top50_logcpm)))  # Z-score per gene

annotation_col <- data.frame(
  Genotype  = DGE$samples$genotype,
  row.names = colnames(top50_logcpm)
)

# Display in document
pheatmap(top50_scaled,
         annotation_col = annotation_col,
         cluster_rows   = TRUE,
         cluster_cols   = TRUE,
         show_rownames  = TRUE,
         show_colnames  = TRUE,
         main           = "Top 50 DE Genes",
         fontsize_row   = 6,
         fontsize_col   = 8,
         color          = colorRampPalette(c("blue", "white", "red"))(100))

# Save to file
pheatmap(top50_scaled,
         annotation_col = annotation_col,
         cluster_rows   = TRUE,
         cluster_cols   = TRUE,
         show_rownames  = TRUE,
         show_colnames  = TRUE,
         main           = "Top 50 DE Genes",
         fontsize_row   = 6,
         fontsize_col   = 8,
         color          = colorRampPalette(c("blue", "white", "red"))(100),
         filename       = here(figures_dir, "heatmap_top50.png"),
         width          = 8,
         height         = 10)
```

## Export Results Tables

```{r export-contrasts}
all_results$gene <- rownames(all_results)

write.csv(all_results,
          here(results_dir, "de_results_all.csv"),
          row.names = FALSE)

# Significant genes only
sig_results <- all_results %>% filter(adj.P.Val < 0.05 & abs(logFC) > 1)

write.csv(sig_results,
          here(results_dir, "de_results_significant.csv"),
          row.names = FALSE)

cat("✓ de_results_all.csv:        ", nrow(all_results), "genes\n")
cat("✓ de_results_significant.csv:", nrow(sig_results), "genes\n")
```

# Summary and Key Takeaways

## What We Covered

1. **Data Loading**: RSEM counts from nf-core pipeline
2. **Quality Control**: Library sizes, MDS plots, sample correlations
3. **Filtering & Normalization**: Removing low counts, TMM normalization
4. **Simple Design**: Two-group comparison with limma-voom
5. **Visualization**: Volcano plots, MA plots, heatmaps

## Key Concepts to Remember

::: {.callout-important}
## Critical Points

1. **Filter before normalizing** - remove genes with insufficient counts
2. **TMM normalization** handles library size and composition bias
3. **voom transformation** converts counts to log-scale with precision weights
4. **Design matrix** defines what you're testing
5. **Contrasts** let you ask specific biological questions
6. **Multiple testing correction** (FDR/BH) is essential
7. **LogFC and p-value together** tell the full story (effect size + significance)
:::

## Next Steps

- **Gene Set Enrichment Analysis**: Are DE genes enriched for specific pathways/functions?
- **Validation**: qRT-PCR or additional experiments to confirm key findings
- **Biological Interpretation**: Connect results back to your biological question
- **Reproducibility**: Document your analysis, save your code, version control!

## Resources

- [limma User's Guide](https://bioconductor.org/packages/release/bioc/vignettes/limma/inst/doc/usersguide.pdf)
- [edgeR User's Guide](https://bioconductor.org/packages/release/bioc/vignettes/edgeR/inst/doc/edgeRUsersGuide.pdf)
- [RNA-seq analysis in R (Harvard Chan Bioinformatics Core)](https://hbctraining.github.io/DGE_workshop/)
- UF Research Computing: [https://www.rc.ufl.edu/](https://www.rc.ufl.edu/)

---

# Part 4: Pathway Enrichment Analysis

::: {.callout-note}
## From Genes to Biology

We've identified thousands of differentially expressed genes, but what does this mean biologically? 

**Pathway enrichment analysis** helps us answer:
- What biological processes are affected?
- What pathways are dysregulated?
- Are there common themes among DE genes?

This transforms gene lists into biological insights!
:::

## Conceptual Overview: Over-Representation Analysis

**The Basic Question**: Are my DE genes **enriched** in certain pathways/processes compared to what we'd expect by chance?

::: {.columns}
::: {.column width="50%"}
**Example:**

- Total genes in genome: 20,000
- Genes in "DNA Repair" pathway: 200 (1%)
- Your DE genes: 1,000
- Expected in DNA Repair: 10 genes (1% of 1,000)
- **Actual in DNA Repair: 50 genes**
:::

::: {.column width="50%"}
**Statistical Test:**

Is 50 significantly different from 10?

→ Use **hypergeometric test** (Fisher's exact test)

If p < 0.05: "DNA Repair is enriched!"

**Biological interpretation**: Your treatment affects DNA repair processes
:::
:::

::: {.callout-warning}
## Important Assumption

ORA assumes all DE genes are equally important. It treats a gene with logFC = 5 (32-fold change) the same as logFC = 0.6 (1.5-fold change).

**GSEA** (Gene Set Enrichment Analysis) addresses this by using ranked gene lists, but it's more complex. We'll mention it at the end.
:::

## Setup: Load Required Packages

```{r load-pathway-packages}
# Install if needed (uncomment):
# BiocManager::install(c("clusterProfiler", "org.Mm.eg.db", "enrichplot", "pathview"))

library(clusterProfiler)
library(org.Mm.eg.db)   # Mouse annotation — use org.Hs.eg.db for human
library(enrichplot)
library(tidyverse)
library(patchwork)
library(here)

set.seed(42)

figures_dir <- here("demo-analysis", "output", "differential-expression", "figures")

dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(figures_dir, showWarnings = FALSE, recursive = TRUE)

# Check organism database
org.Mm.eg.db
```

::: {.callout-tip}
## Species-Specific Annotation Packages

- **Human**: `org.Hs.eg.db`
- **Mouse**: `org.Mm.eg.db`
- **Rat**: `org.Rn.eg.db`
- **Fly**: `org.Dm.eg.db`
- **Zebrafish**: `org.Dr.eg.db`
- **C. elegans**: `org.Ce.eg.db`

Install the appropriate one for your species!
:::

## Prepare Gene Lists

We'll use the limma-voom results from Part 3 (the simple PRMT7_KD vs WT comparison).

```{r prepare-gene-lists}
all_results <- read.csv(here(results_dir, "de_results_all.csv"))
rownames(all_results) <- all_results$gene

sig_genes  <- all_results %>% filter(adj.P.Val < 0.05 & abs(logFC) > 1)
up_genes   <- sig_genes  %>% filter(logFC > 1)
down_genes <- sig_genes  %>% filter(logFC < 1)

cat("Total DE genes (FDR < 0.05) & FC > 2x:", nrow(sig_genes), "\n")
cat("Upregulated in KD:          ", nrow(up_genes), "\n")
cat("Downregulated in KD:        ", nrow(down_genes), "\n")
```

## Convert Gene Symbols to Entrez IDs

::: {.callout-warning}
## Gene ID Conversion Issues

Not all genes convert successfully:

- **Novel genes**: Not yet in annotation databases
- **Non-coding RNAs**: May lack Entrez IDs  
- **Pseudogenes**: Often not annotated
- **Outdated symbols**: Gene nomenclature changes

**Typical conversion rate**: 85-95%

This is normal and expected - don't worry if 5-15% fail to convert!
:::

```{r convert-to-entrez}
# Get all gene symbols from our results
gene_entrez <- bitr(rownames(sig_genes),
                    fromType = "SYMBOL",
                    toType   = "ENTREZID",
                    OrgDb    = org.Mm.eg.db)

# Background: all genes tested
background_entrez <- bitr(rownames(all_results),
                           fromType = "SYMBOL",
                           toType   = "ENTREZID",
                           OrgDb    = org.Mm.eg.db)

universe <- background_entrez$ENTREZID

cat("DE genes submitted:     ", nrow(sig_genes), "\n")
cat("Successfully converted: ", nrow(gene_entrez), "\n")
cat("Background genes:       ", length(universe), "\n")
```


## Background Gene Set

::: {.callout-important}
## Why Background Matters

**Wrong approach**: Compare to all genes in genome (~20,000)

**Right approach**: Compare to genes expressed in your experiment (~13,000)

Using the correct background prevents:
- **False positives** from untested genes
- **Biased enrichment** from tissue-specific expression

**Always use your filtered gene set as background!**
:::

For accurate enrichment testing, we need to define the **background** (universe) - all genes that COULD have been detected:

```{r prepare-background}
# All genes that passed filtering and were tested
background_symbols <- rownames(all_results)

# Convert background to Entrez IDs
background_entrez <- bitr(background_symbols,
                          fromType = "SYMBOL",
                          toType = "ENTREZID",
                          OrgDb = org.Mm.eg.db)

cat("Background genes:", nrow(background_entrez), "\n")

# Extract just the Entrez IDs
universe <- background_entrez$ENTREZID
```


## Gene Ontology (GO) Enrichment

GO provides a standardized vocabulary for gene function across three domains:

- **Biological Process (BP)**: What the gene does (e.g., "DNA repair", "cell cycle")
- **Molecular Function (MF)**: Biochemical activity (e.g., "kinase activity", "DNA binding")
- **Cellular Component (CC)**: Where it acts (e.g., "nucleus", "mitochondrion")

### Biological Process Enrichment

```{r up-vs-down}
up_entrez <- gene_entrez %>%
  filter(SYMBOL %in% rownames(up_genes)) %>%
  pull(ENTREZID)

down_entrez <- gene_entrez %>%
  filter(SYMBOL %in% rownames(down_genes)) %>%
  pull(ENTREZID)

```


```{r enrichGO-BP}
# Run GO enrichment for Biological Process
ego_up <- enrichGO(gene         = up_entrez,
                   universe     = universe,
                   OrgDb        = org.Mm.eg.db,
                   ont          = "BP",
                   pvalueCutoff = 0.05,
                   readable     = TRUE)

ego_down <- enrichGO(gene         = down_entrez,
                     universe     = universe,
                     OrgDb        = org.Mm.eg.db,
                     ont          = "BP",
                     pvalueCutoff = 0.05,
                     readable     = TRUE)

cat("GO terms — upregulated:  ", nrow(ego_up), "\n")
cat("GO terms — downregulated:", nrow(ego_down), "\n")
```

::: {.callout-tip}
## Understanding the Output

Key columns:

- **Description**: GO term name
- **GeneRatio**: X/Y = X genes in your list / Y genes in term
- **BgRatio**: A/B = A genes in term / B genes in background
- **pvalue**: Hypergeometric test p-value
- **p.adjust**: FDR-corrected p-value
- **Count**: Number of your genes in this term
- **geneID**: Which genes (gene symbols if readable=TRUE)

**Rich Factor** = GeneRatio / BgRatio  
→ Higher = more enriched
:::

### Visualize GO Results: Dot Plot

```{r dotplot-GO, fig.width=10, fig.height=8}
p1 <- dotplot(ego_up,   showCategory = 10, title = "Upregulated in KD") +
      theme(axis.text.y = element_text(size = 8))

p2 <- dotplot(ego_down, showCategory = 10, title = "Downregulated in KD") +
      theme(axis.text.y = element_text(size = 8))

p1 | p2

ggsave(here(figures_dir, "GO_up_vs_down.png"),
       width = 12, height = 8, dpi = 300)
```

::: {.callout-note}
## Reading the Dot Plot

- **X-axis**: Gene Ratio (proportion of genes in term)
- **Y-axis**: GO terms (sorted by p-value)
- **Dot size**: Number of genes (Count)
- **Dot color**: Adjusted p-value (darker = more significant)

**Look for**:
- Large dots = many genes
- Dark colors = highly significant
- High GeneRatio = large proportion of term covered
:::

Shows which specific genes belong to which GO terms. Brown nodes are GO terms,
colored nodes are genes, and hub genes appear in multiple terms.

Gene-concept networks show which genes belong to which terms. We can color the genes by their fold-change to see expression patterns.
```{r prepare-foldchange-vectors}
# Extract fold changes for coloring
# Need a named vector: names = gene symbols, values = logFC

# For upregulated genes
up_gene_symbols <- rownames(up_genes)
up_foldchanges <- up_genes$logFC
names(up_foldchanges) <- up_gene_symbols

# For downregulated genes
down_gene_symbols <- rownames(down_genes)
down_foldchanges <- down_genes$logFC
names(down_foldchanges) <- down_gene_symbols

# Check the format
head(up_foldchanges)
head(down_foldchanges)

# Check ranges
cat("Upregulated logFC range:", 
    round(min(up_foldchanges), 2), "to", 
    round(max(up_foldchanges), 2), "\n")
cat("Downregulated logFC range:", 
    round(min(down_foldchanges), 2), "to", 
    round(max(down_foldchanges), 2), "\n")
```

### Visualize GO Results: Gene-Concept Network

Shows which genes belong to which terms:

```{r cnetplot, fig.width=12, fig.height=10}
p_up <- cnetplot(ego_up, 
                 showCategory = 5, 
                 foldChange = up_foldchanges, 
                 node_label = "all", 
                 cex_label_gene = 0.7, 
                 cex_label_category = 1.2, 
                 colorEdge = TRUE) + 
  scale_color_gradient(low = "white",
                       high = "red",
                       name = "Log2FC\n(Upregulated)") +
  labs(title = "Gene-Concept Network: Upregulated DEGS - Top 5 GO Terms") + 
  theme(legend.position = "right")

# Display the plot
print(p_up)

# Save the plot
ggsave(here(figures_dir, "GO_up_cnetplot.png"), plot = p_up, 
       width = 12, height = 10, dpi = 300)

p_down <- cnetplot(ego_down, 
                   showCategory = 5, 
                   foldChange = down_foldchanges, 
                   node_label = "all", 
                   cex_label_gene = 0.7, 
                   cex_label_category = 1.2, 
                   colorEdge = TRUE) + 
  scale_color_gradient(low = "white",
                       high = "blue",
                       name = "Log2FC\n(Downregulated)") +
  labs(title = "Gene-Concept Network: Downregulated DEGs - Top 5 GO Terms") + 
  theme(legend.position = "right")

# Display the plot
print(p_down)

# Save the plot
ggsave(here(figures_dir, "GO_down_cnetplot.png"), plot = p_down,
       width = 12, height = 10, dpi = 300)
```

::: {.callout-note}
## Gene-Concept Networks

- **Brown nodes**: GO terms
- **Colored nodes**: Genes (color = fold change if provided)
- **Edges**: Gene belongs to term

**Useful for**:
- Seeing which genes drive enrichment
- Identifying "hub genes" (in multiple terms)
- Understanding term overlap
:::

### Simplify Redundant Terms

GO is hierarchical and many terms overlap. Simplification removes redundant
terms while keeping the most significant representative from each cluster.

```{r go-simplify}
# Simplify GO results by removing redundant terms
ego_up_simple <- clusterProfiler::simplify(ego_up,
                          cutoff = 0.7,      # Similarity threshold
                          by = "p.adjust",   # Keep most significant
                          select_fun = min)

cat("Original GO terms:", nrow(ego_up), "\n")
cat("After simplification:", nrow(ego_up_simple), "\n\n")

# Compare
cat("Top 10 simplified terms:\n")
head(ego_up_simple, 10)
```

```{r go-simplified-dotplot, fig.width=10, fig.height=6}
dotplot(ego_up_simple,
        showCategory = 10,
        title        = "Simplified GO-BP Enrichment")

ggsave(here(figures_dir, "GO_BP_simplified_dotplot.png"),
       width = 10, height = 6, dpi = 300)
```       
       
## KEGG Pathway Enrichment

KEGG provides curated pathway maps for metabolism, signaling, disease, etc.

```{r enrichKEGG}
# KEGG uses organism codes: "mmu" = mouse, "hsa" = human
kegg_up <- enrichKEGG(gene = up_entrez,
                      universe = universe, 
                      organism = "mmu",      # Mouse
                      pAdjustMethod = "BH",
                      pvalueCutoff = 0.05,
                      qvalueCutoff = 0.05)

kegg_down <- enrichKEGG(gene = down_entrez,
                        universe = universe, 
                        organism = "mmu",      # Mouse
                        pAdjustMethod = "BH",
                        pvalueCutoff = 0.05,
                        qvalueCutoff = 0.05)

# Add gene symbols (KEGG uses Entrez IDs)
kegg_up <- setReadable(kegg_up, 
                       OrgDb = org.Mm.eg.db, 
                       keyType = "ENTREZID")

kegg_down <- setReadable(kegg_down, 
                         OrgDb = org.Mm.eg.db, 
                         keyType = "ENTREZID")

# Clean pathway names - remove " - Mus musculus" suffix
kegg_up@result$Description <- gsub(" - .*$", "", kegg_up@result$Description)
kegg_down@result$Description <- gsub(" - .*$", "", kegg_down@result$Description)

# Summary
cat("Enriched KEGG pathways - Upregulated DEGs: ", nrow(kegg_up), "\n")
cat("Enriched KEGG pathways - Downregulated DEGs: ", nrow(kegg_down), "\n\n")

# Show results
head(kegg_up, 10)
```

::: {.callout-tip}
## Common KEGG Organism Codes

- **mmu**: Mouse (*Mus musculus*)
- **hsa**: Human (*Homo sapiens*)
- **rno**: Rat (*Rattus norvegicus*)
- **dme**: Fly (*Drosophila melanogaster*)
- **dre**: Zebrafish (*Danio rerio*)
- **cel**: C. elegans

Find more: https://www.genome.jp/kegg/catalog/org_list.html
:::

### Visualize KEGG Results

```{r kegg-dotplot, fig.width=10, fig.height=6}
# Dot plot
dotplot(kegg_up, 
        showCategory = 15,
        title = "KEGG Pathway Enrichment")

ggsave(here("demo-analysis", "output", "differential-expression", "figures", "KEGG_dotplot.png"), width = 10, height = 6, dpi = 300)
```

### Compare GO vs KEGG

```{r compare-databases-direction, fig.width=12, fig.height=10}
# Prepare GO data (up and down)
ego_up_df <- as.data.frame(ego_up) %>%
  mutate(Database = "GO-BP", 
         Direction = "Up",
         Term = Description) %>%
  dplyr::select(Database, Direction, Term, Count, FDR = p.adjust) %>%
  head(10)

ego_down_df <- as.data.frame(ego_down) %>%
  mutate(Database = "GO-BP", 
         Direction = "Down",
         Term = Description) %>%
  dplyr::select(Database, Direction, Term, Count, FDR = p.adjust) %>%
  head(10)

# Prepare KEGG data (up and down)
kegg_up_df <- as.data.frame(kegg_up) %>%
  mutate(Database = "KEGG", 
         Direction = "Up",
         Term = Description) %>%
  dplyr::select(Database, Direction, Term, Count, FDR = p.adjust) %>%
  head(10)

kegg_down_df <- as.data.frame(kegg_down) %>%
  mutate(Database = "KEGG", 
         Direction = "Down",
         Term = Description) %>%
  dplyr::select(Database, Direction, Term, Count, FDR = p.adjust) %>%
  head(10)

# Combine all
comparison_df <- bind_rows(ego_up_df, ego_down_df, 
                           kegg_up_df, kegg_down_df)

# Plot: facet by database, color by direction
ggplot(comparison_df, aes(x = Count, y = reorder(Term, Count), fill = Direction)) +
  geom_col() +
  scale_fill_manual(values = c("Up" = "red", "Down" = "blue")) +
  labs(title = "Top Enriched Terms: GO vs KEGG (Up and Down)",
       x = "Number of Genes", 
       y = "Term",
       fill = "Direction") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 8)) +
  facet_wrap(~Database, scales = "free_y", ncol = 1)

ggsave(here(figures_dir, "GO_vs_KEGG_comparison_direction.png"), 
       width = 12, height = 10, dpi = 300)
```

::: {.callout-note}
## GO vs KEGG: When to Use Each?

**Gene Ontology (GO):**
- ✓ Comprehensive (more terms)
- ✓ Broad biological processes
- ✓ Good for exploratory analysis
- ✗ Can be redundant (many overlapping terms)

**KEGG Pathways:**
- ✓ Curated signaling/metabolic pathways
- ✓ Pathway diagrams available
- ✓ Less redundancy
- ✗ Fewer pathways (more focused)

**Recommendation**: Run both! They complement each other.
:::



## Export Results

```{r export-pathway-results}

```{r export-pathway-results-final}
# ===== Export GO-BP Results =====
write.csv(as.data.frame(ego_up),
          here(results_dir, "GO_BP_upregulated.csv"),
          row.names = FALSE)

write.csv(as.data.frame(ego_down),
          here(results_dir, "GO_BP_downregulated.csv"),
          row.names = FALSE)

# ===== Export KEGG Results =====
write.csv(as.data.frame(kegg_up),
          here(results_dir, "KEGG_upregulated.csv"),
          row.names = FALSE)

write.csv(as.data.frame(kegg_down),
          here(results_dir, "KEGG_downregulated.csv"),
          row.names = FALSE)

# ===== Create Summary Table =====
pathway_summary <- data.frame(
  Database = rep(c("GO-BP", "KEGG"), each = 2),
  Direction = rep(c("Up", "Down"), 2),
  N_Terms = c(nrow(ego_up), nrow(ego_down), 
              nrow(kegg_up), nrow(kegg_down)),
  Top_Term = c(
    ego_up$Description[1],
    ego_down$Description[1],
    kegg_up$Description[1],
    kegg_down$Description[1]
  ),
  Top_FDR = formatC(c(
    ego_up$p.adjust[1],
    ego_down$p.adjust[1],
    kegg_up$p.adjust[1],
    kegg_down$p.adjust[1]
  ), format = "e", digits = 2)
)

write.csv(pathway_summary,
          here(results_dir, "pathway_summary.csv"),
          row.names = FALSE)

# ===== Print Summary =====
cat("\n=== Pathway Enrichment Results Exported ===\n\n")
print(pathway_summary)

cat("\nFiles saved:\n")
cat("  ✓ GO_BP_upregulated.csv (", nrow(ego_up), " terms)\n", sep = "")
cat("  ✓ GO_BP_downregulated.csv (", nrow(ego_down), " terms)\n", sep = "")
cat("  ✓ KEGG_upregulated.csv (", nrow(kegg_up), " terms)\n", sep = "")
cat("  ✓ KEGG_downregulated.csv (", nrow(kegg_down), " terms)\n", sep = "")
cat("  ✓ pathway_summary.csv\n")
cat("\nLocation:", results_dir, "\n")
```

---

## Advanced Topic: Gene Set Enrichment Analysis (GSEA)

::: {.callout-note}
## What is GSEA?

**Over-Representation Analysis (ORA)** uses a binary cutoff:
- Gene is DE (FDR < 0.05) → Include
- Gene is not DE → Exclude

**Gene Set Enrichment Analysis (GSEA)** uses ALL genes:
- Rank all genes by statistic (e.g., t-statistic, logFC)
- Test if a gene set is enriched at top or bottom of ranked list
- No arbitrary cutoff needed!

**Advantages:**
- Uses full dataset (more information)
- Detects coordinated changes (many small effects)
- No cutoff bias

**Disadvantages:**
- More complex to interpret
- Requires proper ranking metric
- Takes longer to run
:::

### When to Use GSEA vs ORA

::: {.columns}
::: {.column width="50%"}
**Use ORA when:**
- Clear DE cutoff makes sense (FDR < 0.05)
- Strong effect sizes
- Want simple interpretation
- Quick exploratory analysis
- Teaching beginners
:::

::: {.column width="50%"}
**Use GSEA when:**
- Subtle, coordinated changes
- No clear DE cutoff
- Small effect sizes across pathway
- More rigorous analysis needed
- Publishing results
:::
:::


---

## Common Pitfalls in Pathway Analysis

::: {.callout-warning}
## Mistakes to Avoid

**1. Wrong background**
❌ Using all genome genes  
✅ Use genes tested in your experiment

**2. Ignoring multiple testing**
❌ p-value < 0.05  
✅ FDR (q-value) < 0.05

**3. Over-interpreting single genes**
- One gene can appear in many terms
- Focus on biological themes, not single genes

**4. Cherry-picking results**
- Report all enriched terms, not just favorites
- Include "uninteresting" results too

**5. Ignoring effect size**
- Term with 3 genes vs 50 genes
- Both significant, but 50 is stronger evidence

**6. Not simplifying GO**
- Many redundant GO terms
- Use `simplify()` to reduce redundancy

**7. Database version matters**
- GO updates quarterly
- KEGG updates irregularly
- Document database versions!
:::

## Summary and Next Steps

# Workshop Summary

::: {.callout-important}
## What You Learned Today

**Data Preparation:**
✓ Loading and preparing RSEM count data  
✓ Gene ID conversion and annotation  
✓ Creating DGEList objects  

**Quality Control:**
✓ Filtering lowly expressed genes  
✓ TMM normalization  
✓ MDS plots for sample QC  
✓ Correlation heatmaps  

**Differential Expression:**
✓ limma-voom workflow  
✓ edgeR exact test (alternative approach)  
✓ Design matrices and contrasts  
✓ Multiple testing correction  

**Pathway Analysis:**
✓ GO enrichment (Biological Process)  
✓ KEGG pathway analysis  
✓ Analyzing up/down genes separately  
✓ Gene-concept networks with fold-change colors  

**Reproducibility:**
✓ Using `here()` for portable paths  
✓ Documenting with sessionInfo()  
✓ Creating publication-quality figures  
✓ Exporting results tables  

**Next Steps:**
- Apply these methods to your own data
- Explore GO Molecular Function and Cellular Component
- Try GSEA for ranked gene analysis
- Consider batch correction if needed
- Validate top genes experimentally!
:::

Good luck with your analyses! 🎉

### Recommended Workflow

```{mermaid}
%%| fig-width: 10
flowchart TD
    A[DE Gene List] --> B[Convert to Entrez IDs]
    B --> C[Define Background]
    C --> D[Run enrichGO]
    C --> E[Run enrichKEGG]
    D --> F[Simplify GO Terms]
    F --> G[Visualize Results]
    E --> G
    G --> H{Interpret Biology}
    H --> I[Export Tables & Figures]
    H --> J[Follow-up Experiments]
```

### Resources for Learning More

**clusterProfiler Documentation:**
- Book: https://yulab-smu.top/biomedical-knowledge-mining-book/
- Vignettes: https://bioconductor.org/packages/release/bioc/html/clusterProfiler.html

**Pathway Databases:**
- GO: http://geneontology.org/
- KEGG: https://www.genome.jp/kegg/
- Reactome: https://reactome.org/
- MSigDB: https://www.gsea-msigdb.org/gsea/msigdb/

**Further Reading:**
- Reimand et al. 2019 *Nature Protocols* - Pathway enrichment best practices
- Subramanian et al. 2005 *PNAS* - Original GSEA paper
- Yu et al. 2012 *OMICS* - clusterProfiler paper

---

**Congratulations!** You've completed a full RNA-seq differential expression workflow from raw counts to biological interpretation! 🎉


## Session Information

```{r sessioninfo}
sessionInfo()
```
