---
title: "Optional: edgeR Exact Test and GREIN Reproducibility"
subtitle: "UF Health Cancer Institute Bioinformatics Workshop"
author: "UFHCI BCB-SR Bioinformatics Unit"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: false
    code-tools: true
    code-overflow: wrap
    embed-resources: true
    theme: cosmo
    df-print: paged
execute:
  warning: false
  message: false
---

# Introduction

This notebook explores an alternative analysis approach using the edgeR exact
test and compares results to the limma-voom analysis completed in
`03_differential_expression.qmd`.

It also uses this comparison as a case study in reproducibility — specifically,
how to reverse-engineer an analysis from a public database (GREIN) when the
methods documentation is incomplete.

**Requires:** `DGE_filtered_normalized.rds` and `de_results_all.csv` produced
by `02_quality_control.qmd` and `03_differential_expression.qmd`.

---

# Setup

```{r setup}
library(limma)
library(edgeR)
library(tidyverse)
library(here)

set.seed(42)

results_dir <- here("demo-analysis", "output", "02-differential-expression", "results")
figures_dir <- here("demo-analysis", "output", "02-differential-expression", "figures")

dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(figures_dir, showWarnings = FALSE, recursive = TRUE)
```

---

# Load Data

Load the normalized DGEList from Part 1 and the limma-voom results from Part 2.

```{r load-data}
DGE <- readRDS(here("demo-analysis", "output", "02-differential-expression",
                    "DGE_filtered_normalized.rds"))

limma_results <- read.csv(here(results_dir, "de_results_all.csv"),
                          row.names = 1)

cat("Genes:", nrow(DGE), "| Samples:", ncol(DGE), "\n")
cat("limma-voom results loaded:", nrow(limma_results), "genes\n")
```

---

# GREIN-Style Filtering

Public databases like GREIN use specific filtering criteria that are not always
fully documented. Through careful comparison we determined their actual
approach — which differs from what they describe.

```{r grein-filtering}
# GREIN's stated method: "CPM > 0"
# GREIN's actual implementation: count >= 4 in at least min_group_size samples

min_group_size <- min(table(DGE$samples$genotype))
cat("Minimum group size:", min_group_size, "\n")

keep_grein <- rowSums(DGE$counts >= 4) >= min_group_size
DGE_grein  <- DGE[keep_grein, , keep.lib.sizes = FALSE]

cat("Genes with our filterByExpr:", nrow(DGE), "\n")
cat("Genes with GREIN-style filter:", nrow(DGE_grein), "\n")
cat("(GREIN reports ~15,871 genes for this dataset)\n")
```

::: {.callout-warning}
## Reproducibility Lesson 1
GREIN's documentation states "CPM > 0" but their actual implementation
uses **count >= 4**. This is a common problem in bioinformatics — methods
descriptions are often incomplete or inaccurate. Always document your
exact thresholds and test reproducibility.
:::

# TMM Normalization

```{r normalize}
DGE_grein <- calcNormFactors(DGE_grein, method = "TMM")
summary(DGE_grein$samples$norm.factors)
```

---

# Estimate Dispersion

edgeR models count variability using dispersion parameters. We estimate
common, trended, and tagwise dispersion.

```{r dispersion, fig.width=8, fig.height=6}
design_grein <- model.matrix(~genotype, data = DGE_grein$samples)

DGE_grein <- estimateDisp(DGE_grein, design = design_grein)

plotBCV(DGE_grein,
        main = "Biological Coefficient of Variation")
```

::: {.callout-note}
## BCV Plot
**BCV** = Biological Coefficient of Variation = sqrt(dispersion)

- **Black dots**: Gene-specific dispersion
- **Blue line**: Common dispersion
- **Red curve**: Trended dispersion

Typical values: 0.1–0.2 for well-controlled experiments, 0.4–0.5 for
biological replicates.
:::

---

# Exact Test

```{r exact-test}
et_grein <- exactTest(DGE_grein, pair = c("PRMT7_WT", "PRMT7_KD"))

summary(decideTests(et_grein, adjust.method = "BH", p.value = 0.05))

topTags(et_grein, n = 10)
```

---

# GREIN's Hidden Filter

GREIN applies an additional undocumented filter to their reported DE gene
counts.

```{r hidden-filter}
all_edger_results <- topTags(et_grein, n = Inf)$table

cat("DE genes FDR < 0.05:", sum(all_edger_results$FDR < 0.05), "\n")

# GREIN's actual criteria: FDR < 0.05 AND logCPM > 1
grein_de_genes <- all_edger_results[all_edger_results$FDR < 0.05 &
                                    all_edger_results$logCPM > 1, ]

cat("DE genes FDR < 0.05 AND logCPM > 1:", nrow(grein_de_genes), "\n")
cat("(GREIN reports 8,803 for this comparison)\n")
```

::: {.callout-warning}
## Reproducibility Lesson 2
GREIN applies a **logCPM > 1** filter to their DE gene counts that is
never mentioned in their methods documentation. This removes ~1,700
lowly expressed genes from the reported list.

Without this information, you cannot reproduce their exact results. This
is why sharing actual code — not just prose descriptions — is critical.
:::

---

# Visualization

## Volcano Plot

```{r volcano, fig.width=8, fig.height=6}
grein_de_genes$category <- case_when(
  grein_de_genes$logFC >  1 ~ "Up (>2-fold)",
  grein_de_genes$logFC < -1 ~ "Down (>2-fold)",
  grein_de_genes$logFC >  0 ~ "Up",
  TRUE                       ~ "Down"
)

ggplot(grein_de_genes, aes(x = logFC, y = -log10(PValue), color = category)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_color_manual(values = c("Up (>2-fold)"   = "darkred",
                                "Down (>2-fold)" = "darkblue",
                                "Up"             = "red",
                                "Down"           = "blue")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", alpha = 0.5) +
  labs(title    = "Volcano Plot: KD vs WT (edgeR, GREIN-matched)",
       subtitle = paste0("FDR < 0.05 & logCPM > 1 (n = ",
                         nrow(grein_de_genes), " genes)"),
       x     = "Log2 Fold Change",
       y     = "-Log10 P-value",
       color = "Category") +
  theme_minimal()

ggsave(here(figures_dir, "volcano_plot_edger_grein.png"),
       width = 8, height = 6, dpi = 300)
```

## MA Plot

```{r ma-plot, fig.width=8, fig.height=6}
ggplot(grein_de_genes, aes(x = logCPM, y = logFC, color = category)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_color_manual(values = c("Up (>2-fold)"   = "darkred",
                                "Down (>2-fold)" = "darkblue",
                                "Up"             = "red",
                                "Down"           = "blue")) +
  geom_hline(yintercept = 0,        linetype = "solid",  alpha = 0.5) +
  geom_hline(yintercept = c(-1, 1), linetype = "dashed", alpha = 0.3) +
  labs(title    = "MA Plot: KD vs WT (edgeR, GREIN-matched)",
       subtitle = paste0("FDR < 0.05 & logCPM > 1 (n = ",
                         nrow(grein_de_genes), " genes)"),
       x     = "Average Expression (log2-CPM)",
       y     = "Log2 Fold Change",
       color = "Category") +
  theme_minimal()

ggsave(here(figures_dir, "ma_plot_edger_grein.png"),
       width = 8, height = 6, dpi = 300)
```

---

# Compare edgeR vs limma-voom

```{r compare-methods}
limma_matched <- limma_results %>%
  filter(adj.P.Val < 0.05, AveExpr > 1)

comparison <- data.frame(
  gene       = rownames(grein_de_genes),
  edgeR_logFC = grein_de_genes$logFC,
  edgeR_FDR   = grein_de_genes$FDR,
  limma_logFC = limma_results[rownames(grein_de_genes), "logFC"],
  limma_FDR   = limma_results[rownames(grein_de_genes), "adj.P.Val"]
)

cat("Fold change correlation between methods:",
    round(cor(comparison$edgeR_logFC, comparison$limma_logFC,
              use = "complete.obs"), 3), "\n")

ggplot(comparison, aes(x = limma_logFC, y = edgeR_logFC)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title    = "Fold Change Comparison: edgeR vs limma-voom",
       subtitle = "High correlation indicates robust findings",
       x = "limma-voom log2FC",
       y = "edgeR log2FC") +
  theme_minimal()

edger_genes <- rownames(grein_de_genes)
limma_genes <- rownames(limma_matched)

cat("\nGene overlap:\n")
cat("  Both methods:", length(intersect(edger_genes, limma_genes)), "\n")
cat("  edgeR only:  ", length(setdiff(edger_genes, limma_genes)), "\n")
cat("  limma only:  ", length(setdiff(limma_genes, edger_genes)), "\n")
cat("  Agreement:   ",
    round(100 * length(intersect(edger_genes, limma_genes)) /
          length(union(edger_genes, limma_genes)), 1), "%\n")
```

::: {.callout-important}
## Which Method Should You Use?

**edgeR exact test:**
- Simple two-group comparisons
- Count-based (no transformation)
- Matches published GREIN analyses

**limma-voom:**
- Complex designs (multiple factors, interactions, time series)
- Contrast matrices
- Variable sample quality (`voomWithQualityWeights`)

Both methods typically agree >95% for simple comparisons.
:::

---

# Export

```{r export}
edger_export <- grein_de_genes
edger_export$gene <- rownames(edger_export)

write.csv(edger_export,
          here(results_dir, "edger_grein_matched_results.csv"),
          row.names = FALSE)

writeLines(capture.output(sessionInfo()),
           here(results_dir, "sessionInfo_edger.txt"))

cat("✓ edger_grein_matched_results.csv\n")
cat("✓ sessionInfo_edger.txt\n")
```

---

# Session Information

```{r sessioninfo}
sessionInfo()
```
