---
title: "Optional: Advanced Designs — Two-Factor Analysis"
author: "UFHCI BCB-SR Bioinformatics Unit"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

Our PRMT7 experiment has only one factor (genotype). To demonstrate how to
handle more complex experimental designs, we use a published *Drosophila
melanogaster* temperature adaptation dataset with a proper 2×2 factorial
structure: two populations (Maine and Panama) × two temperatures (low and
high) × three replicates.

This script is standalone — it does not require output from the other workshop
scripts.

**You will learn:**

- How to build a two-factor additive design matrix and interpret main effects
- How to build an interaction model using group-means parameterization
- How to define and interpret contrast matrices
- How to classify parallel vs. divergent response patterns

These skills apply to any multi-factor experiment: genotype × treatment,
tissue × disease, time × condition, and so on.

---

## Setup

```{r libraries}
library(limma)
library(edgeR)
library(tidyverse)
library(ggrepel)
library(pheatmap)
library(RColorBrewer)
library(here)

set.seed(42)

results_dir <- here("demo-analysis", "output", "differential-expression", "results")
figures_dir <- here("demo-analysis", "output", "differential-expression", "figures")

dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(figures_dir, showWarnings = FALSE, recursive = TRUE)
```

---

## Load Data

```{r load-data}
drosophila_counts <- read.table(
  here("demo-analysis", "data", "two-factor-design", "salmon.merged.gene_counts.tsv"),
  header    = TRUE,
  row.names = "gene_name"
)
drosophila_counts$gene_id <- NULL

drosophila_info <- read.table(
  here("demo-analysis", "data", "two-factor-design", "dme_elev_samples.tsv"),
  header           = TRUE,
  stringsAsFactors = FALSE
)

# Combined factor for filtering
drosophila_info$pop_temp <- paste(drosophila_info$population,
                                   drosophila_info$temp, sep = "_")

cat("Experimental Design:\n")
print(table(drosophila_info$population, drosophila_info$temp))
drosophila_info
```

---

## Create DGEList, Filter, Normalize

```{r dge}
DGE_dros               <- DGEList(counts = drosophila_counts, samples = drosophila_info)
DGE_dros$samples$group <- factor(drosophila_info$pop_temp)
keep_dros              <- filterByExpr(DGE_dros, group = DGE_dros$samples$group)
DGE_dros               <- DGE_dros[keep_dros, , keep.lib.sizes = FALSE]
DGE_dros               <- calcNormFactors(DGE_dros, method = "TMM")

cat("Genes after filtering:", nrow(DGE_dros), "\n")
```

---

## QC — MDS Plot

Maine and Panama samples should cluster apart. Low and high temperature samples
should also separate. If the temperature separation differs between populations,
that is evidence of an interaction — which we test in Part 2 below.

```{r mds, fig.width=8, fig.height=6}
mds_dros <- plotMDS(DGE_dros, top = 500, plot = FALSE, gene.selection = "common")

mds_dros_df <- data.frame(
  sample      = rownames(DGE_dros$samples),
  dim1        = mds_dros$x,
  dim2        = mds_dros$y,
  population  = DGE_dros$samples$population,
  temperature = DGE_dros$samples$temp
)

ggplot(mds_dros_df, aes(x = dim1, y = dim2,
                         color = population, shape = temperature)) +
  geom_point(size = 4) +
  geom_text_repel(aes(label = sample), size = 3) +
  labs(title = "MDS Plot: Drosophila Temperature Experiment",
       x = paste0("Dim 1 (", round(mds_dros$var.explained[1] * 100, 1), "%)"),
       y = paste0("Dim 2 (", round(mds_dros$var.explained[2] * 100, 1), "%)"),
       color = "Population", shape = "Temperature") +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")
```

---

## Part 1: Additive Two-Factor Model

An additive model (`~population + temp`) tests the main effect of each factor
while assuming they act independently. The coefficients represent:

- **(Intercept)**: Maine + High temperature (both reference levels)
- **populationpanama**: Panama vs Maine, averaged across temperatures
- **templow**: Low vs High temperature, averaged across populations

```{r design-additive}
DGE_dros$samples$population <- factor(DGE_dros$samples$population,
                                       levels = c("maine", "panama"))
DGE_dros$samples$temp       <- factor(DGE_dros$samples$temp,
                                       levels = c("high", "low"))

design_2factor <- model.matrix(~population + temp, data = DGE_dros$samples)
design_2factor
```

```{r fit-additive, fig.width=8, fig.height=5}
vwts_2factor <- voomWithQualityWeights(DGE_dros,
                                        design           = design_2factor,
                                        normalize.method = "none",
                                        plot             = TRUE)

fit_2factor <- lmFit(vwts_2factor, design_2factor)
fit_2factor <- eBayes(fit_2factor, robust = TRUE)

summary(decideTests(fit_2factor, adjust.method = "fdr", p.value = 0.05))
```

```{r results-additive}
temp_results <- topTable(fit_2factor, coef = "templow",
                          adjust.method = "BH", number = Inf)
pop_results  <- topTable(fit_2factor, coef = "populationpanama",
                          adjust.method = "BH", number = Inf)

cat("Temperature main effect (FDR < 0.05):",
    sum(temp_results$adj.P.Val < 0.05), "\n")
cat("Population main effect (FDR < 0.05): ",
    sum(pop_results$adj.P.Val < 0.05), "\n\n")

cat("Top 10 temperature-responsive genes:\n")
print(head(temp_results, 10))
```

```{r volcano-additive, fig.width=8, fig.height=6}
temp_results$category <- case_when(
  temp_results$adj.P.Val < 0.05 & temp_results$logFC >  1 ~ "Up in Low",
  temp_results$adj.P.Val < 0.05 & temp_results$logFC < -1 ~ "Up in High",
  temp_results$adj.P.Val < 0.05                            ~ "Significant",
  TRUE                                                      ~ "Not Significant"
)

ggplot(temp_results, aes(x = logFC, y = -log10(P.Value), color = category)) +
  geom_point(alpha = 0.5, size = 1) +
  scale_color_manual(values = c("Up in Low"       = "blue",
                                "Up in High"      = "red",
                                "Significant"     = "orange",
                                "Not Significant" = "grey80")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", alpha = 0.5) +
  labs(title    = "Temperature Main Effect (Averaged Across Populations)",
       subtitle = "Low vs High Temperature",
       x = "Log2 Fold Change (Low vs High)", y = "-Log10 P-value") +
  theme_minimal()

ggsave(here(figures_dir, "volcano_drosophila_temperature.png"),
       width = 8, height = 6, dpi = 300)
```

Note: this volcano plot shows the temperature effect **averaged** across both
populations. If the effect differs substantially between Maine and Panama, this
average may be misleading — the interaction model below tests this directly.

---

## Part 2: Interaction Model with Contrasts

The interaction model gives each group combination its own coefficient, then
uses contrast matrices to test specific biological hypotheses. The interaction
contrast tests whether the temperature effect differs between populations.

### Group-Means Design

```{r design-interaction}
DGE_dros$samples$group_combined <- factor(
  paste(DGE_dros$samples$population, DGE_dros$samples$temp, sep = "_"),
  levels = c("maine_high", "maine_low", "panama_high", "panama_low")
)

design_interaction           <- model.matrix(~0 + group_combined,
                                              data = DGE_dros$samples)
colnames(design_interaction) <- levels(DGE_dros$samples$group_combined)
design_interaction
```

### Define Contrasts

Each contrast is a linear combination of group means that tests a specific
biological question.

- **Maine_TempEffect**: Temperature response in Maine only
- **Panama_TempEffect**: Temperature response in Panama only
- **Interaction**: Whether temperature response differs between populations.
  Significant genes show divergent responses; non-significant genes show
  parallel responses

```{r contrasts}
cont_matrix <- makeContrasts(
  Maine_TempEffect  = maine_low - maine_high,
  Panama_TempEffect = panama_low - panama_high,
  Pop_at_High       = panama_high - maine_high,
  Pop_at_Low        = panama_low - maine_low,
  Interaction       = (maine_low - maine_high) - (panama_low - panama_high),
  levels            = design_interaction
)
cont_matrix
```

### Fit and Apply Contrasts

```{r fit-interaction, fig.width=8, fig.height=5}
vwts_interaction <- voomWithQualityWeights(DGE_dros,
                                            design           = design_interaction,
                                            normalize.method = "none",
                                            plot             = TRUE)

fit_interaction <- lmFit(vwts_interaction, design_interaction)
fit_contrasts   <- contrasts.fit(fit_interaction, cont_matrix)
fit_contrasts   <- eBayes(fit_contrasts, robust = TRUE)

contrast_summary <- decideTests(fit_contrasts, adjust.method = "fdr", p.value = 0.05)
summary(contrast_summary)
```

```{r venn, fig.width=8, fig.height=6}
# Venn diagram showing overlap between the two population temperature effects
# and the interaction
vennDiagram(contrast_summary[, c("Maine_TempEffect",
                                   "Panama_TempEffect",
                                   "Interaction")],
            circle.col = c("blue", "red", "green"),
            main       = "Overlap of DE Genes Across Contrasts")
```

### Extract Results

```{r extract-contrasts}
maine_temp  <- topTable(fit_contrasts, coef = "Maine_TempEffect",
                         adjust.method = "BH", number = Inf, sort.by = "P")
panama_temp <- topTable(fit_contrasts, coef = "Panama_TempEffect",
                         adjust.method = "BH", number = Inf, sort.by = "P")
interaction <- topTable(fit_contrasts, coef = "Interaction",
                         adjust.method = "BH", number = Inf, sort.by = "P")

cat("Maine temperature effect (FDR < 0.05): ",
    sum(maine_temp$adj.P.Val < 0.05), "\n")
cat("Panama temperature effect (FDR < 0.05):",
    sum(panama_temp$adj.P.Val < 0.05), "\n")
cat("Interaction (FDR < 0.05):              ",
    sum(interaction$adj.P.Val < 0.05), "\n")
```

### Classify Response Patterns

Genes are classified based on which contrasts they are significant in.
Parallel genes show conserved thermal stress responses across both populations.
Divergent genes show evidence of evolutionary divergence in temperature
response.

```{r classify-patterns}
all_genes <- rownames(maine_temp)

contrast_results <- data.frame(
  gene              = all_genes,
  Maine_logFC       = maine_temp[all_genes, "logFC"],
  Maine_FDR         = maine_temp[all_genes, "adj.P.Val"],
  Panama_logFC      = panama_temp[all_genes, "logFC"],
  Panama_FDR        = panama_temp[all_genes, "adj.P.Val"],
  Interaction_logFC = interaction[all_genes, "logFC"],
  Interaction_FDR   = interaction[all_genes, "adj.P.Val"]
) %>%
  mutate(
    Maine_sig       = Maine_FDR < 0.05,
    Panama_sig      = Panama_FDR < 0.05,
    Interaction_sig = Interaction_FDR < 0.05,
    pattern = case_when(
      Maine_sig & Panama_sig & !Interaction_sig &
        sign(Maine_logFC) == sign(Panama_logFC) ~ "Parallel",
      Interaction_sig                           ~ "Divergent",
      Maine_sig  & !Panama_sig                  ~ "Maine-specific",
      Panama_sig & !Maine_sig                   ~ "Panama-specific",
      TRUE                                      ~ "Not significant"
    )
  )

print(table(contrast_results$pattern))
```

### Heatmap of Top Divergent Genes

```{r heatmap-divergent, fig.width=10, fig.height=8}
top_divergent <- contrast_results %>%
  filter(pattern == "Divergent") %>%
  arrange(Interaction_FDR) %>%
  head(50) %>%
  pull(gene)

divergent_expr <- vwts_interaction$E[top_divergent, ]

col_annotation <- data.frame(
  Population  = DGE_dros$samples$population,
  Temperature = DGE_dros$samples$temp,
  row.names   = colnames(divergent_expr)
)

pheatmap(divergent_expr,
         scale          = "row",
         annotation_col = col_annotation,
         cluster_cols   = FALSE,
         show_rownames  = TRUE,
         main           = "Top 50 Genes with Divergent Temperature Responses",
         color          = colorRampPalette(c("blue", "white", "red"))(100),
         fontsize_row   = 6)
```

---

## Export Results

```{r export}
write.csv(contrast_results,
          here(results_dir, "drosophila_contrast_results.csv"),   row.names = FALSE)
write.csv(maine_temp,
          here(results_dir, "drosophila_maine_temperature.csv"))
write.csv(panama_temp,
          here(results_dir, "drosophila_panama_temperature.csv"))
write.csv(interaction,
          here(results_dir, "drosophila_interaction.csv"))
write.csv(contrast_results %>% filter(pattern == "Parallel"),
          here(results_dir, "drosophila_parallel_response_genes.csv"),  row.names = FALSE)
write.csv(contrast_results %>% filter(pattern == "Divergent"),
          here(results_dir, "drosophila_divergent_response_genes.csv"), row.names = FALSE)

cat("✓ All results exported\n")
```

---

## Session Information

```{r sessioninfo}
sessionInfo()
```
