---
title: "Optional: edgeR Exact Test and GREIN Reproducibility"
author: "UFHCI BCB-SR Bioinformatics Unit"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

This script explores an alternative approach to the same analysis using the
edgeR exact test and compares results to the limma-voom analysis in Part 2.
It also uses this comparison as a case study in reproducibility — specifically,
how undocumented implementation details in a public database (GREIN) can prevent
exact reproduction of published results.

**Requires:** `DGE_filtered_normalized.rds` and `de_results_all.csv` from
Parts 1 and 2.

---

## Setup

```{r libraries}
library(limma)
library(edgeR)
library(tidyverse)
library(here)

set.seed(42)

results_dir <- here("demo-analysis", "output", "differential-expression", "results")
figures_dir <- here("demo-analysis", "output", "differential-expression", "figures")

dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(figures_dir, showWarnings = FALSE, recursive = TRUE)
```

---

## Load Data

```{r load-data}
DGE <- readRDS(here("demo-analysis", "output", "differential-expression",
                    "DGE_filtered_normalized.rds"))

limma_results <- read.csv(here(results_dir, "de_results_all.csv"),
                          row.names = 1)

cat("Genes:", nrow(DGE), "| Samples:", ncol(DGE), "\n")
cat("limma-voom results loaded:", nrow(limma_results), "genes\n")
```

---

## GREIN-Style Filtering

Public databases like GREIN document their filtering as "CPM > 0" but their
actual implementation uses a different threshold. Through careful comparison
we determined the actual criteria: count >= 4 in at least as many samples as
the smallest group.

```{r grein-filtering}
min_group_size <- min(table(DGE$samples$genotype))
cat("Minimum group size:", min_group_size, "\n")

keep_grein <- rowSums(DGE$counts >= 4) >= min_group_size
DGE_grein  <- DGE[keep_grein, , keep.lib.sizes = FALSE]

cat("Genes with filterByExpr:      ", nrow(DGE), "\n")
cat("Genes with GREIN-style filter:", nrow(DGE_grein), "\n")
cat("(GREIN reports ~15,871 genes for this dataset)\n")
```

This is Reproducibility Lesson 1: methods descriptions are often incomplete or
inaccurate. Always document your exact thresholds and test reproducibility.

---

## TMM Normalization

```{r normalize}
DGE_grein <- calcNormFactors(DGE_grein, method = "TMM")
summary(DGE_grein$samples$norm.factors)
```

---

## Estimate Dispersion

edgeR models count variability using dispersion parameters. The BCV
(Biological Coefficient of Variation) plot shows gene-specific dispersion
(black dots), common dispersion (blue line), and trended dispersion (red curve).
Typical BCV values are 0.1–0.2 for well-controlled experiments.

```{r dispersion, fig.width=8, fig.height=6}
design_grein <- model.matrix(~genotype, data = DGE_grein$samples)
DGE_grein    <- estimateDisp(DGE_grein, design = design_grein)

plotBCV(DGE_grein, main = "Biological Coefficient of Variation")
```

---

## Exact Test

The edgeR exact test compares two groups using the negative binomial
distribution directly on counts, without transforming to log-scale as
limma-voom does.

```{r exact-test}
et_grein <- exactTest(DGE_grein, pair = c("PRMT7_WT", "PRMT7_KD"))

summary(decideTests(et_grein, adjust.method = "BH", p.value = 0.05))
topTags(et_grein, n = 10)
```

---

## GREIN's Hidden Filter

GREIN applies an additional undocumented filter to their reported DE gene
counts: FDR < 0.05 AND logCPM > 1. This is never mentioned in their methods
documentation and removes ~1,700 lowly expressed genes from the reported list.
This is Reproducibility Lesson 2: without the actual code, you cannot reproduce
published results.

```{r hidden-filter}
all_edger_results <- topTags(et_grein, n = Inf)$table

cat("DE genes FDR < 0.05:              ", sum(all_edger_results$FDR < 0.05), "\n")

# Apply GREIN's undocumented filter
grein_de_genes <- all_edger_results[all_edger_results$FDR < 0.05 &
                                    all_edger_results$logCPM > 1, ]

cat("DE genes FDR < 0.05 & logCPM > 1:", nrow(grein_de_genes), "\n")
cat("(GREIN reports 8,803 for this comparison)\n")
```

---

## Visualization

### Volcano Plot

```{r volcano, fig.width=8, fig.height=6}
grein_de_genes$category <- case_when(
  grein_de_genes$logFC >  1 ~ "Up (>2-fold)",
  grein_de_genes$logFC < -1 ~ "Down (>2-fold)",
  grein_de_genes$logFC >  0 ~ "Up",
  TRUE                       ~ "Down"
)

ggplot(grein_de_genes, aes(x = logFC, y = -log10(PValue), color = category)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_color_manual(values = c("Up (>2-fold)"   = "darkred",
                                "Down (>2-fold)" = "darkblue",
                                "Up"             = "red",
                                "Down"           = "blue")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", alpha = 0.5) +
  labs(title    = "Volcano Plot: KD vs WT (edgeR, GREIN-matched)",
       subtitle = paste0("FDR < 0.05 & logCPM > 1 (n = ",
                         nrow(grein_de_genes), " genes)"),
       x = "Log2 Fold Change", y = "-Log10 P-value", color = "Category") +
  theme_minimal()

ggsave(here(figures_dir, "volcano_plot_edger_grein.png"),
       width = 8, height = 6, dpi = 300)
```

### MA Plot

```{r ma-plot, fig.width=8, fig.height=6}
ggplot(grein_de_genes, aes(x = logCPM, y = logFC, color = category)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_color_manual(values = c("Up (>2-fold)"   = "darkred",
                                "Down (>2-fold)" = "darkblue",
                                "Up"             = "red",
                                "Down"           = "blue")) +
  geom_hline(yintercept = 0,        linetype = "solid",  alpha = 0.5) +
  geom_hline(yintercept = c(-1, 1), linetype = "dashed", alpha = 0.3) +
  labs(title    = "MA Plot: KD vs WT (edgeR, GREIN-matched)",
       subtitle = paste0("FDR < 0.05 & logCPM > 1 (n = ",
                         nrow(grein_de_genes), " genes)"),
       x = "Average Expression (log2-CPM)",
       y = "Log2 Fold Change", color = "Category") +
  theme_minimal()

ggsave(here(figures_dir, "ma_plot_edger_grein.png"),
       width = 8, height = 6, dpi = 300)
```

---

## Compare edgeR vs limma-voom

Both methods typically agree >95% for simple two-group comparisons — high
agreement increases confidence in the findings. Small differences arise from
different statistical frameworks: edgeR models counts directly using the
negative binomial distribution, while limma-voom transforms counts to
log-scale and applies precision weights.

```{r compare}
limma_matched <- limma_results %>% filter(adj.P.Val < 0.05, AveExpr > 1)

comparison <- data.frame(
  gene        = rownames(grein_de_genes),
  edgeR_logFC = grein_de_genes$logFC,
  limma_logFC = limma_results[rownames(grein_de_genes), "logFC"]
)

cat("logFC correlation:", round(cor(comparison$edgeR_logFC,
                                    comparison$limma_logFC,
                                    use = "complete.obs"), 3), "\n")

ggplot(comparison, aes(x = limma_logFC, y = edgeR_logFC)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Fold Change Comparison: edgeR vs limma-voom",
       x = "limma-voom log2FC", y = "edgeR log2FC") +
  theme_minimal()

edger_genes <- rownames(grein_de_genes)
limma_genes <- rownames(limma_matched)

cat("Both methods:  ", length(intersect(edger_genes, limma_genes)), "\n")
cat("edgeR only:    ", length(setdiff(edger_genes, limma_genes)), "\n")
cat("limma only:    ", length(setdiff(limma_genes, edger_genes)), "\n")
cat("Agreement:     ",
    round(100 * length(intersect(edger_genes, limma_genes)) /
          length(union(edger_genes, limma_genes)), 1), "%\n")
```

---

## Export

```{r export}
edger_export      <- grein_de_genes
edger_export$gene <- rownames(edger_export)

write.csv(edger_export,
          here(results_dir, "edger_grein_matched_results.csv"),
          row.names = FALSE)

writeLines(capture.output(sessionInfo()),
           here(results_dir, "sessionInfo_edger.txt"))

cat("✓ edger_grein_matched_results.csv\n")
cat("✓ sessionInfo_edger.txt\n")
```

---

## Session Information

```{r sessioninfo}
sessionInfo()
```
