---
title: "Optional: Advanced Designs — Two-Factor Analysis"
subtitle: "UF Health Cancer Institute Bioinformatics Workshop"
author: "UFHCI BCB-SR Bioinformatics Unit"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: false
    code-tools: true
    code-overflow: wrap
    embed-resources: true
    theme: cosmo
    df-print: paged
execute:
  warning: false
  message: false
---

# Introduction

This notebook demonstrates limma-voom with a two-factor experimental design.
Our PRMT7 experiment has only one factor (genotype). To show how to handle
more complex designs, we use a published Drosophila temperature adaptation
dataset with a proper 2×2 factorial structure.

**Dataset:** *Drosophila melanogaster* temperature adaptation
(Zhao et al. 2015, *PLoS Genetics*)

**Design:**

- **Population**: Maine (high latitude) vs Panama (low latitude)
- **Temperature**: Low vs High
- **Structure**: 2 × 2 × 3 replicates = 12 samples

**This notebook is standalone** — it does not require output from the
other workshop scripts.

**You will learn:**

- How to build a two-factor additive design matrix
- How to build an interaction model using group-means parameterization
- How to define and interpret contrast matrices
- How to classify parallel vs divergent response patterns

---

# Setup

```{r setup}
library(limma)
library(edgeR)
library(tidyverse)
library(ggrepel)
library(pheatmap)
library(RColorBrewer)
library(here)

set.seed(42)

results_dir <- here("demo-analysis", "output", "02-differential-expression", "results")
figures_dir <- here("demo-analysis", "output", "02-differential-expression", "figures")

dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(figures_dir, showWarnings = FALSE, recursive = TRUE)
```

---

# Load Data

```{r load-data}
drosophila_counts <- read.table(
  here("demo-analysis", "data", "two-factor-design", "salmon.merged.gene_counts.tsv"),
  header     = TRUE,
  row.names  = "gene_name"
)

drosophila_counts$gene_id <- NULL

drosophila_info <- read.table(
  here("demo-analysis", "data", "two-factor-design", "dme_elev_samples.tsv"),
  header          = TRUE,
  stringsAsFactors = FALSE
)

drosophila_info$pop_temp <- paste(drosophila_info$population,
                                   drosophila_info$temp,
                                   sep = "_")

cat("Experimental Design:\n")
print(table(drosophila_info$population, drosophila_info$temp))

drosophila_info
```

---

# Create DGEList, Filter, Normalize

```{r dge}
DGE_dros <- DGEList(counts  = drosophila_counts,
                    samples = drosophila_info)

DGE_dros$samples$group <- factor(drosophila_info$pop_temp)

keep_dros <- filterByExpr(DGE_dros, group = DGE_dros$samples$group)
DGE_dros  <- DGE_dros[keep_dros, , keep.lib.sizes = FALSE]
DGE_dros  <- calcNormFactors(DGE_dros, method = "TMM")

cat("Genes after filtering:", nrow(DGE_dros), "\n")
```

---

# QC — MDS Plot

```{r mds, fig.width=8, fig.height=6}
mds_dros <- plotMDS(DGE_dros, top = 500, plot = FALSE,
                    gene.selection = "common")

mds_dros_df <- data.frame(
  sample     = rownames(DGE_dros$samples),
  dim1       = mds_dros$x,
  dim2       = mds_dros$y,
  population = DGE_dros$samples$population,
  temperature = DGE_dros$samples$temp
)

ggplot(mds_dros_df, aes(x = dim1, y = dim2,
                         color = population,
                         shape = temperature)) +
  geom_point(size = 4) +
  geom_text_repel(aes(label = sample), size = 3) +
  labs(title = "MDS Plot: Drosophila Temperature Experiment",
       x = paste0("Dim 1 (", round(mds_dros$var.explained[1] * 100, 1), "%)"),
       y = paste0("Dim 2 (", round(mds_dros$var.explained[2] * 100, 1), "%)"),
       color = "Population", shape = "Temperature") +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")
```

::: {.callout-tip}
## What to Look For
- Do Maine and Panama samples cluster apart?
- Do low and high temperature samples separate?
- Does the temperature separation differ between populations?
  That last point is an interaction — what we test in Part 2 of this notebook.
:::

---

# Part 1: Additive Two-Factor Model

An additive model (`~population + temp`) tests the main effect of each
factor while assuming they act independently.

```{r design-additive}
DGE_dros$samples$population <- factor(DGE_dros$samples$population,
                                       levels = c("maine", "panama"))
DGE_dros$samples$temp       <- factor(DGE_dros$samples$temp,
                                       levels = c("high", "low"))

design_2factor <- model.matrix(~population + temp,
                                data = DGE_dros$samples)

design_2factor
colnames(design_2factor)
```

::: {.callout-note}
## Additive Model Coefficients

- **(Intercept)**: Maine + High temperature (both reference levels)
- **populationpanama**: Panama vs Maine, averaged across temperatures
- **templow**: Low vs High temperature, averaged across populations

The model assumes the temperature effect is the same magnitude in both
populations. If that assumption is violated, you need an interaction model
(Part 2 below).
:::

```{r fit-additive, fig.width=8, fig.height=5}
vwts_2factor <- voomWithQualityWeights(DGE_dros,
                                        design           = design_2factor,
                                        normalize.method = "none",
                                        plot             = TRUE)

fit_2factor <- lmFit(vwts_2factor, design_2factor)
fit_2factor <- eBayes(fit_2factor, robust = TRUE)

summary(decideTests(fit_2factor, adjust.method = "fdr", p.value = 0.05))
```

```{r results-additive}
temp_results <- topTable(fit_2factor,
                          coef          = "templow",
                          adjust.method = "BH",
                          number        = Inf)

pop_results  <- topTable(fit_2factor,
                          coef          = "populationpanama",
                          adjust.method = "BH",
                          number        = Inf)

cat("Temperature main effect (FDR < 0.05):",
    sum(temp_results$adj.P.Val < 0.05), "genes\n")
cat("Population main effect (FDR < 0.05):",
    sum(pop_results$adj.P.Val < 0.05), "genes\n\n")

cat("Top 10 temperature-responsive genes:\n")
print(head(temp_results, 10))
```

```{r volcano-additive, fig.width=8, fig.height=6}
temp_results$category <- case_when(
  temp_results$adj.P.Val < 0.05 & temp_results$logFC >  1 ~ "Up in Low",
  temp_results$adj.P.Val < 0.05 & temp_results$logFC < -1 ~ "Up in High",
  temp_results$adj.P.Val < 0.05                            ~ "Significant",
  TRUE                                                      ~ "Not Significant"
)

ggplot(temp_results, aes(x = logFC, y = -log10(P.Value), color = category)) +
  geom_point(alpha = 0.5, size = 1) +
  scale_color_manual(values = c("Up in Low"       = "blue",
                                "Up in High"      = "red",
                                "Significant"     = "orange",
                                "Not Significant" = "grey80")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", alpha = 0.5) +
  labs(title    = "Temperature Main Effect (Averaged Across Populations)",
       subtitle = "Low vs High Temperature",
       x = "Log2 Fold Change (Low vs High)", y = "-Log10 P-value",
       color = "Category") +
  theme_minimal()

ggsave(here(figures_dir, "volcano_drosophila_temperature.png"),
       width = 8, height = 6, dpi = 300)
```

::: {.callout-warning}
## Limitation of the Additive Model
This volcano plot shows the temperature effect **averaged** across both
populations. If the temperature effect is very different in Maine vs
Panama, this average may be misleading. The interaction model in Part 2
tests this directly.
:::

---

# Part 2: Interaction Model with Contrasts

The interaction model tests whether the temperature effect differs between
populations by giving each group combination its own coefficient.

## Group-Means Design

```{r design-interaction}
DGE_dros$samples$group_combined <- factor(
  paste(DGE_dros$samples$population, DGE_dros$samples$temp, sep = "_"),
  levels = c("maine_high", "maine_low", "panama_high", "panama_low")
)

design_interaction <- model.matrix(~0 + group_combined,
                                    data = DGE_dros$samples)

colnames(design_interaction) <- levels(DGE_dros$samples$group_combined)

design_interaction
```

::: {.callout-note}
## Group-Means Parameterization
Using `~0 + group_combined` gives each group its own coefficient rather
than computing differences from a reference. This makes it easy to define
any contrast we want.
:::

## Define Contrasts

```{r contrasts}
cont_matrix <- makeContrasts(
  Maine_TempEffect   = maine_low - maine_high,
  Panama_TempEffect  = panama_low - panama_high,
  Pop_at_High        = panama_high - maine_high,
  Pop_at_Low         = panama_low - maine_low,
  Interaction        = (maine_low - maine_high) - (panama_low - panama_high),
  levels             = design_interaction
)

cont_matrix
```

::: {.callout-important}
## Understanding the Interaction Contrast

`Interaction = (maine_low - maine_high) - (panama_low - panama_high)`

This tests whether the temperature effect differs between populations:

- **Significant positive**: Stronger low-temp response in Maine
- **Significant negative**: Stronger low-temp response in Panama
- **Not significant**: Parallel responses — same temperature effect in both populations
:::

## Fit Model and Apply Contrasts

```{r fit-interaction, fig.width=8, fig.height=5}
vwts_interaction <- voomWithQualityWeights(DGE_dros,
                                            design           = design_interaction,
                                            normalize.method = "none",
                                            plot             = TRUE)

fit_interaction  <- lmFit(vwts_interaction, design_interaction)
fit_contrasts    <- contrasts.fit(fit_interaction, cont_matrix)
fit_contrasts    <- eBayes(fit_contrasts, robust = TRUE)

contrast_summary <- decideTests(fit_contrasts,
                                 adjust.method = "fdr",
                                 p.value       = 0.05)
summary(contrast_summary)
```

```{r venn, fig.width=8, fig.height=6}
vennDiagram(contrast_summary[, c("Maine_TempEffect",
                                   "Panama_TempEffect",
                                   "Interaction")],
            circle.col = c("blue", "red", "green"),
            main       = "Overlap of DE Genes Across Contrasts")
```

::: {.callout-tip}
## Reading the Venn Diagram
- **Maine ∩ Panama (no interaction)**: Parallel temperature responses
- **Interaction only**: Divergent responses between populations
- **Maine only / Panama only**: Population-specific temperature responses
:::

## Extract Results

```{r extract-contrasts}
maine_temp  <- topTable(fit_contrasts, coef = "Maine_TempEffect",
                         adjust.method = "BH", number = Inf, sort.by = "P")
panama_temp <- topTable(fit_contrasts, coef = "Panama_TempEffect",
                         adjust.method = "BH", number = Inf, sort.by = "P")
interaction <- topTable(fit_contrasts, coef = "Interaction",
                         adjust.method = "BH", number = Inf, sort.by = "P")

cat("Maine temperature effect (FDR < 0.05): ",
    sum(maine_temp$adj.P.Val < 0.05), "genes\n")
cat("Panama temperature effect (FDR < 0.05):",
    sum(panama_temp$adj.P.Val < 0.05), "genes\n")
cat("Interaction (FDR < 0.05):              ",
    sum(interaction$adj.P.Val < 0.05), "genes\n")
```

## Classify Response Patterns

```{r classify-patterns}
all_genes <- rownames(maine_temp)

contrast_results <- data.frame(
  gene             = all_genes,
  Maine_logFC      = maine_temp[all_genes, "logFC"],
  Maine_FDR        = maine_temp[all_genes, "adj.P.Val"],
  Panama_logFC     = panama_temp[all_genes, "logFC"],
  Panama_FDR       = panama_temp[all_genes, "adj.P.Val"],
  Interaction_logFC = interaction[all_genes, "logFC"],
  Interaction_FDR  = interaction[all_genes, "adj.P.Val"]
) %>%
  mutate(
    Maine_sig       = Maine_FDR < 0.05,
    Panama_sig      = Panama_FDR < 0.05,
    Interaction_sig = Interaction_FDR < 0.05,
    pattern = case_when(
      Maine_sig & Panama_sig & !Interaction_sig &
        sign(Maine_logFC) == sign(Panama_logFC) ~ "Parallel",
      Interaction_sig                           ~ "Divergent",
      Maine_sig  & !Panama_sig                  ~ "Maine-specific",
      Panama_sig & !Maine_sig                   ~ "Panama-specific",
      TRUE                                      ~ "Not significant"
    )
  )

cat("\nResponse pattern classification:\n")
print(table(contrast_results$pattern))
```

::: {.callout-note}
## Biological Interpretation

**Parallel**: Conserved thermal stress responses — same direction and
magnitude in both populations

**Divergent**: Evidence of evolutionary divergence — populations have
adapted different responses to temperature

**Population-specific**: Response only detectable in one population —
may reflect local adaptation or statistical power differences
:::

## Heatmap of Divergent Genes

```{r heatmap-divergent, fig.width=10, fig.height=8}
top_divergent <- contrast_results %>%
  filter(pattern == "Divergent") %>%
  arrange(Interaction_FDR) %>%
  head(50) %>%
  pull(gene)

divergent_expr <- vwts_interaction$E[top_divergent, ]

col_annotation <- data.frame(
  Population  = DGE_dros$samples$population,
  Temperature = DGE_dros$samples$temp,
  row.names   = colnames(divergent_expr)
)

pheatmap(divergent_expr,
         scale          = "row",
         annotation_col = col_annotation,
         cluster_cols   = FALSE,
         show_rownames  = TRUE,
         show_colnames  = TRUE,
         main           = "Top 50 Genes with Divergent Temperature Responses",
         color          = colorRampPalette(c("blue", "white", "red"))(100),
         fontsize_row   = 6)
```

---

# Export Results

```{r export}
write.csv(contrast_results,
          here(results_dir, "drosophila_contrast_results.csv"),
          row.names = FALSE)
write.csv(maine_temp,
          here(results_dir, "drosophila_maine_temperature.csv"))
write.csv(panama_temp,
          here(results_dir, "drosophila_panama_temperature.csv"))
write.csv(interaction,
          here(results_dir, "drosophila_interaction.csv"))

parallel_genes  <- contrast_results %>% filter(pattern == "Parallel")
divergent_genes <- contrast_results %>% filter(pattern == "Divergent")

write.csv(parallel_genes,
          here(results_dir, "drosophila_parallel_response_genes.csv"),
          row.names = FALSE)
write.csv(divergent_genes,
          here(results_dir, "drosophila_divergent_response_genes.csv"),
          row.names = FALSE)

cat("✓ drosophila_contrast_results.csv\n")
cat("✓ drosophila_maine_temperature.csv\n")
cat("✓ drosophila_panama_temperature.csv\n")
cat("✓ drosophila_interaction.csv\n")
cat("✓ drosophila_parallel_response_genes.csv\n")
cat("✓ drosophila_divergent_response_genes.csv\n")
```

---

# Key Takeaways

::: {.callout-important}
## When to Use Each Approach

**Additive model** (`~factor1 + factor2`):
- Tests main effects
- Simpler, more power if no interaction
- Use when you expect similar responses across groups

**Interaction model** (`~0 + group` + contrasts):
- Tests specific hypotheses
- Detects divergent responses
- Maximum flexibility
- Use when you expect group-specific responses

**General principle**: Start with the additive model. If biology suggests
the effect of one factor might depend on another, test for interaction.

This framework applies to any multi-factor experiment: genotype ×
treatment, tissue × disease, time × condition, and so on.
:::

---

# Session Information

```{r sessioninfo}
sessionInfo()
```
