---
title: "RNA-seq Differential Expression Analysis with limma-voom"
subtitle: "UF Health Cancer Institute Bioinformatics Workshop - Session 2"
author: "UFHCI BCB-SR Bioinformatics Unit"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: false
    code-tools: true
    code-overflow: wrap
    embed-resources: true
    theme: cosmo
    df-print: paged
execute:
  warning: false
  message: false
---

# Introduction

In this session, we'll work through a complete differential expression (DE) analysis using RNA-seq data. We'll be using the output from the nf-core/rnaseq pipeline (STAR alignment + RSEM quantification) that was configured in Session 1.

**Learning Objectives:**

- Load and prepare RNA-seq count data from RSEM
- Perform quality control and exploratory data analysis
- Understand filtering, normalization, and the limma-voom workflow
- Conduct differential expression analysis for simple and complex designs
- Create publication-quality visualizations
- Interpret and export results

## The Analysis Workflow

```{mermaid}
%%| fig-width: 8
flowchart TD
    A[RSEM Gene Counts] --> B[Create DGEList Object]
    B --> C[Filter Low Expression]
    C --> D[TMM Normalization]
    D --> E[Quality Control & Visualization]
    E --> F[Design Matrix]
    F --> G[voom Transformation + Quality Weights]
    G --> H[Linear Model Fitting]
    H --> I[Empirical Bayes]
    I --> J[Extract Results]
    J --> K[Visualization & Interpretation]
```

# Setup

## Load Required Libraries

```{r setup}
# Load required packages
library(limma)       # Differential expression analysis
library(edgeR)       # RNA-seq data structures and filtering
library(tidyverse)   # Data manipulation and visualization
library(ggrepel)     # Better text labels on plots
library(pheatmap)    # Heatmaps
library(RColorBrewer) # Color palettes

# Set a seed for reproducibility
set.seed(42)

# Create output directory for figures
if (!dir.exists("figures")) {
  dir.create("figures")
}
```

::: {.callout-tip}
## Package Installation
If you haven't installed these packages yet, you can do so with:
```r
install.packages(c("tidyverse", "ggrepel", "pheatmap", "RColorBrewer"))
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(c("limma", "edgeR"))
```
:::

## Set File Paths

```{r paths}
# Define paths to your data files
# These should be outputs from the nf-core/rnaseq pipeline
counts_file <- "path/to/rsem.merged.gene_counts.tsv"
sample_info_file <- "path/to/sample_info.tsv"

# For this workshop, we'll use example paths
# Update these to match your actual file locations
```

::: {.callout-note}
## nf-core Output Structure
The nf-core/rnaseq pipeline with RSEM produces a merged count matrix file typically named:
`rsem.merged.gene_counts.tsv` in the `star_rsem/` results directory. This file contains raw counts for all samples across all genes.
:::

# Part 1: Data Loading and Preparation

## Load Gene Count Data

```{r load-counts}
# Load the RSEM gene counts
# The nf-core pipeline produces a merged file with:
# - gene_id: Ensembl gene ID
# - gene_name: Gene symbol
# - One column per sample with count data

expression_data <- read.table(counts_file, 
                              header = TRUE, 
                              row.names = "gene_name",
                              check.names = FALSE)

# Remove the gene_id column (we're using gene_name as row names)
expression_data$gene_id <- NULL

# Quick look at the data
head(expression_data, n = 3)
dim(expression_data)
```

::: {.callout-important}
## RSEM vs Salmon
Note that RSEM produces **expected counts** which are suitable for downstream analysis with limma-voom or DESeq2. These are not the same as TPM values - make sure you're using the `.genes.results` counts, not TPM values!
:::

## Load Sample Metadata

```{r load-metadata}
# Load sample information
# This file should contain:
# - sample: sample names (matching column names in expression_data)
# - Any experimental factors (treatment, genotype, time, etc.)

sample_info <- read.table(sample_info_file, 
                          header = TRUE, 
                          stringsAsFactors = FALSE)

# Display the sample information
sample_info

# Quick check: do sample names match?
all(sample_info$sample %in% colnames(expression_data))
```

::: {.callout-tip}
## Sample Metadata Best Practices
Your sample metadata file should include:
- Clear, consistent sample names matching your count data
- All experimental factors (treatment, batch, individual, etc.)
- Any known technical variables (sequencing run, library prep date)
- Keep a data dictionary explaining what each column means!
:::

# Part 2: Quality Control and Exploratory Analysis

## Create DGEList Object

The `DGEList` object from edgeR is a convenient container for RNA-seq data that bundles counts, sample information, and gene annotations together.

```{r create-dge}
# Create DGEList object
# This combines count data with sample metadata
DGE <- DGEList(counts = expression_data,
               samples = sample_info)

# Inspect the object
DGE
```

## Filter Lowly Expressed Genes

Not all genes are expressed in your samples. Filtering out genes with very low or zero counts:

- Reduces noise in the analysis
- Improves statistical power (fewer tests = less multiple testing burden)
- Speeds up computation

```{r filter-genes}
# The filterByExpr function determines which genes have 
# sufficient counts for reliable DE testing
# It uses the experimental design (group) to determine 
# the minimum number of samples that should have adequate counts

keep.exprs <- filterByExpr(DGE, group = DGE$samples$group)

# How many genes are we keeping?
table(keep.exprs)

# Filter the DGEList
DGE <- DGE[keep.exprs, , keep.lib.sizes = FALSE]

cat("After filtering, we have", nrow(DGE), "genes remaining\n")
```

::: {.callout-note}
## Understanding filterByExpr
`filterByExpr()` uses a sensible heuristic: a gene should be expressed at a minimum level in at least some proportion of samples from the smallest group. By default:
- Minimum count: 10 reads
- Minimum proportion: Depends on library sizes and design
This is more sophisticated than simple count cutoffs!
:::

## Library Size and Composition

Before normalization, let's check library sizes (total read counts per sample).

```{r library-sizes}
# Calculate library sizes
lib_sizes <- DGE$samples$lib.size

# Create a data frame for plotting
lib_size_df <- data.frame(
  sample = rownames(DGE$samples),
  lib_size = lib_sizes,
  group = DGE$samples$group
)

# Plot library sizes
ggplot(lib_size_df, aes(x = sample, y = lib_size / 1e6, fill = group)) +
  geom_col() +
  labs(title = "Library Sizes After Filtering",
       x = "Sample",
       y = "Library Size (Millions of Reads)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Set2")
```

::: {.callout-warning}
## What to Look For
- **Large differences** (>2-3x) in library sizes might indicate technical issues
- **Very low libraries** (<5M reads for human/mouse) may have insufficient depth
- Some variation is normal and will be handled by normalization
:::

## TMM Normalization

Trimmed Mean of M-values (TMM) normalization accounts for:

1. **Library size differences** - some samples have more total reads
2. **Composition bias** - a few highly expressed genes can "steal" reads from other genes

```{r normalize}
# Calculate normalization factors
DGE <- calcNormFactors(DGE, method = "TMM")

# Check normalization factors
# Values close to 1 indicate little composition bias
DGE$samples
```

::: {.callout-note}
## TMM Normalization Factors
- Factors close to 1.0: little correction needed
- Factors <1: this sample has upward composition bias (effective library is smaller)
- Factors >1: this sample has downward composition bias (effective library is larger)
These factors are used internally by limma-voom; you don't manually apply them to counts.
:::

## Multidimensional Scaling (MDS) Plot

MDS is similar to PCA and shows overall similarity between samples. Samples that are similar should cluster together.

```{r mds-plot, fig.width=8, fig.height=6}
# Calculate MDS
mds <- plotMDS(DGE, top = 500, plot = FALSE, gene.selection = "common")

# Create a data frame for ggplot
mds_df <- data.frame(
  sample = rownames(DGE$samples),
  dim1 = mds$x,
  dim2 = mds$y,
  group = DGE$samples$group
)

# Create MDS plot
ggplot(mds_df, aes(x = dim1, y = dim2, color = group, label = sample)) +
  geom_point(size = 4, alpha = 0.8) +
  geom_text_repel(size = 3, max.overlaps = 20) +
  labs(title = "MDS Plot - Sample Similarity",
       subtitle = "Based on top 500 most variable genes",
       x = paste0("Leading logFC dim 1 (", round(mds$var.explained[1] * 100, 1), "%)"),
       y = paste0("Leading logFC dim 2 (", round(mds$var.explained[2] * 100, 1), "%)")) +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_color_brewer(palette = "Set1")

# Save the plot
ggsave("figures/mds_plot.png", width = 8, height = 6, dpi = 300)
```

::: {.callout-important}
## Interpreting the MDS Plot
**Good signs:**
- Samples from the same group cluster together
- Clear separation between different conditions
- No obvious outliers

**Red flags:**
- Outlier samples far from their group
- No separation between conditions (weak signal)
- Samples clustering by batch instead of biology
:::

## Sample Correlation Heatmap

A complementary view showing pairwise sample correlations.

```{r correlation-heatmap, fig.width=8, fig.height=7}
# Get log-CPM values for visualization
logcpm <- cpm(DGE, log = TRUE)

# Calculate correlation matrix
cor_matrix <- cor(logcpm, method = "spearman")

# Create annotation for samples
annotation_col <- data.frame(
  Group = DGE$samples$group,
  row.names = colnames(logcpm)
)

# Create heatmap
pheatmap(cor_matrix,
         annotation_col = annotation_col,
         annotation_row = annotation_col,
         main = "Sample Correlation Heatmap",
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(0.7, 1, length.out = 101),
         fontsize = 8,
         filename = "figures/correlation_heatmap.png",
         width = 8,
         height = 7)

# Also display in the document
pheatmap(cor_matrix,
         annotation_col = annotation_col,
         annotation_row = annotation_col,
         main = "Sample Correlation Heatmap",
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(0.7, 1, length.out = 101),
         fontsize = 8)
```

::: {.callout-tip}
## Correlation Expectations
- **Within-group correlation**: typically 0.95-0.99
- **Between-group correlation**: depends on how different conditions are (0.85-0.98)
- **Outliers**: samples with consistently lower correlations (<0.90) to all others
:::

# Part 3: Differential Expression - Simple Two-Group Design

Let's start with a simple comparison: two conditions, like "treated vs. control" or "mutant vs. wildtype".

## Create Design Matrix

The design matrix tells limma which samples belong to which groups.

```{r design-simple}
# For this example, assume we have a 'treatment' factor with levels "control" and "treated"
# Adjust this based on your actual experimental design

# Make sure the factor levels are set correctly
# The first level will be the reference (denominator in fold-change)
DGE$samples$treatment <- factor(DGE$samples$treatment, 
                                levels = c("control", "treated"))

# Create design matrix
design_simple <- model.matrix(~treatment, data = DGE$samples)

# View the design matrix
design_simple

# Column names tell us what each coefficient represents
colnames(design_simple)
```

::: {.callout-note}
## Understanding the Design Matrix
- **Intercept (column 1)**: Average expression in the reference group
- **treatmenttreated (column 2)**: Difference between treated and control (log2 fold-change)

If a gene has:
- Positive coefficient: higher expression in treated
- Negative coefficient: higher expression in control
:::

## Run limma-voom with Quality Weights

The voom transformation:
1. Converts counts to log2-CPM
2. Estimates the mean-variance relationship
3. Calculates precision weights for each observation

Quality weights down-weight outlier samples.

```{r voom-simple, fig.width=8, fig.height=5}
# Run voom with quality weights
# normalize.method = "none" because we already did TMM normalization
vwts_simple <- voomWithQualityWeights(DGE, 
                                      design = design_simple,
                                      normalize.method = "none",
                                      plot = TRUE)
```

::: {.callout-note}
## Interpreting voom Plots
**Left panel (mean-variance trend):**
- Shows the relationship between average expression and variability
- Red curve = the trend being modeled
- At low expression: high variance (noisy)
- At high expression: lower variance (reliable)

**Right panel (sample weights):**
- All samples should have weights close to 1.0
- Weights <0.5 indicate problematic samples
- Weights are used to down-weight outliers without excluding them
:::

## Fit Linear Model

```{r fit-simple}
# Fit linear model to each gene
fit_simple <- lmFit(vwts_simple, design_simple)

# Apply empirical Bayes moderation
# This borrows information across genes to improve variance estimates
fit_simple <- eBayes(fit_simple, robust = TRUE)

# Quick summary of DE genes at FDR < 0.05
summary(decideTests(fit_simple, adjust.method = "fdr", p.value = 0.05))
```

::: {.callout-tip}
## Understanding decideTests Output
- **Down**: Genes significantly downregulated in treated vs. control (negative logFC)
- **NotSig**: Genes not significantly different
- **Up**: Genes significantly upregulated in treated vs. control (positive logFC)
:::

## Extract Top Results

```{r top-results}
# Get the top differentially expressed genes
# coef = 2 refers to the treatment effect (second column of design matrix)
top_genes <- topTable(fit_simple, 
                      coef = 2,
                      adjust.method = "BH",
                      number = 20,
                      sort.by = "P")

# Display top genes
top_genes
```

## Volcano Plot

A volcano plot shows log-fold change vs. statistical significance.

```{r volcano-simple, fig.width=8, fig.height=6}
# Extract all results
all_results <- topTable(fit_simple, 
                        coef = 2,
                        adjust.method = "BH",
                        number = Inf,
                        sort.by = "none")

# Add significance categories
all_results$significant <- case_when(
  all_results$adj.P.Val < 0.05 & all_results$logFC > 1 ~ "Up",
  all_results$adj.P.Val < 0.05 & all_results$logFC < -1 ~ "Down",
  all_results$adj.P.Val < 0.05 ~ "Significant",
  TRUE ~ "Not Significant"
)

# Create volcano plot
ggplot(all_results, aes(x = logFC, y = -log10(P.Value), color = significant)) +
  geom_point(alpha = 0.5, size = 1) +
  scale_color_manual(values = c("Up" = "red", 
                                "Down" = "blue", 
                                "Significant" = "orange",
                                "Not Significant" = "grey")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", alpha = 0.5) +
  labs(title = "Volcano Plot: Treated vs. Control",
       x = "Log2 Fold Change",
       y = "-Log10 P-value",
       color = "Significance") +
  theme_minimal() +
  theme(legend.position = "right")

ggsave("figures/volcano_plot_simple.png", width = 8, height = 6, dpi = 300)
```

## MA Plot

An MA plot shows log-fold change vs. average expression.

```{r ma-plot, fig.width=8, fig.height=6}
# Create MA plot
ggplot(all_results, aes(x = AveExpr, y = logFC, color = significant)) +
  geom_point(alpha = 0.5, size = 1) +
  scale_color_manual(values = c("Up" = "red", 
                                "Down" = "blue", 
                                "Significant" = "orange",
                                "Not Significant" = "grey")) +
  geom_hline(yintercept = 0, linetype = "solid", alpha = 0.5) +
  labs(title = "MA Plot: Treated vs. Control",
       x = "Average Expression (log2-CPM)",
       y = "Log2 Fold Change",
       color = "Significance") +
  theme_minimal()

ggsave("figures/ma_plot_simple.png", width = 8, height = 6, dpi = 300)
```

# Part 4: Complex Design - Two-Factor Analysis

Now let's extend to a more complex design. Suppose we have two factors:
- **Treatment**: control vs. treated
- **Genotype**: wildtype vs. mutant

We want to understand:
1. Main effect of treatment (across both genotypes)
2. Main effect of genotype (across both treatments)
3. Interaction: does the treatment effect depend on genotype?

## Create Two-Factor Design Matrix

```{r design-complex}
# Define factors with explicit levels
# Adjust these based on your actual experimental design
DGE$samples$genotype <- factor(DGE$samples$genotype, 
                               levels = c("wildtype", "mutant"))
DGE$samples$treatment <- factor(DGE$samples$treatment, 
                                levels = c("control", "treated"))

# Create two-factor design matrix
design_2factor <- model.matrix(~genotype + treatment, data = DGE$samples)

# View the design
design_2factor
colnames(design_2factor)
```

::: {.callout-note}
## Two-Factor Model Coefficients
- **Intercept**: Expression in wildtype + control (reference combination)
- **genotypemutant**: Effect of mutant vs. wildtype (averaged across treatments)
- **treatmenttreated**: Effect of treated vs. control (averaged across genotypes)

This is an **additive model** - it assumes the treatment effect is the same in both genotypes.
:::

## Fit Two-Factor Model

```{r voom-2factor, fig.width=8, fig.height=5}
# Run voom with quality weights for 2-factor design
vwts_2factor <- voomWithQualityWeights(DGE, 
                                       design = design_2factor,
                                       normalize.method = "none",
                                       plot = TRUE)

# Fit model
fit_2factor <- lmFit(vwts_2factor, design_2factor)
fit_2factor <- eBayes(fit_2factor, robust = TRUE)

# Summary
summary(decideTests(fit_2factor, adjust.method = "fdr", p.value = 0.05))
```

## Extract Results for Each Coefficient

```{r results-2factor}
# Treatment effect (main effect)
treatment_results <- topTable(fit_2factor, 
                              coef = "treatmenttreated",
                              adjust.method = "BH",
                              number = Inf)

cat("Number of genes DE for treatment:", 
    sum(treatment_results$adj.P.Val < 0.05), "\n")

# Genotype effect (main effect)
genotype_results <- topTable(fit_2factor, 
                             coef = "genotypemutant",
                             adjust.method = "BH",
                             number = Inf)

cat("Number of genes DE for genotype:", 
    sum(genotype_results$adj.P.Val < 0.05), "\n")
```

# Part 5: Advanced - Interaction Model and Contrasts

For more flexibility, we can use an **interaction model** and **contrasts** to test specific hypotheses.

## Create Interaction Design

```{r design-interaction}
# Create a combined factor for all group combinations
DGE$samples$group_combined <- factor(
  paste(DGE$samples$genotype, DGE$samples$treatment, sep = "_"),
  levels = c("wildtype_control", "wildtype_treated", 
             "mutant_control", "mutant_treated")
)

# Create design with no intercept (each group gets its own coefficient)
design_interaction <- model.matrix(~0 + group_combined, data = DGE$samples)

# Clean up column names
colnames(design_interaction) <- levels(DGE$samples$group_combined)

design_interaction
```

## Define Contrasts

Contrasts let us test specific biological questions:

```{r contrasts}
# Define contrasts of interest
cont.matrix <- makeContrasts(
  # Treatment effect in wildtype
  WT_TrtEffect = wildtype_treated - wildtype_control,
  
  # Treatment effect in mutant
  Mut_TrtEffect = mutant_treated - mutant_control,
  
  # Interaction: is treatment effect different between genotypes?
  Interaction = (mutant_treated - mutant_control) - 
                (wildtype_treated - wildtype_control),
  
  # Genotype effect in control
  Ctrl_GenoEffect = mutant_control - wildtype_control,
  
  # Genotype effect in treated
  Trt_GenoEffect = mutant_treated - wildtype_treated,
  
  levels = design_interaction
)

cont.matrix
```

## Fit Contrast Model

```{r fit-contrasts, fig.width=8, fig.height=5}
# Run voom
vwts_contrast <- voomWithQualityWeights(DGE, 
                                        design = design_interaction,
                                        normalize.method = "none",
                                        plot = TRUE)

# Fit model
fit_contrast <- lmFit(vwts_contrast, design_interaction)

# Apply contrasts
fit_contrast <- contrasts.fit(fit_contrast, cont.matrix)

# Empirical Bayes
fit_contrast <- eBayes(fit_contrast, robust = TRUE)

# Summary across all contrasts
summary(decideTests(fit_contrast, adjust.method = "fdr", p.value = 0.05))
```

## Visualize Contrast Results

```{r venn-diagram, fig.width=7, fig.height=7}
# Venn diagram showing overlap between contrasts
contrast_sig <- decideTests(fit_contrast, adjust.method = "fdr", p.value = 0.05)

# Focus on treatment effects in each genotype
vennDiagram(contrast_sig[, c("WT_TrtEffect", "Mut_TrtEffect")],
            circle.col = c("blue", "red"),
            main = "Overlap of Treatment Effects")
```

## Extract Genes with Specific Patterns

```{r pattern-extraction}
# Get full results for treatment effects
wt_trt <- topTable(fit_contrast, coef = "WT_TrtEffect", 
                   number = Inf, adjust.method = "BH")
mut_trt <- topTable(fit_contrast, coef = "Mut_TrtEffect", 
                    number = Inf, adjust.method = "BH")
interaction <- topTable(fit_contrast, coef = "Interaction", 
                       number = Inf, adjust.method = "BH")

# Merge results
all_contrasts <- data.frame(
  gene = rownames(wt_trt),
  WT_logFC = wt_trt$logFC,
  WT_pval = wt_trt$adj.P.Val,
  Mut_logFC = mut_trt$logFC,
  Mut_pval = mut_trt$adj.P.Val,
  Int_pval = interaction$adj.P.Val
)

# Find genes with treatment effect in BOTH genotypes (no interaction)
parallel_response <- all_contrasts %>%
  filter(WT_pval < 0.05 & Mut_pval < 0.05) %>%
  filter(Int_pval > 0.05) %>%
  filter(sign(WT_logFC) == sign(Mut_logFC))

cat("Genes with parallel response to treatment:", nrow(parallel_response), "\n")
head(parallel_response)

# Find genes with treatment effect ONLY in mutant (significant interaction)
mutant_specific <- all_contrasts %>%
  filter(Mut_pval < 0.05) %>%
  filter(WT_pval > 0.05) %>%
  filter(Int_pval < 0.05)

cat("\nGenes with mutant-specific treatment response:", nrow(mutant_specific), "\n")
head(mutant_specific)
```

# Part 6: Visualization and Export

## Heatmap of Top DE Genes

```{r heatmap-top-genes, fig.width=8, fig.height=10}
# Get top 50 genes from simple treatment comparison
top50_genes <- rownames(topTable(fit_simple, coef = 2, number = 50))

# Extract log-CPM values for these genes
top50_logcpm <- cpm(DGE[top50_genes, ], log = TRUE)

# Z-score normalize for better visualization
top50_scaled <- t(scale(t(top50_logcpm)))

# Create annotation
annotation_col <- data.frame(
  Treatment = DGE$samples$treatment,
  row.names = colnames(top50_logcpm)
)

# Create heatmap
pheatmap(top50_scaled,
         annotation_col = annotation_col,
         scale = "none",
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         show_rownames = TRUE,
         show_colnames = TRUE,
         main = "Top 50 DE Genes (Treatment Effect)",
         fontsize_row = 6,
         fontsize_col = 8,
         color = colorRampPalette(c("blue", "white", "red"))(100))

# Save
pheatmap(top50_scaled,
         annotation_col = annotation_col,
         scale = "none",
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         show_rownames = TRUE,
         show_colnames = TRUE,
         main = "Top 50 DE Genes (Treatment Effect)",
         fontsize_row = 6,
         fontsize_col = 8,
         color = colorRampPalette(c("blue", "white", "red"))(100),
         filename = "figures/heatmap_top50.png",
         width = 8,
         height = 10)
```

## Export Results Tables

```{r export-results}
# Create output directory
if (!dir.exists("results")) {
  dir.create("results")
}

# Export simple design results
write.csv(all_results, 
          file = "results/simple_design_all_genes.csv",
          row.names = TRUE)

# Export top DE genes
write.csv(top_genes,
          file = "results/simple_design_top20.csv",
          row.names = TRUE)

# Export contrast results
write.csv(all_contrasts,
          file = "results/contrast_results_merged.csv",
          row.names = FALSE)

# Export parallel response genes
write.csv(parallel_response,
          file = "results/parallel_response_genes.csv",
          row.names = FALSE)

cat("Results exported to 'results/' directory\n")
```

# Summary and Key Takeaways

## What We Covered

1. **Data Loading**: RSEM counts from nf-core pipeline
2. **Quality Control**: Library sizes, MDS plots, sample correlations
3. **Filtering & Normalization**: Removing low counts, TMM normalization
4. **Simple Design**: Two-group comparison with limma-voom
5. **Complex Design**: Two-factor model with main effects
6. **Advanced Contrasts**: Interaction model for specific biological questions
7. **Visualization**: Volcano plots, MA plots, heatmaps

## Key Concepts to Remember

::: {.callout-important}
## Critical Points

1. **Filter before normalizing** - remove genes with insufficient counts
2. **TMM normalization** handles library size and composition bias
3. **voom transformation** converts counts to log-scale with precision weights
4. **Quality weights** handle outlier samples without removing them
5. **Design matrix** defines what you're testing
6. **Contrasts** let you ask specific biological questions
7. **Multiple testing correction** (FDR/BH) is essential
8. **LogFC and p-value together** tell the full story (effect size + significance)
:::

## Next Steps

- **Gene Set Enrichment Analysis**: Are DE genes enriched for specific pathways/functions?
- **Validation**: qRT-PCR or additional experiments to confirm key findings
- **Biological Interpretation**: Connect results back to your biological question
- **Reproducibility**: Document your analysis, save your code, version control!

## Resources

- [limma User's Guide](https://bioconductor.org/packages/release/bioc/vignettes/limma/inst/doc/usersguide.pdf)
- [edgeR User's Guide](https://bioconductor.org/packages/release/bioc/vignettes/edgeR/inst/doc/edgeRUsersGuide.pdf)
- [RNA-seq analysis in R (Harvard Chan Bioinformatics Core)](https://hbctraining.github.io/DGE_workshop/)
- UF Research Computing: [https://www.rc.ufl.edu/](https://www.rc.ufl.edu/)

---

## Session Information

```{r sessioninfo}
sessionInfo()
```
